---
title: "Entomological Database Creation – Cienfuegos"
author: "Rita Verstraeten"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: true
    toc_depth: 3
    number_sections: true
editor_options:
  chunk_output_type: console
---

## Install R packages
```{r setup, include = FALSE}
# renv::init()  # creates renv/ folder and lockfile
# Install necessary packages
# if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
# if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
library(readxl)
library(dplyr)
# install.packages(c("lubridate", "readr"))
library(lubridate)
library(readr)
library(DT)
library(ggplot2)
# if (interactive() && isTRUE(capabilities("httpgd"))) {
#   httpgd::hgd(silent = TRUE)  # use httpgd only in interactive sessions
# }
library(httpgd)

knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)

# install.packages(c("rmarkdown", "rticles", "pagedown", "blogdown"))
```


# Entomological data
```{r read the entomo file}

# general path to all data files
data_path <- "/media/rita/New Volume/Documenten/DI-MOB/Data Sharing"

# path to the mosquito surveillance data 
entomo_path <- file.path(data_path, "WP2_Entomological_Data", "Entomo_surveillance_11-24_secondary_DB_Month_Mz_v05082025.xlsx")

# Read and merge the Excel files
# Specify col_types to avoid warnings about AP column mixing logical/text
entomo_data <- read_excel(entomo_path, sheet = "Data", col_types = c(rep("guess", 5), "text", rep("guess", 10)))
entomo_data <- entomo_data %>%
  mutate(year_month = as.Date(paste(Year, Month, "01", sep = "-")))%>%  # create year-month date
  # select(year_month, CODIGO_DIMOB, CP, Area, Inspected_houses, Houses_pos_L)
  select(year_month, CODIGO_DIMOB, Inspected_houses, Houses_pos_L)

entomo_data$Inspected_houses <- as.integer(entomo_data$Inspected_houses)
entomo_data$Houses_pos_L <- as.integer(entomo_data$Houses_pos_L)

print(tail(entomo_data))
# View(entomo_data)
```

Now that the entomological data is loaded, we can perform some logical checks.
```{r logical checks entomo}
message("Number of missing Manzana names: ", sum(entomo_data$CODIGO_DIMOB == "DD")) # 168!!
```

# Epidemiological data
```{r read the epi data}
epi_path <- file.path(data_path, "WP1_Epidemiological_Data", "DENV", "DENV_secondaryDB_12_24_spatial_smoothed_v14102025_forsharing.xlsx")

epi_data <- read_excel(epi_path, sheet="DENV data")
print(head(epi_data))
print(colnames(epi_data))
```

Keep you only the relevant columns: 
- date of diagnosis
- manzana code
<!-- - CP
- Area
- CMF --> we will be getting this from the shapefiles direclty. 

```{r keep selected epi columns}
epi_data <- epi_data %>%
  mutate(date = as.Date(Date_diagnosis),              # ensure date is Date type
         year_month = floor_date(date, "month")) %>%    # extract year-month as Date
  # select(year_month, CODIGO_DIMOB, CP, Area, CMF) 
  select(year_month, CODIGO_DIMOB) 

head(epi_data)
```

Summarize the weekly cases to monthly values per house block (CODIGO_DIMOB)
```{r count cases per block-month}
epi_data <- epi_data %>%
  group_by(CODIGO_DIMOB, year_month) %>%
  summarise(
    cases = n(),
    # CP = first(CP),
    # Area = first(Area),
    # CMF = first(CMF),
    .groups = "drop"
  ) %>%
  arrange(year_month, CODIGO_DIMOB)
head(epi_data)
# DT::datatable(epi_data)
```

# Combine entomological & epidemiological data 

```{r combine entomo and epi based on time and block}
# ensure both are Date
stopifnot(inherits(entomo_data$year_month, "Date"))
stopifnot(inherits(epi_data$year_month, "Date"))

# # aggregate entomo by month/block (sums here; adjust if you need mean/other)
# message("number of rows before aggregation:", nrow(entomo_data))
# entomo_month <- entomo_data %>%
#   group_by(CODIGO_DIMOB, year_month) %>%
#   summarise(
#     Total_houses = sum(Total_houses, na.rm = TRUE),
#     Houses_pos_L = sum(Houses_pos_L, na.rm = TRUE),
#     .groups = "drop"
#   )
# message("number of rows after aggregation:", nrow(entomo_month)) # -> the same, so not needed here

# join and fill missing combos with zeros
ento_epi_data <- full_join(entomo_data, epi_data, by = c("CODIGO_DIMOB", "year_month")) %>%
  mutate(
    Inspected_houses = coalesce(Inspected_houses, 0), # return Inspected_houses, or 0 if NA
    Houses_pos_L = coalesce(Houses_pos_L, 0),
    cases = coalesce(cases, 0)
  ) %>%
  arrange(year_month, CODIGO_DIMOB)

```

# Filter for 2015-2017
```{r filter for 2015-2017}
ento_epi_data <- ento_epi_data %>%
  filter(year_month >= as.Date("2015-01-01") & year_month <= as.Date("2017-12-31"))
# DT::datatable(ento_epi_data)

```

# Create some plots 

## Load the maps 
the shapefile corresponding to the Manzanas in Cienfuegos is in Projected CRS: NAD27 / Cuba Norte. 
This is very important to keep in mind when joining the environmental data later on. 

```{r load the mz maps}
library(sf)
shapefile_path <- "/media/rita/New Volume/Documenten/DI-MOB/Data Sharing/WP1_Cartographic_data/Administrative borders"
# Read the shapefile
dimob_manzanas <- st_read(file.path(shapefile_path, "Manzanas vectores 21_11_2024_CMF column uncleaned/manzanero_AS_1a8_19112024.shp"))
dimob_areas <- st_read(file.path(shapefile_path, "Area de Salud/AS LIMITES 3022025.shp"))
dimob_cps <- st_read(file.path(shapefile_path, "Consejo Popular/CONSEJOS Populares Cfgos 3022025.shp"))

# head(dimob_manzanas)
# head(dimob_areas)
# head(dimob_cps)
```

### Remove all accents and set to uppercase for consistency
```{r clean names in shapefiles}
library(stringi)

# Clean NOMBRE column in all shapefiles: remove accents and uppercase
# dimob_manzanas <- dimob_manzanas %>%
#   mutate(CODIGO_ = stri_trans_general(CODIGO_, "Latin-ASCII"),
#          CODIGO_ = toupper(CODIGO_))

# dimob_areas <- dimob_areas %>%
#   mutate(NOMBRE = stri_trans_general(NOMBRE, "Latin-ASCII"),
#          NOMBRE = toupper(NOMBRE))

dimob_cps <- dimob_cps %>%
  mutate(NOMBRE = stri_trans_general(NOMBRE, "Latin-ASCII"),
         NOMBRE = toupper(NOMBRE),
         NOMBRE = gsub("-", " ", NOMBRE))

head(dimob_cps$NOMBRE)
```


```{r safety check}
# let's check that the manzana names in the shapefile correspond to those in the ento_epi_data
shapefile_mz_names <- unique(dimob_manzanas$CODIGO_)
data_mz_names <- unique(ento_epi_data$CODIGO_DIMOB)

# print number of unique manzanas in shapefiles
cat("Number of unique manzanas in shapefile: \n", length(shapefile_mz_names), "\n") # 1410

# check which manzanas ARE in the shapefiles, but not in the data
cat("\nManzanas in shapefile but not in data:\n")
unique(setdiff(shapefile_mz_names, data_mz_names))  # NA               "A4CPSLMzTaller"

# check which manzanas ARE in the data, but not in the shapefiles
cat("\nManzanas in data but not in shapefile: \n")
unique(setdiff(data_mz_names, shapefile_mz_names))  #  "A7CNAMzPPESC" "DD"   
length(unique(setdiff(data_mz_names, shapefile_mz_names)))  # should be 0 -> is now 2

```

Let's filter out the data-points in the 2 manzanas that are not in the shapefile.
```{r filter out datapoint that do not have a valid location}
cat("Number of rows before filtering: ", nrow(ento_epi_data), "\n") #50760
ento_epi_data <- ento_epi_data %>%
  filter(!CODIGO_DIMOB %in% c("A7CNAMzPPESC", "DD"))
cat("Number of rows after filtering: ", nrow(ento_epi_data), "\n") # 50688 -> 72 removed
```

## Add the CP, Area, and CMF to each manzana based on the shapefiles
```{r add CP CMF and Area to ento_epi_data}
# After loading dimob_manzanas, extract the non-geometry columns
manzana_info <- st_drop_geometry(dimob_manzanas) %>%
  select(CODIGO_, AS, CP, CMF) %>%
  rename(CODIGO_DIMOB = CODIGO_) %>%
  distinct()  # keep one row per manzana in case of duplicates

# Then join to ento_epi_data
ento_epi_data <- ento_epi_data %>%
  left_join(manzana_info, by = "CODIGO_DIMOB")

head(ento_epi_data)
```


Let's try out a plot for all the Mzs with entomo data for the entire year of 2015. 

```{r plot entomo data 2015}
# aggregate to annual: sum counts, keep unique inspected_houses per block
ento_epi_year <- ento_epi_data %>%
  mutate(year = floor_date(year_month, "year")) %>%
  group_by(CODIGO_DIMOB, year) %>%
  summarise(
    Houses_pos_L = sum(Houses_pos_L, na.rm = TRUE),
    cases = sum(cases, na.rm = TRUE),
    Inspected_houses = first(Inspected_houses),
    .groups = "drop"
  ) %>%
  arrange(year, CODIGO_DIMOB)

# View(ento_epi_year)
```

Plot the Houses_pos_L for 2015 per manzana
add the areas map in the background for context
```{r plot 2015 to 2017 entomo}

# Plot Houses_pos_L for 2015–2017 on the manzana map (faceted)
plot_entomo_facets <- ento_epi_year %>%
  filter(year %in% as.Date(c("2015-01-01", "2016-01-01", "2017-01-01"))) %>%
  rename(CODIGO_ = CODIGO_DIMOB) %>%
  right_join(dimob_manzanas, by = "CODIGO_") %>%
  st_as_sf() %>%
  filter(!is.na(year)) %>%
  ggplot() +
  geom_sf(data = dimob_areas, fill = NA, color = "grey60", linewidth = 0.4) + # add the area boundaries in background
  geom_sf(aes(fill = Houses_pos_L), color = NA) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = "Nr. of houses positive for larvae by Consejo Popular",
       fill = "Houses_pos_L") +
  facet_wrap(~year, ncol = 3) +
  theme_minimal()

plot_entomo_facets
```

Plot the cases for 2015
```{r plot 2015 epi data}
plot_entomo_cases <- ento_epi_year %>%
  filter(year %in% as.Date(c("2015-01-01", "2016-01-01", "2017-01-01"))) %>%
  rename(CODIGO_ = CODIGO_DIMOB) %>%
  right_join(dimob_manzanas, by = "CODIGO_") %>%
  st_as_sf() %>%
  filter(!is.na(year)) %>%
  ggplot() +
  geom_sf(data = dimob_areas, fill = NA, color = "grey60", linewidth = 0.4) +
  geom_sf(aes(fill = cases), color = NA) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = "Nr. of cases by Manzana",
       fill = "Cases") +
  facet_wrap(~year, ncol = 3) +
  theme_minimal()

plot_entomo_cases
```

## Now we make the plot per CP 
```{r plot the CP shapefile alone}
# Quick plot of the CP shapefile
ggplot(dimob_cps) +
  geom_sf(fill = "lightblue", color = "black") +
  geom_sf_text(aes(label = CP), size = 2, check_overlap = TRUE) +
  labs(title = "Consejo Popular Boundaries") +
  theme_minimal()
```

```{r plot case data for 2015 to 2017 per CP}
# Aggregate to annual by CP
ento_epi_year_cp <- ento_epi_data %>%
  mutate(year = floor_date(year_month, "year")) %>%
  group_by(CP, year) %>%
  summarise(
    Houses_pos_L = sum(Houses_pos_L, na.rm = TRUE),
    cases = sum(cases, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year, CP)

plot_cases_cp <- ento_epi_year_cp %>%
  filter(year %in% as.Date(c("2015-01-01", "2016-01-01", "2017-01-01"))) %>%
  right_join(dimob_cps, by = c("CP" = "NOMBRE"), relationship = "many-to-many") %>%
  st_as_sf() %>%
  filter(!is.na(year)) %>%
  ggplot() +
  geom_sf(data = dimob_areas, fill = NA, color = "grey60", linewidth = 0.4) +
  geom_sf(aes(fill = cases), color = NA) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = "Cases by Consejo Popular", fill = "Cases") +
  facet_wrap(~year, ncol = 3) +
  theme_minimal()

plot_cases_cp
```

# average manzana surface area
```{r average mz area}
dimob_manzanas <- dimob_manzanas %>%
  st_transform(crs = 26917)  # NAD27 / UTM zone 17N
dimob_manzanas$area_m2 <- st_area(dimob_manzanas)
average_area_m2 <- mean(as.numeric(dimob_manzanas$area_m2))
median_area_m2 <- median(as.numeric(dimob_manzanas$area_m2))

average_area_m2  # ~93693.96 m2 -> ~ 100.000m2
```
This average area per manzana seems way too large... 
```{r plot the manzanas with area}
ggplot(dimob_manzanas) +
  geom_sf(aes(fill = as.numeric(area_m2)/10000), color = "grey40", linewidth = 0.1) + # area in hectares
  geom_sf_text(aes(label = CODIGO_), size = 1.5, check_overlap = TRUE) +
  scale_fill_viridis_c(option = "plasma", name = "Area (ha)") +
  labs(title = "Manzanas by Area") +
  theme_minimal()
```

```{r histogram of manzana areas}
# Create histogram of manzana areas
ggplot(dimob_manzanas, aes(x = as.numeric(area_m2))) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Distribution of Manzana Areas (log scale)",
       x = "Area (m²)",
       y = "Frequency") +
  theme_minimal() +
  geom_vline(xintercept = average_area_m2, color = "red", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = median_area_m2, color = "blue", linetype = "dashed", linewidth = 1) +
  annotate("text", x = average_area_m2, y = Inf, 
           label = paste0("Mean: ", round(average_area_m2, 0), " m²"), 
           hjust = -0.1, vjust = 2, color = "red") +
  annotate("text", x = median_area_m2, y = Inf, 
           label = paste0("Median: ", round(median_area_m2, 0), " m²"), 
           hjust = -0.1, vjust = 4, color = "blue")
```

There is a very large variation in manzana sizes. Averaging over the manzana surface therefore doesn't seem like a safe option. 
What would make most sense -> to use the houseblock location itself (centroid) + 400m buffer around it to extract environmental data.

**Actually**, the manzanas that have classification "TB" in the "MZ" variable indicate that these are *terreno baldio*, which means emtpy terrain. 
So, these manzanas are actually not of interest for our analysis given that there will not be any mosquito measures in absence of houses (and therefore population). 

### Filter manzanas to exclude those with "TB" in the MZ field
```{r filter manzanas without TB}
# Keep only manzanas that do NOT have "TB" in the MZ field
dimob_manzanas_filtered <- dimob_manzanas %>%
  filter(!grepl("TB", MZ, ignore.case = TRUE))

cat("Original number of manzanas:", nrow(dimob_manzanas), "\n") #1480
cat("Number of manzanas after filtering out TB:", nrow(dimob_manzanas_filtered), "\n") # 1415
cat("Number of TB manzanas removed:", nrow(dimob_manzanas) - nrow(dimob_manzanas_filtered), "\n") # 65 removed
```

### Re-plot the manzanas + area with indication of the TB manzanas
```{r plot the non-TB manzanas}
dimob_manzanas_filtered <- dimob_manzanas_filtered %>%
  st_transform(crs = 26917)  # NAD27 / UTM zone 17N
dimob_manzanas_filtered$area_m2 <- st_area(dimob_manzanas_filtered)
average_area_m2 <- mean(as.numeric(dimob_manzanas_filtered$area_m2))
median_area_m2 <- median(as.numeric(dimob_manzanas_filtered$area_m2))
average_area_m2  # ~29258.98 m2 -> ~ 30.000 m2

# plot the filtered manzanas shapefile
ggplot(dimob_manzanas_filtered) +
  geom_sf(aes(fill = as.numeric(area_m2)/10000), color = "grey40", linewidth = 0.1) + # area in hectares
  geom_sf_text(aes(label = CODIGO_), size = 1.5, check_overlap = TRUE) +
  scale_fill_viridis_c(option = "plasma", name = "Area (ha)") +
  labs(title = "Manzanas by Area") +
  theme_minimal()
```

```{r manzana area histogram after filtering out empty spaces}

# histogram of the manzana areas after filtering out TB manzanas
ggplot(dimob_manzanas_filtered, aes(x = as.numeric(area_m2))) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Distribution of Manzana Areas (log scale)",
       x = "Area (m²)",
       y = "Frequency") +
  theme_minimal() +
  geom_vline(xintercept = average_area_m2, color = "red", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = median_area_m2, color = "blue", linetype = "dashed", linewidth = 1) +
  annotate("text", x = average_area_m2, y = Inf, 
           label = paste0("Mean: ", round(average_area_m2, 0), " m²"), 
           hjust = -0.1, vjust = 2, color = "red") +
  annotate("text", x = median_area_m2, y = Inf, 
           label = paste0("Median: ", round(median_area_m2, 0), " m²"), 
           hjust = -0.1, vjust = 4, color = "blue")
```

However, for the larger manzanas, I wonder whether we have the exact houseblock location coordinates. 


Some manzanas are in fact huge. 


Final choice: use centroid per manzana + 500m buffer to extract environmental data.
This gives us an area of $\pi$*400^2 = 502.400 m2, which is larger than the average manzana size (around 30,000 m2 after removing TB manzanas).



# Add the environmental data to the ento_epi_data






