---
title: "Entomological–Epidemiological–Environmental Database – Cienfuegos"
author: "Rita Verstraeten"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: true
    toc_depth: 3
    number_sections: true
editor_options:
  chunk_output_type: console
---
# Entomological–Epidemiological–Environmental Database (Cienfuegos)
## Target table structure
The core table for the analysis should look like: 

Observations table (one row per block-time combination):

|Column | Type | Description |
|---|---|---|
|manzana | String | manzana identifier (1 to B) |
|time | Integer | time period identifier (1 to T) |
|N_HH | Integer | Number of households inspected (universo) |
|Cases | Integer | Number of dengue cases observed in that block-time |
|Houses_pos_IS | Integer | Number of positive mosquito samples detected |
|n_bt | Integer | Total inspection events (N_HH + reactive adjustments) |
|ndvi | Numeric | Normalized Difference Vegetation Index for manzana-month |
|ndmi | Numeric | Normalized Difference Moisture Index for manzana-month |
|ndwi | Numeric | Normalized Difference Water Index for manzana-month |
|wS2M | Numeric | average windspeed at 2 meter height in m/s |
|landcover | categorical | dominant landcover value for that mz |
|wui | Integer | TRUE/FALSE value for wild-land urban interface for that manzana |
|wi | Integer | TRUE/FALSE value for water interface for that manzana (or within 100m of manzana) |
|urban | Integer | TRUE/FALSE value for whether/not urban for that manzana |
|has_aljibes | Binary | TRUE/FALSE value for whether/not there are underground cisternas within a 20m radius of the manzana |
|nr_aljibes | Numeric | Number of underground water cisternas within 20m of the manzana |
|water_shortage | Binary | TRUE/FALSE value for whether/not manzanas have reduced water provision (more days between water provision cycles) |
|water_containers | Continuous | Average number of water containers per person available at the popular counsil level |
|avg_temp | Numeric | average temperature (°C) for that manzana-month |
|total_precip | Numeric | cummulative precipitation (mm) for that manzana-month | max_precip_daily
|precip_max_day | Numeric | cummulative precipitation (mm/24h) for the day in that manzana-month with the highest precipitation | 
|consec_rainy_days | Numeric | Maximum number of consecutive days with rain (> 1 mm) per month | 
|rel_hum | Numeric | average relative humidity (%) for that manzana-month |

<!-- |omega |	Numeric	| Reactive bias weight (calculated from C_bt) | -->

## Setup: packages
```{r setup, include = FALSE}
# Activate project renv (if present) without forcing full restore
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
}

# Install only packages needed by this notebook
required_pkgs <- c("readxl", "dplyr", "lubridate", "readr", "DT", "ggplot2", "tidyr", "sf", "stringi", "writexl", "scales")
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_pkgs) > 0) {
  message("Installing missing packages: ", paste(missing_pkgs, collapse = ", "))
  if (requireNamespace("renv", quietly = TRUE)) {
    renv::install(missing_pkgs)
  } else {
    install.packages(missing_pkgs)
  }
}

library(readxl)
library(dplyr)
library(lubridate)
library(readr)
library(DT)
library(ggplot2)
library(tidyr)  # for pivot_longer
library(sf)
# if (interactive() && isTRUE(capabilities("httpgd"))) {
#   httpgd::hgd(silent = TRUE)  # use httpgd only in interactive sessions
# }
# library(httpgd)

knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)
save_images_to <- "/media/rita/New Volume/Documenten/DI-MOB/Other Data/Env_data_cuba/images"
# install.packages(c("rmarkdown", "rticles", "pagedown", "blogdown"))
```


# Entomological data
```{r read the entomo file}

# general path to all data files
data_path <- "/media/rita/New Volume/Documenten/DI-MOB/Data Sharing"

# path to the mosquito surveillance data 
entomo_path <- file.path(data_path, "WP2_Entomological_Data", "Entomo_surveillance_11-24_secondary_DB_Month_Mz_v05082025.xlsx")

# Read and merge the Excel files
# Specify col_types to avoid warnings about AP column mixing logical/text
entomo_data <- read_excel(entomo_path, sheet = "Data", col_types = c(rep("guess", 5), "text", rep("guess", 10)))
entomo_data <- entomo_data %>%
  mutate(year_month = as.Date(paste(Year, Month, "01", sep = "-")))%>%  # create year-month date
  # select(year_month, CODIGO_DIMOB, CP, Area, Inspected_houses, Houses_pos_L)
  select(year_month, CODIGO_DIMOB, Inspected_houses, Houses_pos_IS)

entomo_data$Inspected_houses <- as.integer(entomo_data$Inspected_houses)
entomo_data$Houses_pos_IS <- as.integer(entomo_data$Houses_pos_IS)

print(tail(entomo_data))
# View(entomo_data)
```

Now that the entomological data is loaded, we can perform some logical checks.
```{r logical checks entomo}
message("Number of missing Manzana names: ", sum(entomo_data$CODIGO_DIMOB == "DD")) # 168!!

entomo_check <- entomo_data %>%
  filter(year_month < as.Date(2020-01-01) & year_month >= as.Date(2016-01-01))
message("Number of missing Manzana names: ", sum(entomo_check$CODIGO_DIMOB == "DD")) # 0
```

# Epidemiological data
In the epidemiological data we will add both DENV and ZIKV data. 
This is done because the main focus of the analysis is on environmental drives of Aedes mosquito presence. 
The disease data is only used to correct for reactive mosquito surveillance post febrile-illness report at the hospital.  

```{r read the epi data}
denv_path <- file.path(data_path, "WP1_Epidemiological_Data", "DENV", "DENV_secondaryDB_12_24_spatial_smoothed_v14102025_forsharing.xlsx")
zikv_path <- file.path(data_path, "WP1_Epidemiological_Data", "ZIKA", "ZIKA_secondaryDB_16-18_spatial_v09042025_forsharing.xlsx")

denv_data <- read_excel(denv_path, sheet="DENV data")
zikv_data <- read_excel(zikv_path, sheet="ZIKA data")
print(head(denv_data))
print(colnames(denv_data))
print(colnames(zikv_data))
```

```{r combine DENV and ZIKV data}

# cat("\nPresent in ZIKV but not in DENV data: ", setdiff(colnames(zikv_data), colnames(denv_data)))  # check which columns are different
# cat("\nPresent in DENV but not in ZIKV data: ",setdiff(colnames(denv_data), colnames(zikv_data)))  # check which columns are different

zikv_data <- zikv_data %>%
  select("CODIGO_DIMOB", "Date_diagnosis")  # keep only columns present in DENV data

zikv_data$arbovirus <- "ZIKV"
denv_data$arbovirus <- "DENV"
# add the zikv data to the denv data based on date of diagnosis and manzana code

epi_data <- bind_rows(denv_data, zikv_data) %>%
  arrange(Date_diagnosis, CODIGO_DIMOB)
```

Keep you only the relevant columns: 
- date of diagnosis
- manzana code
<!-- - CP
- Area
- CMF --> ~~we will be getting this from the shapefiles direclty.~~ We will be getting this from the Mz distribution xlsx

```{r keep selected epi columns}
epi_data <- epi_data %>%
  mutate(date = as.Date(Date_diagnosis),              # ensure date is Date type
         year_month = floor_date(date, "month")) %>%    # extract year-month as Date
  # select(year_month, CODIGO_DIMOB, CP, Area, CMF) 
  select(year_month, CODIGO_DIMOB, arbovirus) 

head(epi_data)
```

Summarize the weekly cases to monthly values per house block (CODIGO_DIMOB)
```{r count cases per block-month}
epi_data <- epi_data %>%
  group_by(CODIGO_DIMOB, year_month) %>%
  summarise(
    cases = n(),
    # CP = first(CP),
    # Area = first(Area),
    # CMF = first(CMF),
    .groups = "drop"
  ) %>%
  arrange(year_month, CODIGO_DIMOB)
head(epi_data)
# DT::datatable(epi_data)
```

# Combine entomological and epidemiological data

```{r combine entomo and epi based on time and block}
# ensure both are Date
stopifnot(inherits(entomo_data$year_month, "Date"))
stopifnot(inherits(epi_data$year_month, "Date"))

# # aggregate entomo by month/block (sums here; adjust if you need mean/other)
# message("number of rows before aggregation:", nrow(entomo_data))
# entomo_month <- entomo_data %>%
#   group_by(CODIGO_DIMOB, year_month) %>%
#   summarise(
#     Total_houses = sum(Total_houses, na.rm = TRUE),
#     Houses_pos_L = sum(Houses_pos_L, na.rm = TRUE),
#     .groups = "drop"
#   )
# message("number of rows after aggregation:", nrow(entomo_month)) # -> the same, so not needed here

# join and fill missing combos with zeros
ento_epi_data <- full_join(entomo_data, epi_data, by = c("CODIGO_DIMOB", "year_month")) %>%
  mutate(
    Inspected_houses = coalesce(Inspected_houses, 0), # return Inspected_houses, or 0 if NA
    Houses_pos_IS = coalesce(Houses_pos_IS, 0), # return Houses_pos_IS, or 0 if NA
    cases = coalesce(cases, 0)
  ) %>%
  arrange(year_month, CODIGO_DIMOB)

```

## Filter to 2015–2019
```{r filter for 2015-2019}
ento_epi_data <- ento_epi_data %>%
  filter(year_month >= as.Date("2015-01-01") & year_month <= as.Date("2019-12-31"))
# DT::datatable(ento_epi_data)

```

# Spatial reference data and ID checks

## Load spatial layers (shapefiles)
the shapefile corresponding to the Manzanas in Cienfuegos is in Projected CRS: NAD27 / Cuba Norte. 
This is very important to keep in mind when joining the environmental data later on. 

```{r load the mz maps}
library(sf)
shapefile_path <- "/media/rita/New Volume/Documenten/DI-MOB/Data Sharing/WP1_Cartographic_data/Administrative borders"
# Read the shapefile
dimob_manzanas_old <- st_read(file.path(shapefile_path, "Manzanas vectores 21_11_2024_CMF column uncleaned/manzanero_AS_1a8_19112024.shp"))
dimob_manzanas <- st_read(file.path(shapefile_path, "Manzanas_cleaned_05032026", "Mz_CMF_Correcto_2022026.shp"))

dimob_areas <- st_read(file.path(shapefile_path, "Area de Salud/AS LIMITES 3022025.shp"))
dimob_cps <- st_read(file.path(shapefile_path, "Consejo Popular/CONSEJOS Populares Cfgos 3022025.shp"))

# head(dimob_manzanas)
# head(dimob_areas)
# head(dimob_cps)
```

## Remove out-of-bound manzana (A8APSMMzFcSAMA)

```{r remove A8APSMMzFcSAMA from the list of manzanas and all the data}
# dimob_manzanas <- dimob_manzanas %>%
#   filter(CODIGO_ != "A8APSMMzFcSAMA")

nrow(dimob_manzanas) # 1414
dimob_manzanas <- dimob_manzanas %>%
  filter(CODIGO_ != "")

nrow(dimob_manzanas) # 1409 (5 removed)


# ento_epi_data <- ento_epi_data %>%
#   filter(CODIGO_DIMOB != "A8APSMMzFcSAMA")
```

### Normalize names for joins
```{r clean names in shapefiles}
library(stringi)

# Clean NOMBRE column in all shapefiles: remove accents and uppercase
# dimob_manzanas <- dimob_manzanas %>%
#   mutate(CODIGO_ = stri_trans_general(CODIGO_, "Latin-ASCII"),
#          CODIGO_ = toupper(CODIGO_))

# dimob_areas <- dimob_areas %>%
#   mutate(NOMBRE = stri_trans_general(NOMBRE, "Latin-ASCII"),
#          NOMBRE = toupper(NOMBRE))

dimob_cps <- dimob_cps %>%
  mutate(NOMBRE = stri_trans_general(NOMBRE, "Latin-ASCII"),
         NOMBRE = toupper(NOMBRE),
         NOMBRE = gsub("-", " ", NOMBRE))

head(dimob_cps$NOMBRE)
```


```{r safety check}
# let's check that the manzana names in the shapefile correspond to those in the ento_epi_data
shapefile_mz_names <- unique(dimob_manzanas$CODIGO_)
data_mz_names <- unique(ento_epi_data$CODIGO_DIMOB)

# print number of unique manzanas in shapefiles
cat("Number of unique manzanas in shapefile: \n", length(shapefile_mz_names), "\n") # 1410

# check which manzanas ARE in the shapefiles, but not in the data
cat("\nManzanas in shapefile but not in data:\n")
unique(setdiff(shapefile_mz_names, data_mz_names))  # "A4CPBVMzAERO"   "A4CPSLMzTaller"

# check which manzanas ARE in the data, but not in the shapefiles
cat("\nManzanas in data but not in shapefile: \n")
unique(setdiff(data_mz_names, shapefile_mz_names))  #   "A8APGuMzFcPSOL" "A8APSMMzFcSAMA" "DD"     
length(unique(setdiff(data_mz_names, shapefile_mz_names)))  # should be 0 -> is now 3

```

Let's filter out the data-points in the 2 manzanas that are not in the shapefile.
```{r filter out datapoint that do not have a valid location}
cat("Number of rows before filtering: ", nrow(ento_epi_data), "\n") #50760 -> 84600 
ento_epi_data <- ento_epi_data %>%
  filter(!CODIGO_DIMOB %in% c("A8APGuMzFcPSOL", "A8APSMMzFcSAMA", "DD" ))
cat("Number of rows after filtering: ", nrow(ento_epi_data), "\n") # 50688 -> 84420 
```


## Manzana allocation to CP/AREA/CMF
```{r load the manzana allocation file}
# mz_allocation_path <- "/media/rita/New Volume/Documenten/DI-MOB/Data Sharing/WP1_ Demographic_Data/Mzlist_overview CMF allocation_Shared between Areas_Original Mzlist_PerArea.xlsx"

# mz_allocation <- suppressWarnings(read_excel(mz_allocation_path, sheet = "unique CMF allocation")) %>%
#   select(1:10) %>% # selects first 10 columns (A-J)
#   select(CODIGO_DIMOB, Area...2, MAP_PC, CMF_unico, MAP_Universo, Poblacion_3) %>%
#   rename(CP = MAP_PC, 
#   AREA = Area...2,
#   CMF = CMF_unico,
#   UNIVERSE = MAP_Universo)

mz_allocation_path <- "/media/rita/New Volume/Documenten/DI-MOB/Data Sharing/WP1_Cartographic_data/Administrative borders/Manzanas_cleaned_05032026/Mzlist_per_Polygon_DIMOB_v03032026_CORRECTED.xlsx"

mz_allocation <- suppressWarnings(readxl::read_excel(mz_allocation_path))

mz_allocation <- mz_allocation %>%
rename(AREA = Area, 
CMF = CMF_unico, 
UNIVERSE = MAP_Universo, 
CP = MAP_PC)

head(mz_allocation)

# print(colnames(mz_allocation))
```

### Add CP/AREA/CMF to ento-epi data


```{r add CP CMF and Area to ento_epi_data}
# Use mz_allocation (Excel) for CP/AREA/CMF instead of shapefiles
allocation_info <- mz_allocation %>%
  select(CODIGO_DIMOB, CP, AREA, CMF, UNIVERSE) %>%
  distinct()

# Join allocation info to ento_epi_data
ento_epi_data <- ento_epi_data %>%
  left_join(allocation_info, by = "CODIGO_DIMOB") %>%
  left_join(dimob_manzanas %>% select(CODIGO_, Vivinds)%>% st_drop_geometry(), by = c("CODIGO_DIMOB" = "CODIGO_"))

head(ento_epi_data)
```

Let's check whether the new maznana shapefile has the same universe as the mz_allocation file 
```{r universe check allocation file vs shapefile}
comparison <- dimob_manzanas %>%
  select(CODIGO_, Vivinds) %>%
  left_join(mz_allocation %>% select(CODIGO_DIMOB, UNIVERSE), by = c("CODIGO_" = "CODIGO_DIMOB")) %>%
  mutate(match = Vivinds == UNIVERSE)
comparison %>% filter(!match)
library(writexl)
write_xlsx(comparison, "comparison_universe_shapefile_mzZllocationFile.xlsx")
```

### Validate Inspected_houses vs UNIVERSE

```{r}
if (all(c("Inspected_houses", "UNIVERSE") %in% colnames(ento_epi_data))) {
  # Compare Inspected_houses with UNIVERSE
  ento_epi_check <- ento_epi_data %>%
    mutate(
      diff = Inspected_houses - UNIVERSE,
      match = Inspected_houses == UNIVERSE
    )

  # Summary of matches/mismatches
  check_summary <- ento_epi_check %>%
    summarise(
      total_rows = n(),
      n_match = sum(match, na.rm = TRUE),
      n_mismatch = sum(!match, na.rm = TRUE),
      pct_mismatch = round(100 * n_mismatch / total_rows, 2)
    )
  print(check_summary)

  # Show mismatching rows
  mismatches <- ento_epi_check %>%
    filter(!match) %>%
    select(CODIGO_DIMOB, Inspected_houses, UNIVERSE, diff)
  if (nrow(mismatches) > 0) {
    print(head(mismatches, 50))
  } else {
    message("All rows match: Inspected_houses == UNIVERSE")
  }

} else {
  cat("Note: Inspected_houses or UNIVERSE not found in ento_epi_data.\n")
}

message("Mismatches alway occurring for same manzana, ",print(unique(mismatches$CODIGO_DIMOB))) #  A5Mz0 A8APSMMzFcSAMA
```

### Validate Inspected_houses vs VIVINDS
```{r}
if (all(c("Inspected_houses", "UNIVERSE") %in% colnames(ento_epi_data))) {
  # Compare Inspected_houses with UNIVERSE
  ento_epi_check <- ento_epi_data %>%
    mutate(
      diff = Inspected_houses - Vivinds,
      match = Inspected_houses == Vivinds
    )

  # Summary of matches/mismatches
  check_summary <- ento_epi_check %>%
    summarise(
      total_rows = n(),
      n_match = sum(match, na.rm = TRUE),
      n_mismatch = sum(!match, na.rm = TRUE),
      pct_mismatch = round(100 * n_mismatch / total_rows, 2)
    )
  print(check_summary)

  # Show mismatching rows
  mismatches <- ento_epi_check %>%
    filter(!match) %>%
    select(CODIGO_DIMOB, Inspected_houses, Vivinds, diff)
  if (nrow(mismatches) > 0) {
    print(head(mismatches, 50))
  } else {
    message("All rows match: Inspected_houses == UNIVERSE")
  }

} else {
  cat("Note: Inspected_houses or UNIVERSE not found in ento_epi_data.\n")
}

message("Mismatches alway occurring for same manzanas, ",cat(unique(mismatches$CODIGO_DIMOB), sep = " ")) #  
```
### Use UNIVERSE as household count

```{r substitute Inspected_houses by UNIVERSE}
ento_epi_data <- ento_epi_data %>%
  mutate(Inspected_houses = UNIVERSE) %>%
  select(-UNIVERSE, -Vivinds)
```

### Use SHAPEFILE VIVINDS as household count

```{r substitute Inspected_houses by Vivinds}
# ento_epi_data <- ento_epi_data %>%
#   mutate(Inspected_houses = Vivinds) %>%
#   select(-UNIVERSE, Vivinds)
```

```{r check manzanas with zero universe}
manzanas_without_houses <- subset(ento_epi_data, ento_epi_data$Inspected_houses == 0) # 70 unique manzanas 
manzanas_without_houses_pos <- subset(manzanas_without_houses, manzanas_without_houses$Houses_pos_IS != 0)
library(writexl)
write_xlsx(manzanas_without_houses, 
           "/media/rita/New Volume/Documenten/DI-MOB/Data Sharing/WP2_Entomological_Data/manzanas_without_houses.xlsx")

# check this against dimob_manzanas
dimob_manzanas_nohouses <- subset(dimob_manzanas, dimob_manzanas$Vivinds == 0)
dimob_manzanas_nohouses <- dimob_manzanas_nohouses %>% 
  filter(MZ != 'TB') %>%
  filter(!CODIGO_ %in% c("A7CNAMzPPESC", "DD", "A8APSMMzFcSAMA", ""))

cat("Number of unique manzanas in entomo data where Inspected_houses is zero: ",
  length(unique(manzanas_without_houses$CODIGO_DIMOB)), "\n")
cat("Number of rows that will be removed from entomological data due to zero household count: ", nrow(manzanas_without_houses), "\n")
```

### Remove zero-household manzanas

```{r remove manzanas without households}
ento_epi_data <- ento_epi_data %>%
  filter(Inspected_houses != 0)
```

## Exploratory maps

Let's try out a plot for all the Mzs with entomo data for the entire year of 2015. 

```{r plot entomo data 2015}
# aggregate to annual: sum counts, keep unique inspected_houses per block
ento_epi_year <- ento_epi_data %>%
  mutate(year = floor_date(year_month, "year")) %>%
  group_by(CODIGO_DIMOB, year) %>%
  summarise(
    Houses_pos_IS = sum(Houses_pos_IS, na.rm = TRUE),
    cases = sum(cases, na.rm = TRUE),
    Inspected_houses = first(Inspected_houses),
    .groups = "drop"
  ) %>%
  arrange(year, CODIGO_DIMOB)

# View(ento_epi_year)
```

Plot the Houses_pos_IS for 2015 per manzana
add the areas map in the background for context
```{r plot 2015 to 2019 entomo}

# Plot Houses_pos_IS for 2015–2019 on the manzana map (faceted)
plot_entomo_facets <- ento_epi_year %>%
  filter(year %in% as.Date(c("2015-01-01", "2016-01-01", "2017-01-01", "2018-01-01", "2019-01-01"))) %>%
  rename(CODIGO_ = CODIGO_DIMOB) %>%
  right_join(dimob_manzanas, by = "CODIGO_") %>%
  st_as_sf() %>%
  filter(!is.na(year)) %>%
  ggplot() +
  geom_sf(data = dimob_areas, fill = NA, color = "grey60", linewidth = 0.4) + # add the area boundaries in background
  geom_sf(aes(fill = Houses_pos_IS), color = NA) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = "Nr. of houses positive for immature stages by Consejo Popular",
       fill = "Houses_pos_IS") +
  facet_wrap(~year, ncol = 3) +
  theme_minimal()

plot_entomo_facets
```

Plot the cases for 2015 per manzana
```{r plot 2015 epi data}
plot_epi_cases <- ento_epi_year %>%
  filter(year %in% as.Date(c("2015-01-01"))) %>%
  rename(CODIGO_ = CODIGO_DIMOB) %>%
  right_join(dimob_manzanas, by = "CODIGO_") %>%
  st_as_sf() %>%
  filter(!is.na(year)) %>%
  ggplot() +
  geom_sf(data = dimob_areas, fill = NA, color = "grey60", linewidth = 0.4) +
  geom_sf(aes(fill = cases), color = NA) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = "Nr. of dengue cases by Manzana",
       fill = "Cases") +
  facet_wrap(~year, ncol = 3) +
  theme_minimal() + theme(
    axis.text = element_blank(),    # removes axis text (numbers)
    axis.title = element_blank()    # removes axis titles (lon/lat)
  )

plot_epi_cases
```

### Cases by Consejo Popular (CP)
```{r plot the CP shapefile alone}
# Quick plot of the CP shapefile
p_cp <- ggplot(dimob_cps) +
  geom_sf(fill = "lightblue", color = "black") +
  geom_sf_text(aes(label = CP), size = 2, check_overlap = TRUE) +
  labs(title = "Consejo Popular Boundaries") +
  theme_minimal()

p_cp
```

```{r plot case data for 2015 to 2019 per CP}
# Aggregate to annual by CP
ento_epi_year_cp <- ento_epi_data %>%
  mutate(year = floor_date(year_month, "year")) %>%
  group_by(CP, year) %>%
  summarise(
    Houses_pos_IS = sum(Houses_pos_IS, na.rm = TRUE),
    cases = sum(cases, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year, CP)

plot_cases_cp <- ento_epi_year_cp %>%
  filter(year %in% as.Date(c("2015-01-01", "2016-01-01", "2017-01-01", "2018-01-01", "2019-01-01"))) %>%
  mutate(year_label = format(year, "%Y")) %>%  # extract just the year
  right_join(dimob_cps, by = c("CP" = "NOMBRE"), relationship = "many-to-many") %>%
  st_as_sf() %>%
  filter(!is.na(year)) %>%
  ggplot() +
  geom_sf(data = dimob_areas, fill = NA, color = "grey60", linewidth = 0.4) +
  geom_sf(aes(fill = cases), color = NA) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = "Cases by Consejo Popular", fill = "Cases") +
  facet_wrap(~year_label, ncol = 3) +
  theme_minimal() + theme(
    axis.text = element_blank(),    # removes axis text (numbers)
    axis.title = element_blank()    # removes axis titles (lon/lat)
  )

plot_cases_cp
```

# Manzana area diagnostics
```{r average mz area}
dimob_manzanas <- dimob_manzanas %>%
  st_transform(crs = 26917)  # NAD27 / UTM zone 17N
dimob_manzanas$area_m2 <- st_area(dimob_manzanas)
average_area_m2 <- mean(as.numeric(dimob_manzanas$area_m2))
median_area_m2 <- median(as.numeric(dimob_manzanas$area_m2))

average_area_m2  # ~93693.96 m2 -> ~ 100.000m2
```

This average area per manzana seems way too large... 
```{r plot the manzanas with area}
ggplot(dimob_manzanas) +
  geom_sf(aes(fill = as.numeric(area_m2)/10000), color = "grey40", linewidth = 0.1) + # area in hectares
  geom_sf_text(aes(label = CODIGO_), size = 1.5, check_overlap = TRUE) +
  scale_fill_viridis_c(option = "plasma", name = "Area (ha)") +
  labs(title = "Manzanas by Area") +
  theme_minimal()
```

```{r histogram of manzana areas}
# Create histogram of manzana areas
ggplot(dimob_manzanas, aes(x = as.numeric(area_m2))) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Distribution of Manzana Areas (log scale)",
       x = "Area (m²)",
       y = "Frequency") +
  theme_minimal() +
  geom_vline(xintercept = average_area_m2, color = "red", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = median_area_m2, color = "blue", linetype = "dashed", linewidth = 1) +
  annotate("text", x = average_area_m2, y = Inf, 
           label = paste0("Mean: ", round(average_area_m2, 0), " m²"), 
           hjust = -0.1, vjust = 2, color = "red") +
  annotate("text", x = median_area_m2, y = Inf, 
           label = paste0("Median: ", round(median_area_m2, 0), " m²"), 
           hjust = -0.1, vjust = 4, color = "blue")
```

There is a very large variation in manzana sizes. Averaging over the manzana surface therefore doesn't seem like a safe option. 
What would make most sense -> to use the houseblock location itself (centroid) + 400m buffer around it to extract environmental data.

**Actually**, the manzanas that have classification "TB" in the "MZ" variable indicate that these are *terreno baldio*, which means emtpy terrain. 
So, these manzanas are actually not of interest for our analysis given that there will not be any mosquito measures in absence of houses (and therefore population). 

### Exclude TB and empty manzanas - already done by Sylke in new files
```{r filter manzanas without TB}
# # Keep only manzanas that do NOT have "TB" in the MZ field and have non-empty MZ names
dimob_manzanas_filtered <- dimob_manzanas %>%
  filter(!grepl("TB", MZ, ignore.case = TRUE) & CODIGO_ != "")

# cat("Original number of manzanas:", nrow(dimob_manzanas), "\n") #1480
# cat("Number of manzanas after filtering out TB:", nrow(dimob_manzanas_filtered), "\n") # 1415
# cat("Number of TB manzanas removed:", nrow(dimob_manzanas) - nrow(dimob_manzanas_filtered), "\n") # 65 removed -> 71
```

### Re-plot areas after TB filter
```{r plot the non-TB manzanas}
dimob_manzanas_filtered <- dimob_manzanas_filtered %>%
  st_transform(crs = 26917)  # NAD27 / UTM zone 17N
dimob_manzanas_filtered$area_m2 <- st_area(dimob_manzanas_filtered)
average_area_m2 <- mean(as.numeric(dimob_manzanas_filtered$area_m2))
median_area_m2 <- median(as.numeric(dimob_manzanas_filtered$area_m2))
average_area_m2  # ~29258.98 m2 -> ~ 30.000 m2

# plot the filtered manzanas shapefile
ggplot(dimob_manzanas_filtered) +
  geom_sf(aes(fill = as.numeric(area_m2)/10000), color = "grey40", linewidth = 0.1) + # area in hectares
  geom_sf_text(aes(label = CODIGO_), size = 1.5, check_overlap = TRUE) +
  scale_fill_viridis_c(option = "plasma", name = "Area (ha)") +
  labs(title = "Manzanas by Area") +
  theme_minimal()
```

```{r manzana area histogram after filtering out empty spaces}

# histogram of the manzana areas after filtering out TB manzanas
ggplot(dimob_manzanas_filtered, aes(x = as.numeric(area_m2))) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Distribution of Manzana Areas (log scale)",
       x = "Area (m²)",
       y = "Frequency") +
  theme_minimal() +
  geom_vline(xintercept = average_area_m2, color = "red", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = median_area_m2, color = "blue", linetype = "dashed", linewidth = 1) +
  annotate("text", x = average_area_m2, y = Inf, 
           label = paste0("Mean: ", round(average_area_m2, 0), " m²"), 
           hjust = -0.1, vjust = 2, color = "red") +
  annotate("text", x = median_area_m2, y = Inf, 
           label = paste0("Median: ", round(median_area_m2, 0), " m²"), 
           hjust = -0.1, vjust = 4, color = "blue")
```

However, for the larger manzanas, I wonder whether we have the exact houseblock location coordinates. 


Some manzanas are in fact huge. 


~~Final choice: use centroid per manzana + 500m buffer to extract environmental data.~~
~~This gives us an area of $\pi$*400^2 = 502.400 m2, which is larger than the average manzana size (around 30,000 m2 after removing TB manzanas).~~

## Add missing manzanas (if needed)


# Environmental data assembly

1. Create one single environmental dataset with all variables + timestamps
  - NDVI -> mazana-month
  - NDMI -> mazana-month
  - NDWI -> mazana-month
  - temperature -> month
  - precipitation -> month
  - max_precip_daily -> month
  - humidity -> month
  - wind speed -> month
  - WI -> manzana
  - WUI -> manzana
  - LandCover -> manzana
  - urbanization -> manzana
  - aljibes -> manzana 
  - water_shortage -> manzana 
  - water_containers -> consejo popular


2. Join this environmental dataset to the ento_epi_data based on year_month and manzana

```{r prepare the evironmental dataset first}
env_data_path <- "/mnt/UGent/shares/bionamix/SOFI/Env_data_cuba/CF_Municipality"

ndvi <- read_csv(file.path(env_data_path, "NDVI", "NDVI_per_manzana_nobuf_2015_2025_all_months.csv"), show_col_types = FALSE)
ndwi <- read_csv(file.path(env_data_path, "NDWI", "NDWI_per_manzana_nobuf_2015_2025_all_months.csv"), show_col_types = FALSE)
ndmi <- read_csv(file.path(env_data_path, "NDMI", "NDMI_per_manzana_nobuf_2015_2025_all_months.csv"), show_col_types = FALSE)
windspeed <- read_csv(file.path(env_data_path, "WindSpeed", "wind_speed_2015_2019_monthly.csv"), show_col_types = FALSE)
landcover <- read_csv(file.path(env_data_path, "LandCover", "LandCover_per_manzana_2021.csv"), show_col_types = FALSE)
wi <- read_csv(file.path(env_data_path, "WI", "WI_per_manzana_100mBuffer_classification.csv"), show_col_types = FALSE)
wui <- read_csv(file.path(env_data_path, "WUI", "WUI_per_manzana_NoBuffer_classification.csv"), show_col_types = FALSE)
urban <- read_csv(file.path(env_data_path, "WUI", "WSF_per_manzana_noBuffer_classification.csv"), show_col_types = FALSE)
meteo_daily <- read_xlsx("/media/rita/New Volume/Documenten/DI-MOB/Data Sharing/WP2_Meteorological_Data/Meteorologia_Mcp_Cfgos_2010-2024_pordia_v07022025.xlsx")
aljibes <- read_csv(file.path(env_data_path, "aljibes", "20m_buffer_manzanas_aljibes.csv"), show_col_types=FALSE)
water_shortage <- read_csv(file.path(env_data_path, "abasto", "no_buffer_manzanas_water_shortage.csv"), show_col_types=FALSE)
water_containers <- read_csv(file.path(env_data_path, "abasto", "avg_water_tank_by_mz.csv"), show_col_types=FALSE)
```


```{r keep relevant columns from ndvi, ndwi, and ndmi}
ndvi <- ndvi %>%
  select(manzana_id, year, month, mean_ndvi) %>%
  filter(!is.na(manzana_id) & manzana_id != "None" & manzana_id != "") %>%
  mutate(year_month = paste0(year, "_", sprintf("%02d", month))) %>%
  select(manzana_id, year_month, mean_ndvi) %>%
  distinct()  # remove any duplicates

ndmi <- ndmi %>%
  select(manzana_id, year, month, mean_ndmi) %>%
  filter(!is.na(manzana_id) & manzana_id != "None" & manzana_id != "") %>%
  mutate(year_month = paste0(year, "_", sprintf("%02d", month))) %>%
  select(manzana_id, year_month, mean_ndmi) %>%
  distinct()  

ndwi <- ndwi %>%
  select(manzana_id, year, month, mean_ndwi) %>%
  filter(!is.na(manzana_id) & manzana_id != "None" & manzana_id != "") %>%
  mutate(year_month = paste0(year, "_", sprintf("%02d", month))) %>% 
  select(manzana_id, year_month, mean_ndwi) %>%
  distinct()  
```

```{r checks}
head(ndvi)
```

```{r keep relevant columns for WI and wUI}
wi <- wi %>%
  rename(manzana_id = ID) %>%
  select(manzana_id, is_WI)
wui <- wui %>%
  rename(manzana_id = ID) %>%
  select(manzana_id, is_WUI)
urban <- urban %>% 
  rename(manzana_id = ID) %>%
  select(manzana_id, is_urban)
aljibes <- aljibes %>% 
  rename(manzana_id = CODIGO_) %>%
  select(manzana_id, has_aljibes, nr_aljibes)

water_shortage <- water_shortage %>%
  rename(
    manzana_id = CODIGO_,
    water_shortage = has_shortage) %>%
  select(manzana_id, water_shortage)

water_containers <- water_containers %>%
  rename(manzana_id = CODIGO_DIMOB) %>%
  select(manzana_id, water_containers)
```

```{r keep relevant columns for landcover}
# print(colnames(landcover))
landcover <- landcover %>%
  select(manzana_code, dominant_landcover) %>%
  rename(manzana_id = manzana_code)
```

```{r check windspeed data}
head(windspeed)
windspeed$date <- gsub(windspeed$date, pattern = "-", replacement = "_")
windspeed <- windspeed %>%
  select(!c(year, month, WS2M_min, WS2M_max, WS10M_min, WS10M_mean, WS10M_max, WS50M_min,WS50M_mean, WS50M_max)) %>%
  rename(year_month = date)
```

### Calculate Vapor Pressure Deficit (VPD) based off of RH

It represents the difference between the amount of moisture the air can hold when it is saturated (saturation vapor pressure) and the actual amount of moisture in the air (actual vapor pressure).

The vapour pressure deficit (VPD) is the absolute difference between Es and Ea and is an important metric of atmospheric moisture because it has a near linear relationship with evaporative potential. Thus, as temperature warms, for a given decrease in RH, there will be a larger increase in VPD and the amount of water stress mosquitoes experience Brown et al. (2023). see: https://pmc.ncbi.nlm.nih.gov/articles/PMC10299817

VPD is calculated as the absolute difference between saturation and actual vapor pressure: $e_a$ and $e_s$ 


$$e_s(T) = 0.6108 x exp(\frac{17.27T}{T +  237.3}),$$ 
Where T is in C and $e_s$ is in kPa. 

$$e_a = e_s x \frac{RH}{100}, $$
Where RH is in %

Finally, 

$$VPD = e_s - e_a$$

```{r calculate VPD}
meteo_daily <- read_xlsx("/media/rita/New Volume/Documenten/DI-MOB/Data Sharing/WP2_Meteorological_Data/Meteorologia_Mcp_Cfgos_2010-2024_pordia_v07022025.xlsx")
# Select relevant years
meteo_daily <- meteo_daily %>%
  filter(Ano >= 2015 & Ano <= 2019)

# select relevant columns and create year_month 
meteo_daily <- meteo_daily %>%
  rename(Temp_med = "Temp med", Temp_min = "Temp min", Temp_max = "Temp max", Humedad_r_med = "Humedad r med", Humedad_r_min = "Humedad r min", Humedad_r_max = "Humedad r max") %>%
  mutate(year_month = paste0(Ano, "_", sprintf("%02d", Mes)),
    precip_over_1 = precip_ponderada > 1) 


meteo_daily$date <- as.Date(paste0(meteo_daily$Ano, meteo_daily$Mes, meteo_daily$Dia), format = "%Y_%m_%d") 

# calculate saturation VP
meteo_daily$e_s  <- 0.6108 * exp((17.27*meteo_daily$Temp_med)/(meteo_daily$Temp_med+237.3)) 

# calculate absolute VP 
meteo_daily$e_a <- meteo_daily$e_s * (meteo_daily$Humedad_r_med/100)

# calculate VPD (kPa)
meteo_daily$VPD <- meteo_daily$e_s - meteo_daily$e_a
```

```{r plot VPD over time}
library(ggplot2)

# If year_month is in "YYYY_MM" format:
meteo_daily$date <- as.Date(paste0(meteo_daily$year_month, "_01"), format = "%Y_%m_%d")

ggplot(meteo_daily) +
  geom_line(aes(x = date, y = VPD), color="blue") +
  geom_point(aes(x = date, y = VPD), color="blue") +
  geom_line(aes(x = date, y = Humedad_r_med/100), color="red") +
  geom_point(aes(x = date, y = Humedad_r_med/100), color="red") +
  labs(x = "Month", y = "Vapor Pressure Deficit (kPa)", title = "Monthly VPD (blue) vs relative humidity (red)") +
  theme_minimal()

meteo_daily <- meteo_daily %>%
  select(-e_a, -e_s, -date)
```

### Aggregate meteorological data (daily → monthly)

```{r aggregate daily meteorological data to monthly}
meteo_monthly <- meteo_daily %>%
  group_by(year_month) %>%
  summarise(
    avg_temp = mean(Temp_med, na.rm = TRUE),
    se_avg_temp = sd(Temp_med, na.rm = TRUE) / sqrt(sum(!is.na(Temp_med))),
    min_temp = mean(Temp_min, na.rm = TRUE),
    se_min_temp = sd(Temp_min, na.rm = TRUE) / sqrt(sum(!is.na(Temp_min))),
    max_temp = mean(Temp_max, na.rm = TRUE),
    se_max_temp = sd(Temp_max, na.rm = TRUE) / sqrt(sum(!is.na(Temp_max))),
    rel_hum = mean(Humedad_r_med, na.rm = TRUE),
    se_hum = sd(Humedad_r_med, na.rm = TRUE) / sqrt(sum(!is.na(Humedad_r_med))),
    avg_VPD = mean(VPD, na.rm = TRUE),
    se_VPD = sd(VPD, na.rm = TRUE) / sqrt(sum(!is.na(VPD))),
    total_precip = sum(precip_ponderada, na.rm = TRUE),
    precip_max_day = max(precip_ponderada, na.rm = TRUE),
    consec_rainy_days = {
      runs <- rle(precip_ponderada > 1)
      ifelse(any(runs$values), max(runs$lengths[runs$values]), 0)
    },
    total_rainy_days = sum(precip_ponderada > 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(year_month >= "2016_01" & year_month <= "2019_12")
```



### Create temp x precipitation variable

Temperature categorical variable based off of the 30th and 60th percentiles for the average monthly temperature. 

```{r create temperature categories cold med warm}
# constrict temperatures to include to viable range for Aedes aegypti development 
temp_viable_range <- c(17,35) # not necessary because avg_temp is always between 21 and 28 degrees

# based on the temperature percentiles, create "cold" "med" "warm" climate categories 
# Calculate the percentiles
q <- quantile(meteo_monthly$avg_temp, probs = c(0, 0.3, 0.6, 1), na.rm = TRUE)

# switch to 2 categories instead of 3
q <- quantile(meteo_monthly$avg_temp, probs = c(0, 0.6, 1), na.rm = TRUE)
# Create the categorical variable
meteo_monthly <- meteo_monthly %>%
  mutate(
    temp_cat = cut(
      avg_temp,
      breaks = q,
      labels = c("mild", "warm"),
      include.lowest = TRUE,
      right = FALSE
    )
  )
# based on the total_precipitaion percentiles, create "wet" and "dry" climate categories

```

```{r plot categorical temp over time}

meteo_monthly$year_month_date <- as.Date(paste0(meteo_monthly$year_month, "_01"), format = "%Y_%m_%d")

# Prepare plot_long as before
temp_long <- meteo_monthly %>%
  select(year_month_date, temp_cat, avg_temp, min_temp, max_temp) %>%
  pivot_longer(
    cols = c(avg_temp, min_temp, max_temp),
    names_to = "Type",
    values_to = "Value"
  ) %>%
  mutate(Type = recode(Type, avg_temp = "Mean", min_temp = "Min", max_temp = "Max"))

se_long <- meteo_monthly %>%
  select(year_month_date, se_avg_temp, se_min_temp, se_max_temp) %>%
  pivot_longer(
    cols = c(se_avg_temp, se_min_temp, se_max_temp),
    names_to = "Type",
    values_to = "SE"
  ) %>%
  mutate(Type = recode(Type, se_avg_temp = "Mean", se_min_temp = "Min", se_max_temp = "Max"))

plot_long <- left_join(temp_long, se_long, by = c("year_month_date", "Type"))

# Calculate y-range for the plot
ymin <- min(plot_long$Value - 1.96 * plot_long$SE, na.rm = TRUE)
ymax <- max(plot_long$Value + 1.96 * plot_long$SE, na.rm = TRUE)

# Create a data frame for rectangles (one per month)
rects <- meteo_monthly %>%
  mutate(
    xmin = year_month_date - 15, # half a month before
    xmax = year_month_date + 15  # half a month after
  )

# Plot with per-month background color
p_temp <- ggplot(plot_long, aes(x = year_month_date, y = Value, color = Type, group = Type)) +
  geom_rect(
    data = rects,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = temp_cat),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  geom_ribbon(
    aes(ymin = Value - 1.96 * SE, ymax = Value + 1.96 * SE, fill = Type),
    alpha = 0.2,
    color = NA,
    show.legend = FALSE
  ) +
  scale_color_manual(
    name = "Legend",
    values = c("Mean" = "darkblue", "Max" = "red", "Min" = "orange")
  ) +
  scale_fill_manual(
    name = "Temp Category",
    # values = c("cold" = "skyblue", "med" = "khaki", "warm" = "tomato"),
    values = c("mild" = "skyblue", "warm" = "tomato"),
    guide = "legend"
  ) +
  theme_minimal() + 
  labs(
    title = "Temperature categories",
    x = "Month"
  )

print(p_temp)

ggsave(
  filename = file.path(save_images_to, "Temperature_with_categories.png"),
  plot = p_temp,
  width = 12,
  height = 6,
  dpi = 300,
  bg = "white"
)

meteo_monthly <- meteo_monthly %>%
 select(-year_month_date)
```

Precipitation categorical variable based off of total_precipitation 

```{r create precipitation categorical variable}

# plot the precipitaiton variables on one graph first 
meteo_to_plot <- meteo_monthly %>%
  mutate(year_month_date = as.Date(paste0(gsub("_", "-", as.character(year_month)), "-01")))
scale_factor <- max(meteo_to_plot$total_precip, na.rm = TRUE) / max(meteo_to_plot$consec_rainy_days, na.rm = TRUE)

p_precip <- ggplot(meteo_to_plot, aes(x = year_month_date)) +
  geom_line(aes(y = total_precip, color = "Total Precip"), size = 1) +
  geom_point(aes(y = total_precip, color = "Total Precip")) +
  geom_line(aes(y = precip_max_day, color = "Max Daily Precip"), size = 1) +
  geom_point(aes(y = precip_max_day, color = "Max Daily Precip")) +
  geom_point(aes(y = consec_rainy_days * scale_factor, color = "Consec Rainy Days")) +
  geom_line(aes(y = consec_rainy_days * scale_factor, color = "Consec Rainy Days")) +
  geom_point(aes(y = total_rainy_days * scale_factor, color = "Total Rainy Days")) +
  geom_line(aes(y = total_rainy_days * scale_factor, color = "Total Rainy Days")) +
  scale_color_manual(
    name = NULL,
    values = c(
      "Total Precip" = "darkblue",
      "Max Daily Precip" = "lightblue",
      "Consec Rainy Days" = "yellow",
      "Total Rainy Days" = "darkorange"
    )
  ) +
  scale_y_continuous(
    name = "Precipitation (mm)",
    sec.axis = sec_axis(~ . / scale_factor, name = "Days")
  ) +
  labs(
    title = "Precipitation",
    x = "Month"
  ) +
  theme_minimal() +
  theme(legend.position = "top", legend.justification = c(1, 1))
print(p_precip)
```

```{r create precipitation categories}
# divide precipitation into 3 categories based on percentiles
q <- quantile(meteo_monthly$total_precip, probs = c(0, 0.3, 0.6, 1), na.rm = TRUE)

q <- quantile(meteo_monthly$total_precip, probs = c(0, 0.6, 1), na.rm = TRUE)
# Create the categorical variable
meteo_monthly <- meteo_monthly %>%
  mutate(
    precip_cat = cut(
      total_precip,
      breaks = q,
      labels = c("dry", "wett"),
      include.lowest = TRUE,
      right = FALSE
    )
  )
print(q)
```

In summary, by using the 30th and 60th quantile, we set the limits at: 
- < 51 mm/month for dry
- 51 - 109 mm for medium 
- > 109 mm for wett

```{r plot the precipitation categories}
meteo_monthly$year_month_date <- as.Date(paste0(meteo_monthly$year_month, "_01"), format = "%Y_%m_%d")

# Create a data frame for rectangles (one per month)
rects <- meteo_monthly %>%
  mutate(
    xmin = year_month_date - 15, # half a month before
    xmax = year_month_date + 15  # half a month after
  )
scale_factor <- max(meteo_to_plot$total_precip, na.rm = TRUE) / max(meteo_to_plot$consec_rainy_days, na.rm = TRUE)

p_precip_with_cat <- ggplot(meteo_to_plot, aes(x = year_month_date)) +
  geom_rect(
    data = rects,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = -Inf,
      ymax = max(meteo_to_plot$total_precip, na.rm = TRUE),
      fill = precip_cat
    ),
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  geom_line(aes(y = total_precip, linetype = "Total Precip")) +
  geom_point(aes(y = total_precip), size=1.5) +
  geom_line(aes(y = consec_rainy_days * scale_factor, linetype = "Consec Rainy Days")) +
  geom_point(aes(y = consec_rainy_days * scale_factor), size = 1.5) +
  scale_linetype_manual(
    name = NULL,
    values = c(
      "Total Precip" = "solid",
      "Consec Rainy Days" = "dotdash"
    )
  ) +
  scale_y_continuous(
    name = "Precipitation (mm)",
    sec.axis = sec_axis(~ . / scale_factor, name = "Days")
  ) +
  labs(
    title = "Precipitation categories",
    x = "Month"
  ) +
  # scale_fill_manual(values = c("dry" = "red", "med" = "yellow", "wett" = "skyblue")) +
  scale_fill_manual(values = c("dry" = "red", "wett" = "skyblue")) +
  theme_minimal() +
  theme(legend.position = "right", legend.justification = c(1, 1))
print(p_precip_with_cat)
ggsave(
  filename = file.path(save_images_to, "Precipitation_with_categories.png"),
  plot = p_precip_with_cat,
  width = 12,
  height = 6,
  dpi = 300,
  bg = "white"
)

meteo_monthly <- meteo_monthly %>%
 select(-year_month_date)
```

Let's now look into the correlation between cat_precip and cat_temp
```{r correlation cat precip cat temp}
library(vcd)
# Assuming your variables are cat_precip and cat_temp
table_data <- table(meteo_monthly$temp_cat, meteo_monthly$precip_cat)
dimnames(table_data) <- list(
  "Temperature Category" = levels(meteo_monthly$temp_cat),
  "Precipitation Category" = levels(meteo_monthly$precip_cat)
)

png(file.path(save_images_to,"temp_precip_mosaic_plot.png"), width = 1200, height = 600)

mosaic(
  table_data,
  shade = TRUE,
  legend = TRUE,
  main = "Association between Precipitation and Temperature Categories"
)
dev.off()
```

```{r plot temp x precip categories with original values}
meteo_monthly$year_month_date <- as.Date(paste0(meteo_monthly$year_month, "_01"), format = "%Y_%m_%d")
# Combine plots with patchwork
library(patchwork)
png(file.path(save_images_to,"temp_precip_patchwork.png"), width = 1200, height = 600)

p_temp / p_precip_with_cat
dev.off()
```


**Temporal variables** 
  - temperature -> month (2016-01 -> 2019-12)
  - precipitation -> month (2016-01 -> 2019-12)
  - precip_max_day -> month (2016-01 -> 2019-12)
  - consec_rainy_days -> month 
  - humidity -> month (2016-01 -> 2019-12)
  - wind speed -> month (2016-01 -> 2019-12)

**Spatial variables**
  - WI -> manzana
  - WUI -> manzana
  - LandCover -> manzana
  - aljibes -> manzana
  - water_containers -> manzana (CP)
  - 
  - water shortage classification -> manzana

**Spatio-temporal variables**
  - NDVI -> mazana-month (2016-01 -> 2019-12)
  - NDMI -> mazana-month (2016-01 -> 2019-12)
  - NDWI -> mazana-month (2016-01 -> 2019-12)

## Build temporal variables (monthly)
```{r aggregate the temporal variables based on year_month}
environmental_temp <- meteo_monthly %>%
  left_join(windspeed, by = "year_month") %>%
  rename(WS2M = WS2M_mean)

# select only the temporal var's as of 2016-04, since the spatio-temporal variables are only available from 2016-04 onwards
environmental_temp <- environmental_temp %>%
  filter(year_month >= "2016_01" & year_month <= "2019_12")
```


## Build spatial variables (per manzana)
```{r aggregate the spatial variables based on manzana_id}
setdiff(unique(landcover$manzana_id), wi$manzana_id)  # check if all manzanas in landcover are in wi
setdiff(unique(landcover$manzana_id), wui$manzana_id) # 

environmental_spatial <- landcover %>%
  left_join(wi, by = "manzana_id") %>%
  left_join(wui, by = "manzana_id") %>%
  left_join(urban, by = "manzana_id") %>%
  left_join(aljibes, by = "manzana_id") %>%
  left_join(water_shortage, by = "manzana_id") %>%
  left_join(water_containers, by = "manzana_id")
```

## Build spatio-temporal variables (per manzana-month)
```{r aggregate the spatio-temporal variables based on manzana_id and year_month}

environmental_spatio_temp <- ndvi %>%
  left_join(ndwi, by = c("manzana_id", "year_month"), relationship = "one-to-one") %>%
  left_join(ndmi, by = c("manzana_id", "year_month"), relationship = "one-to-one")

```

## Combine temporal, spatial, and spatio-temporal data
For the temporal variables, we will repeat the same value for all manzanas within the same month, since these variables are only available at the monthly level.
Same for the spatial variables, we will repeat the same value for all months within the same manzana, since these variables are only available at the manzana level.

```{r aggregate all environmental variables into one dataset}

# Define the time range
start_month <- "2016_01"
end_month <- "2019_12"

# Get all unique manzanas from spatial data
all_manzanas <- unique(dimob_manzanas_filtered$CODIGO_)

# Create all expected months in range
all_expected_months <- seq(as.Date("2016-01-01"), as.Date("2019-12-01"), by="month")
all_months <- format(all_expected_months, "%Y_%m")

cat("Building complete grid...\n")
cat("Number of manzanas:", length(all_manzanas), "\n")
cat("Number of months (2016-04 to 2019-12):", length(all_months), "\n")
cat("Expected total rows:", length(all_manzanas) * length(all_months), "\n")

# Create complete grid starting with temporal variables (which have all months)
complete_grid <- expand.grid(
  manzana_id = all_manzanas,
  year_month = all_months,
  stringsAsFactors = FALSE
)

# Start with the complete grid and the temporal variables
env_all <- complete_grid %>%
  left_join(environmental_temp, by = "year_month")

# Join spatio-temporal variables (NDVI, NDMI, NDWI) - these may have NAs
env_all <- env_all %>%
  left_join(environmental_spatio_temp, by = c("manzana_id", "year_month"))

# Join spatial variables (landcover, WI, WUI)
env_all <- env_all %>%
  left_join(environmental_spatial, by = "manzana_id")

env_all <- env_all %>%
  rename(manzana = manzana_id, landcover = dominant_landcover) %>%
  arrange(year_month, manzana)

cat("\nFinal environmental dataset dimensions:", nrow(env_all), "x", ncol(env_all), "\n")
```

## Save environmental dataset
```{r save the final environmental dataset}
final_env_data_path <-  "/mnt/UGent/shares/bionamix/SOFI/Env_data_cuba/CF_Municipality"
final_env_data_path_1 <-  "/media/rita/New Volume/Documenten/DI-MOB/Other Data/Env_data_cuba/data"

write_csv(env_all, file.path(final_env_data_path, "all_environmental_data_per_manzana_2016_01_to_2019_12.csv"))
write_csv(env_all, file.path(final_env_data_path_1, "all_environmental_data_per_manzana_2016_01_to_2019_12.csv"))

```

# Environmental data missingness checks
```{r check final file}
env_all <- read_csv("/mnt/UGent/shares/bionamix/SOFI/Env_data_cuba/CF_Municipality/all_environmental_data_per_manzana_2016_01_to_2019_12.csv")
# Check the final result
head(env_all)
cat("\nFinal environmental dataset dimensions:", nrow(env_all), "x", ncol(env_all), "\n")
cat("\nColumn names:\n")
print(colnames(env_all))

# Check for missing values
cat("\nMissing values per column:\n")
print(colSums(is.na(env_all)))
```

There are some missing values in the environmental dataset. 
- ndvi, ndwi, ndmi -> 16174

```{r quality checks final manzanas in env data}
# let's check whether all year_month - Manzana combinations are in the env_all dataset
length(unique(dimob_manzanas_filtered$CODIGO_))
length(unique(env_all$manzana))

cat("The manzanas in our final environmental dataset are the same as the manzanas in our filtered shapefile: ", 
    all(unique(dimob_manzanas_filtered$CODIGO_) %in% unique(env_all$manzana)) & 
      all(unique(env_all$manzana) %in% unique(dimob_manzanas_filtered$CODIGO_)), "\n")

cat("All the manzanas in our final shapefile are in the environmental dataset: ", 
    all(unique(dimob_manzanas_filtered$CODIGO_) %in% unique(env_all$manzana)), "\n")
```


## Missingness by month and manzana

### Helper: missingness by month
```{r function to check missing data for any variable}

check_missing_by_month <- function(data, variable_name) {
  
  # Create a symbol for the variable
  var_sym <- sym(variable_name)
  
  # Count how much data is missing per month
  missing_summary <- data %>%
    group_by(year_month) %>%
    summarise(
      total_rows = n(),
      missing_count = sum(is.na(!!var_sym)),
      available_count = sum(!is.na(!!var_sym)),
      pct_missing = round(100 * sum(is.na(!!var_sym)) / n(), 1),
      .groups = "drop"
    ) %>%
    arrange(year_month)
  
  cat("\n=====================================\n")
  cat("MISSING DATA ANALYSIS: ", variable_name, "\n")
  cat("=====================================\n\n")
  
  # Show full table
  cat("Summary by month:\n")
  print(missing_summary)
  
  # Find months where ALL data is missing
  completely_empty <- missing_summary %>%
    filter(missing_count == total_rows)
  
  if(nrow(completely_empty) > 0) {
    cat("\n⚠️  MONTHS WITH ALL MISSING DATA:\n")
    cat(paste(completely_empty$year_month, collapse = ", "), "\n")
  } else {
    cat("\n✓ No months have completely missing data\n")
  }
  
  # Find months with some missing data
  partial_empty <- missing_summary %>%
    filter(missing_count > 0 & missing_count < total_rows)
  
  if(nrow(partial_empty) > 0) {
    cat("\n⚠️  MONTHS WITH PARTIAL MISSING DATA:\n")
    print(partial_empty)
  } else {
    cat("✓ No months have partial missing data\n")
  }
  
  cat("\n")
}
```

```{r check missing data by month for NDVI, ndmi, ndwi}
check_missing_by_month(env_all, "mean_ndvi")
check_missing_by_month(env_all, "mean_ndmi")
check_missing_by_month(env_all, "mean_ndwi")
```


## Missingness maps

```{r visualize missing data heatmap function}

visualize_missing_data <- function(data, variable_name, sample_manzanas = NULL) {
  var_sym <- sym(variable_name)

  if (!is.null(sample_manzanas)) {
    plot_data <- data %>% filter(manzana %in% sample_manzanas)
    subtitle <- paste0("(showing ", length(sample_manzanas), " manzanas)")
  } else {
    plot_data <- data
    subtitle <- paste0("(all ", length(unique(data$manzana)), " manzanas)")
  }

  plot_data <- plot_data %>% mutate(is_missing = is.na(!!var_sym))

  p <- ggplot(plot_data, aes(x = year_month, y = manzana, fill = is_missing)) +
  geom_tile(color = "white", size = 0.1, alpha = 1) +
    scale_fill_manual(
      values = c("FALSE" = "#1B5E20", "TRUE" = "#B71C1C"),  # fully opaque
      labels = c("FALSE" = "Available", "TRUE" = "Missing"),
      name = "Data Status"
    ) +
    labs(
      title = paste("Missing Data Pattern:", variable_name),
      subtitle = subtitle,
      x = "Year-Month",
      y = "Manzana"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
      axis.text.y = element_blank(),
      panel.grid = element_blank(),
      legend.position = "bottom"
    )

  return(p)
}
```

```{r visualize NDVI missing data - sample}
# Show a sample of 50 random manzanas for better visibility
set.seed(123)
sample_mz <- sample(unique(env_all$manzana), min(50, length(unique(env_all$manzana))))
visualize_missing_data(env_all, "mean_ndvi", sample_manzanas = sample_mz)
```

```{r missing data visualize ndvi, ndmi, ndwi - all manzanas}
# Or show all manzanas (will be compressed but shows overall pattern)
ndvi_plot <- visualize_missing_data(env_all, "mean_ndvi")
ggsave(
  filename = file.path(save_images_to, "missing_NDVI_block_month.png"),
  plot = ndvi_plot,
  width = 10,
  height = 6,
  dpi = 300,
  bg = "white"
)
# visualize_missing_data(env_all, "mean_ndmi")
# visualize_missing_data(env_all, "mean_ndwi")
```


## Missingness overview (all variables)

```{r missing data overview for all variables}

# Get all variable names (excluding identifiers)
all_vars <- setdiff(colnames(env_all), c("manzana", "year_month"))

# Create summary table for all variables
missing_overview <- data.frame(
  variable = all_vars,
  total_obs = nrow(env_all),
  missing_count = sapply(all_vars, function(v) sum(is.na(env_all[[v]]))),
  pct_missing = sapply(all_vars, function(v) round(100 * mean(is.na(env_all[[v]])), 2))
) %>%
  arrange(desc(pct_missing))

cat("=== MISSING DATA OVERVIEW: ALL VARIABLES ===\n\n")
print(missing_overview)

# Visualize comparison across variables
ggplot(missing_overview, aes(x = reorder(variable, pct_missing), y = pct_missing)) +
  geom_col(aes(fill = pct_missing > 0), show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "#F44336", "FALSE" = "#4CAF50")) +
  scale_y_continuous(limits = c(0, 100)) +
  coord_flip() +
  labs(
    title = "Missing Data Across All Environmental Variables",
    subtitle = paste0("Total observations: ", nrow(env_all)),
    x = "Variable",
    y = "% Missing"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank()
  )

# Identify variable categories by missingness
cat("\n\n=== VARIABLE CATEGORIES BY MISSINGNESS ===\n")
cat("Complete variables (0% missing):\n")
complete_vars <- missing_overview %>% filter(pct_missing == 0)
if(nrow(complete_vars) > 0) {
  cat(paste(complete_vars$variable, collapse = ", "), "\n")
} else {
  cat("None\n")
}

cat("\nPartially missing (0-50%):\n")
partial_vars <- missing_overview %>% filter(pct_missing > 0 & pct_missing <= 50)
if(nrow(partial_vars) > 0) {
  print(partial_vars)
} else {
  cat("None\n")
}

cat("\nMostly missing (>50%):\n")
mostly_missing_vars <- missing_overview %>% filter(pct_missing > 50)
if(nrow(mostly_missing_vars) > 0) {
  print(mostly_missing_vars)
} else {
  cat("None\n")
}
```

## Missingness heatmap (variables × months)

```{r missing data heatmap: variables x months}

# Create a summary of missing data by variable and month
missing_by_var_month <- env_all %>%
  select(year_month, all_of(all_vars)) %>%
  group_by(year_month) %>%
  summarise(across(everything(), ~sum(is.na(.))), .groups = "drop") %>%
  pivot_longer(cols = -year_month, names_to = "variable", values_to = "missing_count")

# Calculate total observations per month
n_manzanas <- length(unique(env_all$manzana))
missing_by_var_month <- missing_by_var_month %>%
  mutate(pct_missing = round(100 * missing_count / n_manzanas, 1))

# Create heatmap
ggplot(missing_by_var_month, aes(x = year_month, y = variable, fill = pct_missing)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_gradient2(
    low = "#4CAF50", 
    mid = "#FFC107",
    high = "#F44336",
    midpoint = 50,
    name = "% Missing",
    limits = c(0, 100)
  ) +
  labs(
    title = "Missing Data Heatmap: All Variables Across Time",
    subtitle = paste0("Each cell shows % of ", n_manzanas, " manzanas with missing data"),
    x = "Year-Month",
    y = "Variable"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    axis.text.y = element_text(size = 9),
    panel.grid = element_blank()
  )
ggsave(
  filename = file.path(save_images_to, "missing_env_data_heatmap.png"),
  width = 10,
  height = 6,
  dpi = 300
)
# Print which months have most complete data
cat("\n\n=== MONTHS WITH MOST COMPLETE DATA ===\n")
complete_by_month <- missing_by_var_month %>%
  group_by(year_month) %>%
  summarise(
    avg_pct_missing = round(mean(pct_missing), 1),
    n_complete_vars = sum(pct_missing == 0),
    .groups = "drop"
  ) %>%
  arrange(avg_pct_missing)

cat("Top 10 months with most complete data:\n")
print(head(complete_by_month, 10))

cat("\n\nMonths with most missing data:\n")
print(tail(complete_by_month, 10))
```

## Impute missing data for NDVI, NDMI, and NDWI

This section implements a hierarchical imputation strategy for the vegetation and moisture indices (NDVI, NDMI, NDWI). The strategy prioritizes:
1. **Temporal interpolation** - Linear interpolation when both neighboring months have data
2. **Manzana-specific climatology** - Monthly mean for each (manzana, month) if sufficient observations exist
3. **Manzana-specific climatology** - Used for all remaining gaps without both neighbors

```{r imputation function for NDVI NDMI NDWI}

impute_NDXI_index <- function(df, index_name) {
  
  cat("\n", strrep("=", 60), "\n")
  cat("Computing manzana-specific monthly climatology for", index_name, "...\n")
  cat(strrep("=", 60), "\n")
  
  # Create a copy for imputation
  df_imputed <- df %>% arrange(manzana, year_month)
  
  # Column names
  mean_col <- paste0("mean_", tolower(index_name))
  flag_col <- paste0("imputed_", tolower(index_name))
  
  # Calculate climatology: mean index for each (manzana, month) across all years
  # Extract month from year_month format (YYYY_MM)
  df_for_clim <- df_imputed %>%
    mutate(month = as.integer(substr(year_month, 6, 7))) %>%
    select(manzana, month, !!mean_col)
  
  climatology <- df_for_clim %>%
    group_by(manzana, month) %>%
    summarise(
      climatology_mean = mean(!!sym(mean_col), na.rm = TRUE),
      n_obs = sum(!is.na(!!sym(mean_col))),
      .groups = "drop"
    )
  
  cat(sprintf("  Climatology computed for %d (manzana, month) combinations\n", 
              nrow(climatology)))
  cat(sprintf("  Average observations per (manzana, month): %.1f\n", 
              mean(climatology$n_obs, na.rm = TRUE)))
  
  # Merge climatology back
  df_imputed <- df_imputed %>%
    mutate(month = as.integer(substr(year_month, 6, 7))) %>%
    left_join(climatology, by = c("manzana", "month"))
  
  # Identify missing values before imputation
  missing_mask <- is.na(df_imputed[[mean_col]])
  n_missing_before <- sum(missing_mask)
  
  # Keep track of which rows were originally missing for plotting/reporting
  df_imputed <- df_imputed %>%
    mutate(was_missing_before = is.na(!!sym(mean_col)))
  
  cat(sprintf("\nAnalyzing %d missing %s values for imputation strategy...\n", 
              n_missing_before, index_name))
  
  # Check for adjacent time points within each manzana
  # Note: year_month is in YYYY_MM format, so we can sort as character
  df_imputed <- df_imputed %>%
    arrange(manzana, year_month) %>%
    group_by(manzana) %>%
    mutate(
      prev_value = lag(!!sym(mean_col), 1),
      next_value = lead(!!sym(mean_col), 1),
      is_missing = is.na(!!sym(mean_col)),
      has_both_neighbors = is_missing & !is.na(prev_value) & !is.na(next_value),
      sufficient_clim = n_obs >= 3
    ) %>%
    ungroup()
  
  # Step 1: Interpolate where BOTH neighbors exist
  cat("\nStep 1: Interpolating where BOTH neighbors exist within manzana...\n")
  
  n_interpolated <- sum(df_imputed$has_both_neighbors, na.rm = TRUE)
  
  df_imputed <- df_imputed %>%
    mutate(
      !!mean_col := ifelse(
        has_both_neighbors,
        (prev_value + next_value) / 2,  # Linear interpolation
        !!sym(mean_col)
      )
    )
  
  cat(sprintf("  Interpolated %d missing values\n", n_interpolated))
  
  # Step 2: Use climatology for gaps without both neighbors
  cat("Step 2: Filling gaps without both neighbors using climatology...\n")
  
  # Recalculate missing mask after interpolation and fill with climatology
  df_imputed <- df_imputed %>%
    mutate(
      is_missing = is.na(!!sym(mean_col)),
      use_manzana_clim = is_missing
    )
  
  n_climatology_manzana <- sum(df_imputed$use_manzana_clim, na.rm = TRUE)
  
  df_imputed <- df_imputed %>%
    mutate(
      !!mean_col := case_when(
        use_manzana_clim ~ climatology_mean,
        TRUE ~ !!sym(mean_col)
      )
    )

  # Persistent flag for values that were imputed in this run
  df_imputed <- df_imputed %>%
    mutate(!!flag_col := was_missing_before & !is.na(!!sym(mean_col)))
  
  # Verify imputation
  n_missing_after <- sum(is.na(df_imputed[[mean_col]]))
  
  cat("\n", strrep("=", 60), "\n")
  cat(sprintf("%s Imputation Summary:\n", index_name))
  cat(sprintf("  ✓ Interpolated (both neighbors present): %d\n", n_interpolated))
  cat(sprintf("  ✓ Manzana-specific climatology: %d\n", n_climatology_manzana))
  cat(sprintf("  Total imputed: %d\n", n_interpolated + n_climatology_manzana))
  cat(sprintf("  Missing values remaining: %d\n", n_missing_after))
  cat(strrep("=", 60), "\n")
  
  # Clean up temporary columns
  df_imputed <- df_imputed %>%
    select(-c(month, climatology_mean, n_obs, prev_value, next_value, 
          is_missing, has_both_neighbors, sufficient_clim, use_manzana_clim,
          was_missing_before))
  
  cat(sprintf("\n✓ %s imputation complete. Final dataset: %d rows\n\n", 
              index_name, nrow(df_imputed)))
  
  return(df_imputed)
}
```

```{r apply imputation to NDVI}
env_all <- impute_NDXI_index(env_all, "NDVI")
```

```{r apply imputation to NDMI}
env_all <- impute_NDXI_index(env_all, "NDMI")
```

```{r apply imputation to NDWI}
env_all <- impute_NDXI_index(env_all, "NDWI")
```

```{r visualize imputation results for NDVI, NDMI, NDWI}

plot_imputation_results <- function(df, index_name, n_samples = 3) {
  
  mean_col <- paste0("mean_", tolower(index_name))
  flag_col <- paste0("imputed_", tolower(index_name))
  
  cat(sprintf("\nVisualizing %s imputation results (3 sample manzanas)...\n", index_name))
  
  # Sample manzanas
  set.seed(42)
  sample_manzanas <- sample(unique(df$manzana), 
                            size = min(n_samples, length(unique(df$manzana))), 
                            replace = FALSE)
  
  fig_list <- list()
  
  for (i in seq_along(sample_manzanas)) {
    sample_manzana <- sample_manzanas[i]
    
    # Get original data (before imputation) - we need to compare against the input
    # Since we've already imputed, we'll just show the current state
    imputed_data <- df %>%
      filter(manzana == sample_manzana) %>%
      arrange(year_month) %>%
      mutate(
        is_imputed = if (flag_col %in% names(.)) !!sym(flag_col) else FALSE,
        point_type = ifelse(is_imputed, "Imputed", "Observed"),
        year_month_date = as.Date(paste0(substr(year_month, 1, 4), "-", 
                                          substr(year_month, 6, 7), "-01"))
      )
    
    p <- ggplot(imputed_data, aes(x = year_month_date, y = !!sym(mean_col))) +
      geom_line(color = "grey60", linewidth = 1, na.rm = TRUE) +
      geom_point(aes(color = point_type), size = 2, na.rm = TRUE) +
      scale_color_manual(values = c("Observed" = "steelblue", "Imputed" = "coral3")) +
      labs(
        title = sprintf("Manzana: %s", sample_manzana),
        x = "Date",
        y = index_name,
        subtitle = sprintf("%d values, %d imputed, %d missing after imputation", 
                          nrow(imputed_data), 
                          sum(imputed_data$is_imputed, na.rm = TRUE),
                          sum(is.na(imputed_data[[mean_col]])))
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor = element_blank()
      )
    
    fig_list[[i]] <- p
  }
  
  # Combine plots
  combined_plot <- gridExtra::grid.arrange(grobs = fig_list, ncol = 1)
  
  return(combined_plot)
}

# Visualize results
library(gridExtra)

ndvi_impute_plot <- plot_imputation_results(env_all, "NDVI")
ggsave(
  filename = file.path(save_images_to, "ndvi_imputation_samples.png"),
  plot = ndvi_impute_plot,
  width = 10,
  height = 8,
  dpi = 150
)

ndmi_impute_plot <- plot_imputation_results(env_all, "NDMI")
ggsave(
  filename = file.path(save_images_to, "ndmi_imputation_samples.png"),
  plot = ndmi_impute_plot,
  width = 10,
  height = 8,
  dpi = 150
)

ndwi_impute_plot <- plot_imputation_results(env_all, "NDWI")
ggsave(
  filename = file.path(save_images_to, "ndwi_imputation_samples.png"),
  plot = ndwi_impute_plot,
  width = 10,
  height = 8,
  dpi = 150
)
```

```{r remove imputation indicator before saving dataset}
# Remove imputation flag columns before exporting/final joins
env_all <- env_all %>%
  select(-starts_with("imputed_"))
```

```{r save the environmental dataset with missing data imputed}
final_env_data_path <-  "/mnt/UGent/shares/bionamix/SOFI/Env_data_cuba/CF_Municipality"
final_env_data_path_1 <-  "/media/rita/New Volume/Documenten/DI-MOB/Other Data/Env_data_cuba/data"

write_csv(env_all, file.path(final_env_data_path, "all_environmental_data_per_manzana_2016_01_to_2019_12_imputed.csv"))
write_csv(env_all, file.path(final_env_data_path_1, "all_environmental_data_per_manzana_2016_01_to_2019_12_imputed.csv"))

```

# Merge ento-epi with environmental data

## Prepare ento-epi data for join
```{r prepare ento_epi_data for joining}

# First, check the date format and convert if needed
cat("ento_epi_data structure:\n")
str(ento_epi_data[1:5, ])

# Convert ento_epi_data year_month to character format matching env_all (YYYY_MM)
ento_epi_join <- ento_epi_data %>% 
  mutate(
    year_month = format(year_month, "%Y_%m"),  # convert Date to "YYYY_MM" format
    manzana = CODIGO_DIMOB  # rename to match env_all
  ) %>%
  filter(year_month >= "2016_01" & year_month <= "2019_12") %>%
  select(manzana, year_month, Inspected_houses, Houses_pos_IS, cases, CMF, CP, AREA) 
head(ento_epi_join)
```

## Join ento-epi with environmental data

```{r join ento_epi with environmental data}

# Filter env_all to only include months that have ento_epi data (2016-04 onwards)
final_dataset <- ento_epi_join %>%
  left_join(env_all, by = c("manzana", "year_month"))

cat("\n=== FINAL DATASET ===\n")
cat("Dimensions:", nrow(final_dataset), "x", ncol(final_dataset), "\n")
cat("\nColumn names:\n")
print(colnames(final_dataset))

cat("\nFirst few rows:\n")
head(final_dataset)

cat("\n\nMissing values in final dataset:\n")
print(colSums(is.na(final_dataset)))

cat("\n\nData type summary:\n")
cat("Entomological data (Inspected_houses, Houses_pos_IS, cases):\n")
cat("  - Inspected_houses missing:", sum(is.na(final_dataset$Inspected_houses)), "\n")
cat("  - Houses_pos_IS missing:", sum(is.na(final_dataset$Houses_pos_IS)), "\n")
cat("  - cases missing:", sum(is.na(final_dataset$cases)), "\n")

missing_ento_epi_join <- final_dataset %>%
  filter(is.na(Inspected_houses) & is.na(Houses_pos_IS) & is.na(cases))
```

## Coverage check: ento-epi vs environmental
```{r check which months/manzanas have ento_epi data}

# Show coverage of ento_epi data
ento_epi_coverage <- final_dataset %>%
  filter(!is.na(Inspected_houses)) %>%
  summarise(
    n_rows = n(),
    n_unique_manzanas = n_distinct(manzana),
    n_unique_months = n_distinct(year_month),
    month_range = paste0(min(year_month), " to ", max(year_month))
  )

cat("\n=== ENTOMOLOGICAL-EPIDEMIOLOGICAL DATA COVERAGE ===\n")
print(ento_epi_coverage)

# Show any months/manzanas in ento_epi but missing from env_all
ento_epi_not_in_env <- ento_epi_join %>%
  anti_join(env_all, by = c("manzana", "year_month"))

if(nrow(ento_epi_not_in_env) > 0) {
  cat("\n⚠️  Rows in ento_epi data NOT in environmental data:\n")
  print(ento_epi_not_in_env)
} else {
  cat("\n✓ All ento_epi data successfully joined with environmental data\n")
}
```

## Consistency check: Inspected_houses by manzana
```{r check consistency of Inspected_houses per manzana}

cat("\n=== CHECKING CONSISTENCY OF Inspected_houses ===\n\n")

# For each manzana, check if Inspected_houses has the same value across all months
inconsistent_manzanas <- final_dataset %>%
  filter(!is.na(Inspected_houses)) %>%
  group_by(manzana) %>%
  summarise(
    n_obs = n(),
    n_unique_values = n_distinct(Inspected_houses),
    min_value = min(Inspected_houses, na.rm = TRUE),
    max_value = max(Inspected_houses, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_unique_values > 1)  # Keep only those with inconsistent values

if(nrow(inconsistent_manzanas) > 0) {
  cat("⚠️  MANZANAS WITH INCONSISTENT Inspected_houses VALUES:\n")
  cat("Number of manzanas with inconsistent values:", nrow(inconsistent_manzanas), "\n\n")
  print(inconsistent_manzanas)
  
  # Show examples
  cat("\n\nExample: First manzana with inconsistent values\n")
  example_mz <- inconsistent_manzanas$manzana[1]
  cat("Manzana:", example_mz, "\n")
  print(final_dataset %>% 
          filter(manzana == example_mz & !is.na(Inspected_houses)) %>%
          select(manzana, year_month, Inspected_houses))
} else {
  cat("✓ All manzanas have CONSISTENT Inspected_houses values across months\n")
  cat("✓ This is correct - Inspected_houses is a static characteristic per manzana\n")
}
```

## Check whether Houses_Pos_IS > Inspected_houses + cases

```{r check for overly large entomological indicators}

# Create a potential maximum for inspection count
final_dataset <- final_dataset %>%
  mutate(potential_max = Inspected_houses + cases)

# Check if Houses_pos_IS exceeds the potential maximum
violations <- final_dataset %>%
  filter(!is.na(Inspected_houses) & !is.na(cases) & !is.na(Houses_pos_IS)) %>%
  filter(Houses_pos_IS > potential_max) %>%
  select(manzana, year_month, Inspected_houses, cases, potential_max, Houses_pos_IS) %>%
  mutate(excess = Houses_pos_IS - potential_max) %>%
  arrange(desc(excess))

cat("\n=== DATA QUALITY CHECK: Houses_pos_IS vs Inspected_houses + cases ===\n")
cat("Total observations with complete data:", 
    nrow(final_dataset %>% filter(!is.na(Inspected_houses) & !is.na(cases) & !is.na(Houses_pos_IS))), "\n")
cat("Observations where Houses_pos_IS > Inspected_houses + cases:", nrow(violations), "\n\n")

if(nrow(violations) > 0) {
  cat("⚠️  DATA QUALITY ISSUE DETECTED!\n")
  cat("The following rows have more positive samples than could plausibly be detected:\n\n")
  print(violations)
  
  # Create visualization
  violations_plot <- ggplot(violations, aes(x = reorder(paste(manzana, year_month), excess), y = excess)) +
    geom_col(fill = "red", alpha = 0.7) +
    coord_flip() +
    labs(
      x = "Manzana-Month",
      y = "Excess (Houses_pos_IS - Inspected_houses - cases)",
      title = "Data Quality Issues: Positive samples exceed plausible maximum",
      subtitle = paste0("Found ", nrow(violations), " problematic observations")
    ) +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 8))
  
  print(violations_plot)
  
  ggsave(
    filename = file.path(save_images_to, "data_quality_violations_Houses_pos_IS.png"),
    plot = violations_plot,
    width = 10,
    height = 8,
    dpi = 300
  )
  
  cat("\n\nCORRECTING DATA:\n")
  cat("Resetting Houses_pos_IS to maximum possible value (Inspected_houses + cases) for affected rows...\n")
  
  # Correct the violations: cap Houses_pos_IS at potential_max
  final_dataset <- final_dataset %>%
    mutate(
      Houses_pos_IS_original = Houses_pos_IS,
      Houses_pos_IS = ifelse(
        Houses_pos_IS > potential_max & !is.na(potential_max),
        potential_max,
        Houses_pos_IS
      )
    )
  
  cat("✓ Correction complete.\n")
  cat("\nCorrected rows:\n")
  corrections <- final_dataset %>%
    filter(!is.na(Houses_pos_IS_original) & Houses_pos_IS_original != Houses_pos_IS) %>%
    select(manzana, year_month, Inspected_houses, cases, potential_max, 
           Houses_pos_IS_original, Houses_pos_IS) %>%
    mutate(reduction = Houses_pos_IS_original - Houses_pos_IS)
  print(corrections)
  
  # Remove the temporary column since we've made the correction
  final_dataset <- final_dataset %>%
    select(-Houses_pos_IS_original)
  
} else {
  cat("✓ No data quality issues detected.\n")
  cat("✓ All Houses_pos_IS values are consistent with Inspected_houses + cases\n")
}

# Remove the temporary column
final_dataset <- final_dataset %>%
  select(-potential_max)
```


## Save final merged dataset
```{r save final dataset with environmental entomological and epi data}
final_epi_entomo_env_data_path <- "/media/rita/New Volume/Documenten/DI-MOB/Other Data/Env_data_cuba/data"
final_epi_entomo_env_data_path_1 <- "/mnt/UGent/shares/bionamix/SOFI/Env_data_cuba/CF_Municipality"
final_epi_entomo_env_data_path_2 <- "/home/rita/PyProjects/DI-MOB-BionamiX/data"

write_csv(final_dataset, file.path(final_epi_entomo_env_data_path, "env_epi_entomo_data_per_manzana_2016_01_to_2019_12.csv"))
write_csv(final_dataset, file.path(final_epi_entomo_env_data_path_1, "env_epi_entomo_data_per_manzana_2016_01_to_2019_12.csv"))
write_csv(final_dataset, file.path(final_epi_entomo_env_data_path_2, "env_epi_entomo_data_per_manzana_2016_01_to_2019_12.csv"))
```

# Correcting for colinnearity (check Descriptive Statistics for measures of colinnearity)
  geom_ribbon(aes(ymin = avg_temp - 1.96 * se_temp, ymax = avg_temp + 1.96 * se_temp), fill = "blue", alpha = 0.2) +

```{r create residuals for precip_max_day in linear relation to total_precip}
env_epi_ento <- read_csv("/mnt/UGent/shares/bionamix/SOFI/Env_data_cuba/CF_Municipality/env_epi_entomo_data_per_manzana_2016_01_to_2019_12.csv")

# na.exclude pads residuals with NA for missing rows so lengths match the dataframe
resid_fit <- lm(precip_max_day ~ precip_cat, data = env_epi_ento, na.action = na.exclude)

env_epi_ento <- env_epi_ento %>%
  mutate(precip_max_day_resid = resid(resid_fit))
```

# Save the final dataset 
```{r save corrected dataset with environmental entomological and epi data}
write_csv(env_epi_ento, file.path(final_epi_entomo_env_data_path, "env_epi_entomo_data_per_manzana_2016_01_to_2019_12_noColinnearity.csv"))
write_csv(env_epi_ento, file.path(final_epi_entomo_env_data_path_1, "env_epi_entomo_data_per_manzana_2016_01_to_2019_12_noColinnearity.csv"))
write_csv(env_epi_ento, file.path(final_epi_entomo_env_data_path_2, "env_epi_entomo_data_per_manzana_2016_01_to_2019_12_noColinnearity.csv"))
```