---
title: "Descriptive Statistics - Environmental Variables"
author: "Rita Verstraeten"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

# Descriptive Statistics: Environmental Variables

This document provides descriptive statistics for environmental variables used in entomological research in Cuba, organized by variable type and their spatial-temporal structure:

- **Temporal variables**: How environmental conditions vary over time (months) across the study region
- **Spatial variables**: How the study locations differ in their geographic and infrastructural characteristics (across 1,409 manzanas)
- **Spatio-temporal variables**: How vegetation and water indices vary both across space and time

|Column | Type | Description |
|---|---|---|
|manzana | String | manzana identifier (1 to B) |
|time | Integer | time period identifier (1 to T) |
|N_HH | Integer | Number of households inspected (universo) |
|Cases | Integer | Number of dengue cases observed in that block-time |
|Houses_pos_IS | Integer | Number of positive mosquito samples detected |
|n_bt | Integer | Total inspection events (N_HH + reactive adjustments) |
|ndvi | Numeric | Normalized Difference Vegetation Index for manzana-month |
|ndmi | Numeric | Normalized Difference Moisture Index for manzana-month |
|ndwi | Numeric | Normalized Difference Water Index for manzana-month |
|wS2M | Numeric | average windspeed at 2 meter height in m/s |
|landcover | categorical | dominant landcover value for that mz |
|wui | Integer | 0/1 value for wild-land urban interface for that manzana |
|wi | Integer | 0/1 value for water interface for that manzana (or within 100m of manzana) |
|urban | Integer | 0/1 value for whether/not urban for that manzana |
|has_aljibes | Binary | 0/1 value for whether/not there are underground cisternas within a 20m radius of the manzana |
|nr_aljibes | Numeric | Number of underground water cisternas within 20m of the manzana |
|avg_temp | Numeric | average temperature (°C) for that manzana-month |
|total_precip | Numeric | cummulative precipitation (mm) for that manzana-month |
|rel_hum | Numeric | average relative humidity (%) for that manzana-month |

## Setup

```{r setup, include=FALSE}
source("renv/activate.R")
library(readxl)
library(dplyr)
library(lubridate)
library(readr)
library(ggplot2)
library(tidyr)
library(corrplot)
library(rstatix)
library(cowplot)
library(gridExtra)
library(knitr)
library(sf)

knitr::opts_chunk$set(echo = FALSE, fig.width = 12, fig.height = 7, message = FALSE, warning = FALSE)

# Set up output directory for plots
output_dir <- "/home/rita/PyProjects/DI-MOB-BionamiX/results/Entomo/images/descriptive"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Load data
env_all <- read_csv("/mnt/UGent/shares/bionamix/SOFI/Env_data_cuba/CF_Municipality/all_environmental_data_per_manzana_2016_04_to_2019_12.csv")

# Load manzanas shapefile (assumed to be already loaded in environment as 'manzanas')
# If not loaded, uncomment and adjust path:
# manzanas <- st_read("path/to/manzanas.shp")

shapefile_path <- "/media/rita/New Volume/Documenten/DI-MOB/Data Sharing/WP1_Cartographic_data/Administrative borders"
manzanas <- st_read(file.path(shapefile_path, "Manzanas vectores 21_11_2024_CMF column uncleaned/manzanero_AS_1a8_19112024.shp"))
manzanas <- manzanas %>% dplyr::filter(MZ != "TB", !is.na(CODIGO_), CODIGO_ != "")
manzanas <- manzanas %>%
  filter(CODIGO_ != "A8APSMMzFcSAMA")
length(unique(manzanas$CODIGO_))

municipality <- st_read(file.path(shapefile_path, "Municipality", "MUNICIPIO CFGOS 3022025.shp"))

# Ensure year_month is recognized as date/factor
if ("year_month" %in% colnames(env_all)) {
  env_all <- env_all %>%
    mutate(year_month = as.factor(year_month))
}
```

---

## Temporal Variables

Temporal variables that vary over time and are measured consistently across all spatial units. Here we examine how environmental conditions change seasonally over the study period (April 2016 - December 2019).

### Overall Summary Statistics

```{r temporal_overall_summary}
temporal_vars <- c("WS2M", "avg_temp", "total_precip", "rel_hum")

temporal_data <- env_all %>%
  select(all_of(temporal_vars)) %>%
  drop_na()

# Create summary statistics table
temporal_summary <- temporal_data %>%
  summarise(
    across(everything(),
      list(
        n = ~n(),
        Mean = ~mean(., na.rm = TRUE),
        SD = ~sd(., na.rm = TRUE),
        Min = ~min(., na.rm = TRUE),
        Q25 = ~quantile(., probs = 0.25, na.rm = TRUE),
        Median = ~median(., na.rm = TRUE),
        Q75 = ~quantile(., probs = 0.75, na.rm = TRUE),
        Max = ~max(., na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value") %>%
  separate(Statistic, into = c("Variable", "Stat"), sep = "_(?=n$|Mean|SD|Min|Q25|Median|Q75|Max)") %>%
  pivot_wider(names_from = "Stat", values_from = "Value")

knitr::kable(temporal_summary, digits = 2, caption = "Temporal Variables - Overall Summary Statistics (Pooled across all locations and months)")
```

### Summary Statistics by Month (Year-Month)

```{r temporal_by_month_summary}
if ("year_month" %in% colnames(env_all)) {
  temporal_by_month <- env_all %>%
    select(all_of(c("year_month", temporal_vars))) %>%
    drop_na() %>%
    group_by(year_month) %>%
    summarise(
      across(all_of(temporal_vars),
        list(
          Mean = ~mean(., na.rm = TRUE),
          SD = ~sd(., na.rm = TRUE),
          Min = ~min(., na.rm = TRUE),
          Max = ~max(., na.rm = TRUE)
        ),
        .names = "{.col}_{.fn}"
      ),
      .groups = "drop"
    ) %>%
    pivot_longer(-year_month, names_to = "Stat_Var", values_to = "Value") %>%
    separate(Stat_Var, into = c("Variable", "Statistic"), sep = "_(?=Mean|SD|Min|Max)") %>%
    pivot_wider(names_from = "Statistic", values_from = "Value")
  
  knitr::kable(temporal_by_month, digits = 2, caption = "Temporal Variables - Summary by Month")
} else {
  cat("Note: year_month variable not found in dataset")
}
```

### Temporal Distributions

```{r temporal_distributions, fig.cap="Distribution of temporal variables (pooled across all locations and months)"}
temporal_data_long <- temporal_data %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")

p <- ggplot(temporal_data_long, aes(x = Value)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7, color = "black") +
  facet_wrap(~Variable, scales = "free", nrow = 2) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    axis.title = element_text(size = 10)
  ) +
  labs(title = "Distribution of Temporal Variables",
       x = "Value",
       y = "Frequency")
print(p)
ggsave(file.path(output_dir, "01_temporal_distributions.png"), p, width = 12, height = 8, dpi = 300)
```

### Temporal Variation Over Time

```{r temporal_by_month, fig.cap="Temporal variables by month - showing seasonal patterns", fig.width=14, fig.height=10}
if ("year_month" %in% colnames(env_all)) {
  temporal_with_month <- env_all %>%
    select(all_of(c("year_month", temporal_vars))) %>%
    drop_na() %>%
    pivot_longer(all_of(temporal_vars), names_to = "Variable", values_to = "Value")
  
  # Plot windspeed variables
  ws_vars <- c("WS2M")
  ws_data <- temporal_with_month %>% filter(Variable %in% ws_vars)
  
  p1 <- ggplot(ws_data, aes(x = year_month, y = Value, color = Variable, group = Variable)) +
    geom_point(size = 2, alpha = 0.7) +
    geom_line(alpha = 0.5) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 90, size = 6),
      plot.title = element_text(size = 11, face = "bold")
    ) +
    labs(title = "Wind Speed Variables",
         x = "Year-Month",
         y = "Wind Speed (m/s)",
         color = "Height (m)")
  
  # Plot individual other temporal variables
  other_vars <- setdiff(temporal_vars, ws_vars)
  
  p2 <- ggplot(temporal_with_month %>% filter(Variable == "avg_temp"), 
               aes(x = year_month, y = Value, color = Variable, group = Variable)) +
    geom_point(size = 2, alpha = 0.7, color = "firebrick") +
    geom_line(alpha = 0.5, color = "firebrick") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 90, size = 6),
      legend.position = "none",
      plot.title = element_text(size = 11, face = "bold")
    ) +
    labs(title = "Average Temperature",
         x = "Year-Month",
         y = "Temperature (°C)")
  
  p3 <- ggplot(temporal_with_month %>% filter(Variable == "total_precip"), 
               aes(x = year_month, y = Value, color = Variable, group = Variable)) +
    geom_point(size = 2, alpha = 0.7, color = "steelblue") +
    geom_line(alpha = 0.5, color = "steelblue") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 90, size = 6),
      legend.position = "none",
      plot.title = element_text(size = 11, face = "bold")
    ) +
    labs(title = "Total Precipitation",
         x = "Year-Month",
         y = "Precipitation (mm)")
  
  p4 <- ggplot(temporal_with_month %>% filter(Variable == "rel_hum"), 
               aes(x = year_month, y = Value, color = Variable, group = Variable)) +
    geom_point(size = 2, alpha = 0.7, color = "mediumseagreen") +
    geom_line(alpha = 0.5, color = "mediumseagreen") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 90, size = 6),
      legend.position = "none",
      plot.title = element_text(size = 11, face = "bold")
    ) +
    labs(title = "Relative Humidity",
         x = "Year-Month",
         y = "Humidity (%)")
  
  # Combine into 2x2 grid
  combined_plot <- cowplot::plot_grid(p1, p2, p3, p4, nrow = 2, ncol = 2)
  
  print(combined_plot)
  ggsave(file.path(output_dir, "02_temporal_by_month.png"), combined_plot, width = 14, height = 10, dpi = 300)
} else {
  cat("Note: year_month variable not found in dataset")
}
```

---

## Spatial Variables

Variables that describe the spatial characteristics of locations. With 1,409 spatial units (manzanas), we examine the geographic heterogeneity of the study region.

## Continuous Spatial Variables (Distribution across 1,409 locations)

Get an idea of the total number of houses in the database
```{r}
```

```{r spatial_continuous_summary}
continuous_spatial <- c("nr_aljibes")

# Get unique locations to understand spatial distribution
spatial_unique <- env_all %>%
  select(all_of(c("manzana", continuous_spatial))) %>%
  distinct(manzana, .keep_all = TRUE) %>%
  drop_na()

continuous_summary <- spatial_unique %>%
  select(all_of(continuous_spatial)) %>%
  summarise(
    across(everything(),
      list(
        n_locations = ~n(),
        Mean = ~mean(., na.rm = TRUE),
        SD = ~sd(., na.rm = TRUE),
        Min = ~min(., na.rm = TRUE),
        Median = ~median(., na.rm = TRUE),
        Max = ~max(., na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value") %>%
  separate(Statistic, into = c("Variable", "Stat"), sep = "_(?=n_locations|Mean|SD|Min|Median|Max)") %>%
  pivot_wider(names_from = "Stat", values_from = "Value")

knitr::kable(continuous_summary, digits = 2, caption = "Continuous Spatial Variables - Distribution Across 1,409 Manzanas")
```

## Binary Spatial Variables

```{r binary_spatial_summary}
binary_spatial <- c("is_WI", "is_WUI", "has_aljibes", "is_urban")

binary_summary <- env_all %>%
  select(all_of(binary_spatial)) %>%
  drop_na() %>%
  summarise(
    across(everything(),
      list(
        Total_n = ~n(),
        Count_TRUE = ~sum(. == TRUE, na.rm = TRUE),
        Count_FALSE = ~sum(. == FALSE, na.rm = TRUE),
        Pct_TRUE = ~sum(. == TRUE, na.rm = TRUE) / n() * 100
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value") %>%
  separate(Statistic, into = c("Variable", "Stat"), sep = "_(?=Total_n|Count_TRUE|Count_FALSE|Pct_TRUE)") %>%
  pivot_wider(names_from = "Stat", values_from = "Value")

knitr::kable(binary_summary, digits = 1, caption = "Descriptive Statistics - Binary Spatial Variables")
```

## Categorical Spatial Variables

```{r categorical_spatial_summary}
categorical_spatial <- c("landcover")

categorical_summary <- env_all %>%
  select(all_of(categorical_spatial)) %>%
  drop_na() %>%
  group_by(landcover) %>%
  summarise(
    Count = n(),
    Percentage = n() / nrow(env_all %>% select(all_of(categorical_spatial)) %>% drop_na()) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(Count))

knitr::kable(categorical_summary, digits = 1, caption = "Descriptive Statistics - Landcover (Categorical)")
```


## Spatial Variables Visualizations

```{r spatial_binary_barplot, fig.cap="Binary spatial variables frequency"}
binary_data_long <- env_all %>%
  select(all_of(binary_spatial)) %>%
  drop_na() %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable, Value) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(Percentage = paste0(round(Count / sum(Count) * 100, 1), "%"), .by = Variable)

p <- ggplot(binary_data_long, aes(x = Variable, y = Count, fill = Value)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = Percentage), position = position_dodge(width = 0.9), vjust = -0.5, size = 3.5) +
  theme_minimal() +
  labs(title = "Binary Spatial Variables Distribution among manzanas",
       x = "Variable",
       y = "Count",
       fill = "Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p)
ggsave(file.path(output_dir, "03_spatial_binary_barplot.png"), p, width = 10, height = 6, dpi = 300)
```

```{r spatial_categorical_barplot, fig.cap="Landcover distribution"}
landcover_data <- env_all %>%
  select(landcover) %>%
  drop_na() %>%
  group_by(landcover) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(Percentage = paste0(round(Count / sum(Count) * 100, 1), "%"))

p <- ggplot(landcover_data, aes(x = reorder(landcover, -Count), y = Count, fill = landcover)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Percentage), vjust = -0.5, size = 3.5) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Landcover Distribution among manzanas",
       x = "Landcover Type",
       y = "Count")
print(p)
ggsave(file.path(output_dir, "04_spatial_categorical_barplot.png"), p, width = 10, height = 6, dpi = 300)
```

### Spatial Distribution Maps

```{r spatial_maps_prep}
# Prepare spatial data - get unique manzana values for each spatial variable
spatial_map_data <- env_all %>%
  select(manzana, all_of(c(binary_spatial, continuous_spatial, categorical_spatial))) %>%
  distinct(manzana, .keep_all = TRUE)

# Join with shapefile
manzanas_spatial <- manzanas %>%
  left_join(spatial_map_data, by = c("CODIGO_" = "manzana"))
```

```{r spatial_maps_binary, fig.cap="Spatial distribution of binary variables across manzanas", fig.width=14, fig.height=12}
# Create maps for binary spatial variables
binary_plots <- list()

for (var in binary_spatial) {
  p <- ggplot() +
    geom_sf(data = municipality, fill = NA, color = "black", size = 0.8) +
    geom_sf(data = manzanas_spatial, aes(fill = !!sym(var)), color = "gray30", size = 0.1) +
    scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "gray90"),
                      na.value = "white",
                      labels = c("TRUE" = "Yes", "FALSE" = "No")) +
    theme_minimal() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(),
          plot.title = element_text(face = "bold", size = 11)) +
    labs(title = var, fill = "")
  
  binary_plots[[var]] <- p
}

# Combine into 2x2 grid
combined_binary <- cowplot::plot_grid(plotlist = binary_plots, nrow = 2, ncol = 2)
print(combined_binary)
ggsave(file.path(output_dir, "08_spatial_maps_binary.png"), combined_binary, width = 14, height = 12, dpi = 300)
```

```{r spatial_maps_continuous, fig.cap="Spatial distribution of nr_aljibes across manzanas", fig.width=10, fig.height=8}
# Map for continuous variable (nr_aljibes)
p <- ggplot() +
  geom_sf(data = municipality, fill = NA, color = "black", size = 0.8) +
  geom_sf(data = manzanas_spatial, aes(fill = nr_aljibes), color = "gray30", size = 0.1) +
  scale_fill_viridis_c(option = "viridis", na.value = "white") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 12)) +
  labs(title = "Number of Aljibes per Manzana",
       fill = "Count")

print(p)
ggsave(file.path(output_dir, "09_spatial_map_nr_aljibes.png"), p, width = 10, height = 8, dpi = 300)
```

```{r spatial_maps_categorical, fig.cap="Spatial distribution of landcover types across manzanas", fig.width=10, fig.height=8}
# Map for categorical variable (landcover)
p <- ggplot(manzanas_spatial) +
  geom_sf(data = municipality, fill = NA, color = "black", size = 0.8) +
  geom_sf(data = manzanas_spatial, aes(fill = landcover), color = "gray30", size = 0.1) +
  scale_fill_brewer(palette = "Set3", na.value = "white") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 12),
        legend.position = "bottom") +
  labs(title = "Landcover Distribution",
       fill = "Landcover Type")

print(p)
ggsave(file.path(output_dir, "10_spatial_map_landcover.png"), p, width = 10, height = 8, dpi = 300)
```

---

## Spatio-Temporal Variables

Variables that vary in both space and time, representing vegetation and water conditions across different time points and locations.

## Summary Statistics - Overall

```{r spatiotemporal_summary}
spatiotemporal_vars <- c("mean_ndvi", "mean_ndmi", "mean_ndwi")

spatiotemporal_data <- env_all %>%
  select(all_of(spatiotemporal_vars)) %>%
  drop_na()

spatiotemporal_summary <- spatiotemporal_data %>%
  summarise(
    across(everything(),
      list(
        n = ~n(),
        Mean = ~mean(., na.rm = TRUE),
        SD = ~sd(., na.rm = TRUE),
        Min = ~min(., na.rm = TRUE),
        Q25 = ~quantile(., probs = 0.25, na.rm = TRUE),
        Median = ~median(., na.rm = TRUE),
        Q75 = ~quantile(., probs = 0.75, na.rm = TRUE),
        Max = ~max(., na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value") %>%
  separate(Statistic, into = c("Variable", "Stat"), sep = "_(?=n$|Mean|SD|Min|Q25|Median|Q75|Max)") %>%
  pivot_wider(names_from = "Stat", values_from = "Value")

knitr::kable(spatiotemporal_summary, digits = 3, caption = "Spatio-Temporal Variables - Overall Summary (Pooled across all 1,409 locations and time periods)")
```

## Distributions

```{r spatiotemporal_distributions, fig.cap="Distribution of spatio-temporal variables (pooled across all locations and months)"}
spatiotemporal_data_long <- spatiotemporal_data %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")

p <- ggplot(spatiotemporal_data_long, aes(x = Value)) +
  geom_histogram(bins = 30, fill = "mediumseagreen", alpha = 0.7, color = "black") +
  facet_wrap(~Variable, scales = "free", nrow = 1) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    axis.title = element_text(size = 10)
  ) +
  labs(title = "Distribution of Spatio-Temporal Variables",
       x = "Index Value",
       y = "Frequency")
print(p)
ggsave(file.path(output_dir, "05_spatiotemporal_distributions.png"), p, width = 14, height = 6, dpi = 300)
```

### Temporal Variation by Month

```{r spatiotemporal_by_month, fig.cap="Spatio-temporal indices by month - showing seasonal patterns"}
if ("year_month" %in% colnames(env_all)) {
  spatial_temporal_by_month <- env_all %>%
    select(all_of(c("year_month", spatiotemporal_vars))) %>%
    drop_na() %>%
    pivot_longer(all_of(spatiotemporal_vars), names_to = "Index", values_to = "Value")
  
  p <- ggplot(spatial_temporal_by_month, aes(x = year_month, y = Value, fill = year_month)) +
    geom_boxplot(alpha = 0.7) +
    facet_wrap(~Index, scales = "free_y", nrow = 1) +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 90, size = 7),
      strip.text = element_text(face = "bold")
    ) +
    labs(title = "Spatio-Temporal Indices by Month - Seasonal Patterns",
         x = "Year-Month",
         y = "Index Value")
  print(p)
  ggsave(file.path(output_dir, "06_spatiotemporal_by_month.png"), p, width = 14, height = 6, dpi = 300)
} else {
  cat("Note: year_month variable not found in dataset")
}
```

### Spatial Variation by Landcover

```{r spatiotemporal_by_landcover, fig.cap="Spatio-temporal indices by landcover"}
spatial_temporal_by_cover <- env_all %>%
  select(all_of(c("landcover", spatiotemporal_vars))) %>%
  drop_na() %>%
  pivot_longer(all_of(spatiotemporal_vars), names_to = "Index", values_to = "Value")

p <- ggplot(spatial_temporal_by_cover, aes(x = landcover, y = Value, fill = landcover)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~Index, scales = "free_y", nrow = 1) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  ) +
  labs(title = "Spatio-Temporal Indices by Landcover - Spatial Heterogeneity",
       x = "Landcover Type",
       y = "Index Value",
       fill = "Landcover")
print(p)
ggsave(file.path(output_dir, "07_spatiotemporal_by_landcover.png"), p, width = 12, height = 6, dpi = 300)
```

---

## Correlations Environmental data

Quick overview of correlations between all environmental numeric variables.

### Correlations in numerical data - raw
NDVI, NDMI, NDWI, Temp, Precip, RH, Windspeed, nr of aljibes

```{r numeric_vars_correlation}
all_numeric <- c(temporal_vars, continuous_spatial, spatiotemporal_vars)

correlation_data <- env_all %>%
  select(all_of(all_numeric)) %>%
  drop_na()

corr_matrix <- cor(correlation_data, method = "spearman")

# Save to file
png(file.path(output_dir, "11a_numeric_vars_correlation.png"), width = 800, height = 800, res = 150)
corrplot(corr_matrix, 
         method = "color",
         type = "upper",
         addCoef.col = "black",
         diag = FALSE,
         tl.col = "black",
         title = "Spearman Correlation raw",
         tl.cex = 0.8)
dev.off()

# Display in notebook
corrplot(corr_matrix, 
         method = "color",
         type = "upper",
         addCoef.col = "black",
         diag = FALSE,
         tl.col = "black",
         title = "Spearman Correlation raw",
         tl.cex = 0.8)
```


### Correlations in numerical data - seasonality removed 
In order to remove seasonality, I will subtract the monthly average from every value of the numeric variables 
```{r deseasonalized_correlation, fig.width=14, fig.height=6}
all_numeric_w_seasonality <- all_numeric[all_numeric != "nr_aljibes"]

# Extract month from year_month for seasonal adjustment
env_deseason <- env_all %>%
  mutate(month = substr(year_month, 6, 7)) %>%
  select(all_of(c("month", all_numeric_w_seasonality)))

# Calculate monthly means for each variable (averaged across years)
monthly_means <- env_deseason %>%
  group_by(month) %>%
  summarise(across(all_of(all_numeric_w_seasonality), ~mean(., na.rm = TRUE)), .groups = "drop")

# Subtract monthly means to deseasonalize
env_deseason <- env_deseason %>%
  left_join(monthly_means, by = "month", suffix = c("", "_monthly_mean")) %>%
  mutate(
    across(
      all_of(all_numeric_w_seasonality),
      ~. - get(paste0(cur_column(), "_monthly_mean")),
      .names = "{.col}_deseason"
    )
  )

# Select only deseasonalized variables
deseason_vars <- paste0(all_numeric_w_seasonality, "_deseason")
correlation_deseason_data <- env_deseason %>%
  select(all_of(deseason_vars)) %>%
  drop_na()

# Calculate correlation on deseasonalized data
corr_matrix_deseason <- cor(correlation_deseason_data, method = "spearman")

# Clean up column names for plotting
colnames(corr_matrix_deseason) <- gsub("_deseason", "", colnames(corr_matrix_deseason))
rownames(corr_matrix_deseason) <- gsub("_deseason", "", rownames(corr_matrix_deseason))

# Save to file
png(file.path(output_dir, "11c_deseasonalized_correlation.png"), width = 800, height = 800, res = 150)
corrplot(corr_matrix_deseason, 
         method = "color",
         type = "upper",
         addCoef.col = "black",
         diag = FALSE,
         tl.col = "black",
         tl.cex = 0.8,
         title = "Spearman Correlation After Removing Seasonality (monthly average)",
         mar = c(0, 0, 2, 0))
dev.off()

# Display in notebook
corrplot(corr_matrix_deseason, 
         method = "color",
         type = "upper",
         addCoef.col = "black",
         diag = FALSE,
         tl.col = "black",
         tl.cex = 0.8,
         title = "Correlation After Removing Seasonality (monthly average)",
         mar = c(0, 0, 2, 0))
```

### Correlation Among Binary Spatial Variables

```{r spatial_categorical_correlation}
# Select binary spatial variables (excluding landcover as it's multi-category categorical)
binary_vars <- c("is_WI", "is_WUI", "is_urban", "has_aljibes")

# Prepare data: convert all to numeric format for correlation
spatial_binary_data <- env_all %>%
  select(all_of(binary_vars)) %>%
  drop_na() %>%
  mutate(
    across(all_of(binary_vars), ~as.numeric(.))
  )

# Calculate Spearman correlation matrix
corr_matrix_categ <- cor(spatial_binary_data, method = "spearman")

# Create correlation plot
png(file.path(output_dir, "11b_spatial_categorical_correlation.png"), width = 800, height = 800, res = 150)
corrplot(corr_matrix_categ,
         method = "color",
         type = "lower",
         addCoef.col = "black",
         diag = FALSE,
         tl.col = "black",
         tl.cex = 0.9,
         title = "Spearman Correlation - Binary Spatial Variables",
         mar = c(0, 0, 2, 0))
dev.off()

# Display in notebook
corrplot(corr_matrix_categ,
         method = "color",
         type = "lower",
         addCoef.col = "black",
         diag = FALSE,
         tl.col = "black",
         tl.cex = 0.9,
         title = "Spearman Correlation - Binary Spatial Variables",
         mar = c(0, 0, 2, 0))
```

---

# Descriptive Statistics - Entomological Indicators
```{r load the entomo_epi_env data}
# Load data
env_epi_ento <- read_csv("/mnt/UGent/shares/bionamix/SOFI/Env_data_cuba/CF_Municipality/env_epi_entomo_data_per_manzana_2016_04_to_2019_12.csv")

# Ensure year_month is recognized as date/factor
if ("year_month" %in% colnames(env_epi_ento)) {
  env_epi_ento <- env_epi_ento %>%
    mutate(year_month = as.factor(year_month))
}
```

## Part 1: Mosquito Survey Data

### Overall Positivity Rates

```{r ento_overall_positivity}
# Calculate positivity rate across all block-month combinations
# Houses_pos_IS: number of houses with immature stages per block-month
# Inspected_houses: number of houses inspected per block (constant over time)

if (all(c("Houses_pos_IS", "Inspected_houses") %in% colnames(env_epi_ento))) {
  overall_positivity <- env_epi_ento %>%
    summarise(
      Total_Block_Months = n(),
      Total_Houses_Inspected = sum(Inspected_houses, na.rm = TRUE),
      Total_Positive_Houses = sum(Houses_pos_IS, na.rm = TRUE),
      Missing_Positivity = sum(is.na(Houses_pos_IS)),
      Overall_Positivity_Rate_Pct = sum(Houses_pos_IS, na.rm = TRUE) / sum(Inspected_houses, na.rm = TRUE) * 100
    )
  
  knitr::kable(overall_positivity, digits = 2, caption = "Overall Mosquito Survey Positivity Rates Across All Block-Months")
} else {
  cat("Note: Houses_pos_IS or Inspected_houses variable not found in dataset.")
}
```

### Total Number of Unique Houses in Database

```{r ento_total_houses}
if ("Inspected_houses" %in% colnames(env_epi_ento)) {
  # Get unique manzanas and their Inspected_houses values (constant per manzana)
  unique_houses <- env_epi_ento %>%
    select(manzana, Inspected_houses) %>%
    distinct(manzana, .keep_all = TRUE) %>%
    drop_na()
  
  total_unique_houses <- sum(unique_houses$Inspected_houses, na.rm = TRUE)
  n_unique_manzanas <- nrow(unique_houses)
  
  cat("\n**Total Number of Houses in Database:**\n\n")
  cat("- Number of unique manzanas: ", n_unique_manzanas, "\n")
  cat("- Total number of inspected houses (summed across unique manzanas): ", total_unique_houses, "\n")
  cat("- Average houses per manzana: ", round(total_unique_houses / n_unique_manzanas, 1), "\n")
} else {
  cat("Note: Inspected_houses variable not found in dataset.")
}
```

### Temporal Distribution of Positivity by Month

```{r ento_temporal_positivity}
if (all(c("Houses_pos_IS", "Inspected_houses", "year_month") %in% colnames(env_epi_ento))) {
  temporal_positivity <- env_epi_ento %>%
    select(year_month, Houses_pos_IS, Inspected_houses) %>%
    drop_na() %>%
    group_by(year_month) %>%
    summarise(
      N_Block_Months = n(),
      Total_Houses_Inspected = sum(Inspected_houses, na.rm = TRUE),
      Total_Positive = sum(Houses_pos_IS, na.rm = TRUE),
      Positivity_Rate_Pct = sum(Houses_pos_IS, na.rm = TRUE) / sum(Inspected_houses, na.rm = TRUE) * 100,
      .groups = "drop"
    ) %>%
    arrange(year_month)
  
  knitr::kable(temporal_positivity, digits = 2, caption = "Mosquito Positivity Rates by Month - Seasonal Patterns")
} else {
  cat("Note: Required variables not found in dataset.")
}
```

### Temporal Trend Visualization

```{r ento_temporal_trend_plot, fig.cap="Mosquito positivity rates over time showing seasonal patterns"}
if (all(c("Houses_pos_IS", "Inspected_houses", "year_month") %in% colnames(env_epi_ento))) {
  temporal_pos_data <- env_epi_ento %>%
    select(year_month, Houses_pos_IS, Inspected_houses) %>%
    drop_na() %>%
    group_by(year_month) %>%
    summarise(
      Positivity_Rate = sum(Houses_pos_IS) / sum(Inspected_houses) * 100,
      N_Blocks = n(),
      .groups = "drop"
    )
  
  p <- ggplot(temporal_pos_data, aes(x = year_month, y = Positivity_Rate, group = 1)) +
    geom_point(size = 3, color = "darkred", alpha = 0.7) +
    geom_line(color = "darkred", alpha = 0.5, linewidth = 1) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 90, size = 8),
      plot.title = element_text(face = "bold", size = 12)
    ) +
    labs(title = "Mosquito Survey Positivity Rate Over Time",
         x = "Year-Month",
         y = "Positivity Rate (%)")
  
  print(p)
  ggsave(file.path(output_dir, "11_ento_temporal_positivity.png"), p, width = 12, height = 6, dpi = 300)
}
```

### Spatial Distribution of Positivity by Manzana

```{r ento_spatial_positivity_summary}
# Summary statistics at manzana level (aggregated across all time periods)
if (all(c("manzana", "Houses_pos_IS", "Inspected_houses") %in% colnames(env_epi_ento))) {
  manzana_positivity <- env_epi_ento %>%
    select(manzana, Houses_pos_IS, Inspected_houses) %>%
    drop_na() %>%
    group_by(manzana) %>%
    summarise(
      N_Block_Months = n(),
      Total_Houses_Inspected = sum(Inspected_houses, na.rm = TRUE),
      Total_Positive = sum(Houses_pos_IS, na.rm = TRUE),
      Positivity_Rate = sum(Houses_pos_IS, na.rm = TRUE) / sum(Inspected_houses, na.rm = TRUE) * 100,
      .groups = "drop"
    ) %>%
    arrange(desc(Positivity_Rate))
  
  # Summary of manzana-level statistics
  manzana_summary <- manzana_positivity %>%
    summarise(
      N_Manzanas = n(),
      Mean_Positivity = mean(Positivity_Rate, na.rm = TRUE),
      SD_Positivity = sd(Positivity_Rate, na.rm = TRUE),
      Min_Positivity = min(Positivity_Rate, na.rm = TRUE),
      Max_Positivity = max(Positivity_Rate, na.rm = TRUE),
      Median_Positivity = median(Positivity_Rate, na.rm = TRUE),
      Zero_Positive_Manzanas = sum(Positivity_Rate == 0, na.rm = TRUE),
      High_Transmission_Manzanas = sum(Positivity_Rate > 50, na.rm = TRUE)
    )
  
  knitr::kable(manzana_summary, digits = 2, caption = "Spatial Heterogeneity in Mosquito Positivity - Manzana Level Summary (Aggregated Across All Time Periods)")
  
  # Show top 20 highest positivity manzanas
  cat("\n\n**Top 20 Manzanas with Highest Mosquito Positivity Rates:**\n")
  knitr::kable(head(manzana_positivity, 20), digits = 2)
} else {
  cat("Note: Required variables for spatial analysis not found in dataset.")
}
```

### Positivity Distribution Across Manzanas

```{r ento_positivity_histogram, fig.cap="Distribution of mosquito positivity rates across manzanas"}
if (all(c("manzana", "Houses_pos_IS", "Inspected_houses") %in% colnames(env_epi_ento))) {
  manzana_pos_dist <- env_epi_ento %>%
    select(manzana, Houses_pos_IS, Inspected_houses) %>%
    drop_na() %>%
    group_by(manzana) %>%
    summarise(
      Positivity_Rate = sum(Houses_pos_IS) / sum(Inspected_houses) * 100,
      .groups = "drop"
    )
  
  p <- ggplot(manzana_pos_dist, aes(x = Positivity_Rate)) +
    geom_histogram(bins = 30, fill = "darkred", alpha = 0.7, color = "black") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 12)) +
    labs(title = "Distribution of Mosquito Positivity Rates Across Manzanas",
         x = "Positivity Rate (%) grouped by manzana",
         y = "Number of Manzanas")
  
  print(p)
  ggsave(file.path(output_dir, "12_ento_positivity_distribution.png"), p, width = 10, height = 6, dpi = 300)
}
```

### Spatial Map of Mosquito Positivity

```{r ento_spatial_map, fig.cap="Spatial distribution of mosquito positivity rates across manzanas - handled for zero-inflation"}
if (all(c("manzana", "Houses_pos_IS", "Inspected_houses") %in% colnames(env_epi_ento))) {
  # Calculate positivity rate per manzana (aggregated across time)
  manzana_pos_map <- env_epi_ento %>%
    select(manzana, Houses_pos_IS, Inspected_houses) %>%
    drop_na() %>%
    group_by(manzana) %>%
    summarise(
      Positivity_Rate = sum(Houses_pos_IS) / sum(Inspected_houses) * 100,
      .groups = "drop"
    )
  
  # Join with shapefile
  manzanas_ento <- manzanas %>%
    left_join(manzana_pos_map, by = c("CODIGO_" = "manzana"))
  
  # Create quantile-based categories to handle zero-inflation
  # Separate zeros and non-zeros
  manzanas_ento <- manzanas_ento %>%
    mutate(
      Positivity_Category = case_when(
        is.na(Positivity_Rate) ~ "No Data",
        Positivity_Rate == 0 ~ "Zero",
        Positivity_Rate > 0 & Positivity_Rate <= quantile(Positivity_Rate[Positivity_Rate > 0], 0.33, na.rm = TRUE) ~ "Low (1st tertile)",
        Positivity_Rate > quantile(Positivity_Rate[Positivity_Rate > 0], 0.33, na.rm = TRUE) & 
        Positivity_Rate <= quantile(Positivity_Rate[Positivity_Rate > 0], 0.67, na.rm = TRUE) ~ "Medium (2nd tertile)",
        Positivity_Rate > quantile(Positivity_Rate[Positivity_Rate > 0], 0.67, na.rm = TRUE) ~ "High (3rd tertile)",
        TRUE ~ "No Data"
      )
    )
  
  # Create map with categorical coloring
  p <- ggplot() +
    geom_sf(data = municipality, fill = NA, color = "black", linewidth = 0.8) +
    geom_sf(data = manzanas_ento, aes(fill = Positivity_Category), color = "gray40", linewidth = 0.1) +
    scale_fill_manual(
      values = c(
        "Zero" = "#f7fbff",
        "Low (1st tertile)" = "#deebf7",
        "Medium (2nd tertile)" = "#9ecae1",
        "High (3rd tertile)" = "#3182bd",
        "No Data" = "white"
      ),
      na.value = "white",
      breaks = c("Zero", "Low (1st tertile)", "Medium (2nd tertile)", "High (3rd tertile)", "No Data")
    ) +
    theme_minimal() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(face = "bold", size = 12),
      legend.position = "right",
      legend.text = element_text(size = 9)
    ) +
    labs(
      title = "Mosquito Survey Positivity Rates Across Manzanas",
      subtitle = "(Stratified by tertiles due to zero-inflation)",
      fill = "Positivity\nCategory"
    )
  
  print(p)
  ggsave(file.path(output_dir, "14_ento_spatial_map_positivity.png"), p, width = 11, height = 9, dpi = 300)
  
  # Print summary of categories for reference
  cat("\n**Positivity Category Breakdown:**\n")
  category_summary <- manzanas_ento %>%
    st_drop_geometry() %>%
    group_by(Positivity_Category) %>%
    summarise(
      Count = n(),
      Mean_Rate = mean(Positivity_Rate, na.rm = TRUE),
      Min_Rate = min(Positivity_Rate, na.rm = TRUE),
      Max_Rate = max(Positivity_Rate, na.rm = TRUE),
      .groups = "drop"
    )
  print(knitr::kable(category_summary, digits = 2))
} else {
  cat("Required columns not found in env_epi_ento dataset.\n")
}
```

### Positivity by Manzana Over Time (Spatio-Temporal)

```{r ento_positivity_by_manzana_time, fig.cap="Mosquito positivity rates by manzana and month - heatmap of all manzanas", fig.width=16, fig.height=10}
if (all(c("manzana", "year_month", "Houses_pos_IS", "Inspected_houses") %in% colnames(env_epi_ento))) {
  # Calculate positivity for each manzana-month combination
  manzana_month_positivity <- env_epi_ento %>%
    select(manzana, year_month, Houses_pos_IS, Inspected_houses) %>%
    drop_na() %>%
    group_by(manzana, year_month) %>%
    summarise(
      Positivity_Rate = sum(Houses_pos_IS) / sum(Inspected_houses) * 100,
      N_Houses_Inspected = sum(Inspected_houses),
      .groups = "drop"
    )
  
  # Get all manzanas ordered by mean positivity (for ordering in plot)
  manzana_order <- manzana_month_positivity %>%
    group_by(manzana) %>%
    summarise(Mean_Pos = mean(Positivity_Rate, na.rm = TRUE), .groups = "drop") %>%
    arrange(manzana) %>%
    pull(manzana)
  
  # Prepare data with ordered manzanas
  heatmap_data <- manzana_month_positivity %>%
    mutate(
      manzana = factor(manzana, levels = manzana_order),
      year_month = factor(year_month)
    )
  
  # Create heatmap
  p <- ggplot(heatmap_data, aes(x = manzana, y = year_month, fill = Positivity_Rate)) +
    geom_tile(color = "gray30", linewidth = 0.1) +
    scale_fill_viridis_c(option = "YlOrRd", direction = 1, na.value = "white", 
                         limits = c(0, 110), trans = "log1p") +
    theme_minimal() +
    theme(
      axis.text.x = element_blank(),
      axis.text.y = element_text(vjust = 0.5, hjust = 1, size = 8),
      plot.title = element_text(face = "bold", size = 12),
      legend.title = element_text(size = 10),
      panel.grid = element_blank()
    ) +
    labs(
      title = "Mosquito Positivity Rates Over Time - All Manzanas (Ordered by Mean Positivity)",
      x = "Manzana",
      y = "Year-Month",
      fill = "Positivity\nRate (%)"
    )
  
  print(p)
  ggsave(file.path(output_dir, "15_ento_positivity_heatmap_manzana_time.png"), p,width = 16, height = 10, dpi = 300)
}
```

### Positivity by CMF over time (spatio-temporal)

First create a unique Area_CMF indicator 
```{r create CMF indicator Area_CMF}
env_epi_ento_CMF <- env_epi_ento %>% 
  mutate(Area_CMF = paste(AREA, CMF, sep = "_"))
```

```{r ento_positivity_by_cmf_time, fig.cap="Mosquito positivity rates by CMF and month - heatmap of all CMFs", fig.width=16, fig.height=10}
if (all(c("Area_CMF", "year_month", "Houses_pos_IS", "Inspected_houses") %in% colnames(env_epi_ento_CMF))) {
  # Calculate positivity for each CMF-month combination (aggregate to CMF level)
  cmf_month_positivity <- env_epi_ento_CMF %>%
    select(Area_CMF, year_month, Houses_pos_IS, Inspected_houses) %>%
    drop_na() %>%
    group_by(Area_CMF, year_month) %>%
    summarise(
      Positivity_Rate = sum(Houses_pos_IS) / sum(Inspected_houses) * 100,
      N_Houses_Inspected = sum(Inspected_houses),
      .groups = "drop"
    )
  
  # Get all Area_CMFs ordered first by Area_CMF, then by mean positivity
  cmf_order <- cmf_month_positivity %>%
    group_by(Area_CMF) %>%
    summarise(Mean_Pos = mean(Positivity_Rate, na.rm = TRUE), .groups = "drop") %>%
    arrange(Area_CMF, desc(Mean_Pos)) %>%
    pull(Area_CMF)
  
  # Prepare data with ordered Area_CMFs
  cmf_heatmap_data <- cmf_month_positivity %>%
    mutate(
      Area_CMF = factor(Area_CMF, levels = cmf_order),
      year_month = factor(year_month)
    )
  
  # Create heatmap
  p <- ggplot(cmf_heatmap_data, aes(x = Area_CMF, y = year_month, fill = Positivity_Rate)) +
    geom_tile(color = "gray30", linewidth = 0.1) +
    scale_fill_viridis_c(option = "YlOrRd", direction = 1, na.value = "white", 
                         limits = c(0, 110), trans = "log1p") +
    theme_minimal() +
    theme(
      axis.text.y = element_text(vjust = 0.5, hjust = 1, size = 8),
      axis.text.x = element_text(angle = 90, size = 9),
      plot.title = element_text(face = "bold", size = 12),
      legend.title = element_text(size = 10),
      panel.grid = element_blank()
    ) +
    labs(
      title = "Mosquito Positivity Rates Over Time - All Area_CMFs (Ordered by Area then Mean Positivity)",
      x = "Area_CMF",
      y = "Year-Month",
      fill = "Positivity\nRate (%)"
    )
  
  print(p)
  ggsave(file.path(output_dir, "16_ento_positivity_heatmap_cmf_time.png"), p, width = 16, height = 10, dpi = 300)
} else {
  cat("Note: CMF or required variables not found in dataset.\n")
}
```

# Descriptive statistics - Cases

## Part 1: purely temporal 

```{r cases per year_month, fig.cap="Nr of dengue cases per month", fig.width=16, fig.height=10}
cases_temporal <- env_epi_ento %>%
  group_by(year_month) %>%
  summarise(Total_cases = sum(cases))

p <- ggplot(cases_temporal, aes(x = year_month, y = Total_cases, group=1)) +
  geom_point(size = 3, color = "darkred", alpha = 0.7) +
  geom_line(color = "darkred", alpha = 0.5, linewidth = 1) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, size = 8),
    plot.title = element_text(face = "bold", size = 12)
  ) +
  labs(title = "Dengue cases over time",
        x = "Year-Month",
        y = "Total dengue case count")

print(p)
ggsave(file.path(output_dir, "17_dengue_cases_temp.png"), p, width = 16, height = 10, dpi = 300)

```

### Dengue Cases by Manzana Over Time (Spatio-Temporal Heatmap)

```{r dengue_cases_by_manzana_time, fig.cap="Dengue cases by manzana and month - heatmap of all manzanas", fig.width=16, fig.height=10}
if (all(c("manzana", "year_month", "cases") %in% colnames(env_epi_ento))) {
  # Calculate total cases for each manzana-month combination
  manzana_month_cases <- env_epi_ento %>%
    select(manzana, year_month, cases) %>%
    drop_na() %>%
    group_by(manzana, year_month) %>%
    summarise(
      Total_Cases = sum(cases),
      .groups = "drop"
    )
  
  # Get all manzanas ordered by mean case count
  manzana_order <- manzana_month_cases %>%
    group_by(manzana) %>%
    summarise(Mean_Cases = mean(Total_Cases, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(Mean_Cases)) %>%
    pull(manzana)
  
  # Prepare data with ordered manzanas
  heatmap_cases_data <- manzana_month_cases %>%
    mutate(
      manzana = factor(manzana, levels = manzana_order),
      year_month = factor(year_month)
    )
  
  # Create heatmap
  p <- ggplot(heatmap_cases_data, aes(x = manzana, y = year_month, fill = Total_Cases)) +
    geom_tile(color = "gray30", linewidth = 0.1) +
    scale_fill_viridis_c(option = "YlOrRd", direction = 1, na.value = "white", 
                         trans = "log1p") +
    theme_minimal() +
    theme(
      axis.text.x = element_blank(),
      axis.text.y = element_text(vjust = 0.5, hjust = 1, size = 8),
      plot.title = element_text(face = "bold", size = 12),
      legend.title = element_text(size = 10),
      panel.grid = element_blank()
    ) +
    labs(
      title = "Dengue Cases Over Time - All Manzanas (Ordered by Mean Case Count)",
      x = "Manzana",
      y = "Year-Month",
      fill = "Total Cases"
    )
  
  print(p)
  ggsave(file.path(output_dir, "18_dengue_cases_heatmap_manzana_time.png"), p, width = 16, height = 10, dpi = 300)
} else {
  cat("Note: Required columns not found in dataset.\n")
}
```