---
title: "Descriptive Statistics - Environmental Variables"
author: "Rita Verstraeten"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

# Descriptive Statistics: Environmental Variables

This document provides descriptive statistics for environmental variables used in entomological research in Cuba, organized by variable type and their spatial-temporal structure:

- **Temporal variables**: How environmental conditions vary over time (months) across the study region
- **Spatial variables**: How the study locations differ in their geographic and infrastructural characteristics (across 1,409 manzanas)
- **Spatio-temporal variables**: How vegetation and water indices vary both across space and time

|Column | Type | Description |
|---|---|---|
|manzana | String | manzana identifier (1 to B) |
|time | Integer | time period identifier (1 to T) |
|N_HH | Integer | Number of households inspected (universo) |
|Cases | Integer | Number of dengue cases observed in that block-time |
|Houses_pos_IS | Integer | Number of positive mosquito samples detected |
|n_bt | Integer | Total inspection events (N_HH + reactive adjustments) |
|ndvi | Numeric | Normalized Difference Vegetation Index for manzana-month |
|ndmi | Numeric | Normalized Difference Moisture Index for manzana-month |
|ndwi | Numeric | Normalized Difference Water Index for manzana-month |
|wS2M | Numeric | average windspeed at 2 meter height in m/s |
|landcover | categorical | dominant landcover value for that mz |
|wui | Integer | TRUE/FALSE value for wild-land urban interface for that manzana |
|wi | Integer | TRUE/FALSE value for water interface for that manzana (or within 100m of manzana) |
|urban | Integer | TRUE/FALSE value for whether/not urban for that manzana |
|has_aljibes | Binary | TRUE/FALSE value for whether/not there are underground cisternas within a 20m radius of the manzana |
|nr_aljibes | Numeric | Number of underground water cisternas within 20m of the manzana |
|water_shortage | Binary | TRUE/FALSE value for whether/not manzanas have reduced water provision (more days between water provision cycles) |
|water_containers | Count | Nnumber of water containers available per manzana |
|avg_temp | Numeric | average temperature (°C) for that manzana-month |
|total_precip | Numeric | cummulative precipitation (mm) for that manzana-month | max_precip_daily
|precip_max_day | Numeric | cummulative precipitation (mm/24h) for the day in that manzana-month with the highest precipitation | 
|rel_hum | Numeric | average relative humidity (%) for that manzana-month |

## Setup

```{r setup, include=FALSE}
source("renv/activate.R")
library(readxl)
library(dplyr)
library(lubridate)
library(readr)
library(ggplot2)
library(tidyr)
library(corrplot)
library(rstatix)
library(cowplot)
library(gridExtra)
library(knitr)
library(sf)

knitr::opts_chunk$set(echo = FALSE, fig.width = 12, fig.height = 7, message = FALSE, warning = FALSE)

# Set up output directory for plots
output_dir <- "/home/rita/PyProjects/DI-MOB-BionamiX/results/Entomo/images/descriptive"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# # Load data
# env_all <- read_csv("/mnt/UGent/shares/bionamix/SOFI/Env_data_cuba/CF_Municipality/all_environmental_data_per_manzana_2016_01_to_2019_12_imputed.csv")

env_all <- read_csv("/media/rita/New Volume/Documenten/DI-MOB/Other Data/Env_data_cuba/data/all_environmental_data_per_manzana_2016_01_to_2019_12_imputed.csv")

# Load manzanas shapefile (assumed to be already loaded in environment as 'manzanas')
# If not loaded, uncomment and adjust path:
# manzanas <- st_read("path/to/manzanas.shp")

shapefile_path <- "/media/rita/New Volume/Documenten/DI-MOB/Data Sharing/WP1_Cartographic_data/Administrative borders"
manzanas <- st_read(file.path(shapefile_path, "Manzanas_cleaned_05032026", "Mz_CMF_Correcto_2022026.shp"))
manzanas <- manzanas %>% dplyr::filter(MZ != "TB", !is.na(CODIGO_), CODIGO_ != "")
manzanas <- manzanas %>%
  filter(CODIGO_ != "A8APSMMzFcSAMA")
length(unique(manzanas$CODIGO_))

municipality <- st_read(file.path(shapefile_path, "Municipality", "MUNICIPIO CFGOS 3022025.shp"))

# Ensure year_month is recognized as date/factor
if ("year_month" %in% colnames(env_all)) {
  env_all <- env_all %>%
    mutate(year_month = as.factor(year_month))
}
```

---

## Temporal Variables

Temporal variables that vary over time and are measured consistently across all spatial units. Here we examine how environmental conditions change seasonally over the study period (April 2016 - December 2019).

### Overall Summary Statistics

```{r temporal_overall_summary}
temporal_vars <- c("WS2M", "avg_temp", "total_precip", "precip_max_day", "consec_rainy_days", "rel_hum")

temporal_data <- env_all %>%
  select(all_of(temporal_vars)) %>%
  drop_na()

# Create summary statistics table
temporal_summary <- temporal_data %>%
  summarise(
    across(everything(),
      list(
        n = ~n(),
        Mean = ~mean(., na.rm = TRUE),
        SD = ~sd(., na.rm = TRUE),
        Min = ~min(., na.rm = TRUE),
        Q25 = ~quantile(., probs = 0.25, na.rm = TRUE),
        Median = ~median(., na.rm = TRUE),
        Q75 = ~quantile(., probs = 0.75, na.rm = TRUE),
        Max = ~max(., na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value") %>%
  separate(Statistic, into = c("Variable", "Stat"), sep = "_(?=n$|Mean|SD|Min|Q25|Median|Q75|Max)") %>%
  pivot_wider(names_from = "Stat", values_from = "Value")

knitr::kable(temporal_summary, digits = 2, caption = "Temporal Variables - Overall Summary Statistics (Pooled across all locations and months)")
```

### Summary Statistics by Month (Year-Month)

```{r temporal_by_month_summary}
if ("year_month" %in% colnames(env_all)) {
  temporal_by_month <- env_all %>%
    select(all_of(c("year_month", temporal_vars))) %>%
    drop_na() %>%
    group_by(year_month) %>%
    summarise(
      across(all_of(temporal_vars),
        list(
          Mean = ~mean(., na.rm = TRUE),
          SD = ~sd(., na.rm = TRUE),
          Min = ~min(., na.rm = TRUE),
          Max = ~max(., na.rm = TRUE)
        ),
        .names = "{.col}_{.fn}"
      ),
      .groups = "drop"
    ) %>%
    pivot_longer(-year_month, names_to = "Stat_Var", values_to = "Value") %>%
    separate(Stat_Var, into = c("Variable", "Statistic"), sep = "_(?=Mean|SD|Min|Max)") %>%
    pivot_wider(names_from = "Statistic", values_from = "Value")
  
  knitr::kable(temporal_by_month, digits = 2, caption = "Temporal Variables - Summary by Month")
} else {
  cat("Note: year_month variable not found in dataset")
}
```

### Temporal Distributions

```{r temporal_distributions, fig.cap="Distribution of temporal variables (pooled across all locations and months)"}
temporal_data_long <- temporal_data %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")

p <- ggplot(temporal_data_long, aes(x = Value)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7, color = "black") +
  facet_wrap(~Variable, scales = "free", nrow = 2) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    axis.title = element_text(size = 10)
  ) +
  labs(title = "Distribution of Temporal Variables",
       x = "Value",
       y = "Frequency")
print(p)
ggsave(file.path(output_dir, "01_temporal_distributions.png"), p, width = 12, height = 8, dpi = 300)
```

### Temporal Variation Over Time

```{r temporal_by_month, fig.cap="Temporal variables by month - showing seasonal patterns", fig.width=14, fig.height=10}

meteo_to_plot <- env_all %>%
  mutate(year_month_date = as.Date(paste0(gsub("_", "-", as.character(year_month)), "-01")))

# Plot 1: avg_temp with 95% CI
p1 <- ggplot(meteo_to_plot, aes(x = year_month_date)) +
  geom_line(aes(y = max_temp), color = "red") +
  geom_point(aes(y = max_temp), color = "red") +
  geom_ribbon(aes(ymin = max_temp - 1.96 * se_max_temp, ymax = max_temp + 1.96 * se_max_temp), fill = "red", alpha = 0.2) +
  
  geom_line(aes(y = avg_temp), color = "orange") +
  geom_point(aes(y = avg_temp), color = "orange") +
  geom_ribbon(aes(ymin = avg_temp - 1.96 * se_avg_temp, ymax = avg_temp + 1.96 * se_avg_temp), fill = "orange", alpha = 0.2) +

  geom_line(aes(y = min_temp), color = "blue") +
  geom_point(aes(y = min_temp), color = "blue") +
  geom_ribbon(aes(ymin = min_temp - 1.96 * se_min_temp, ymax = min_temp + 1.96 * se_min_temp), fill = "blue", alpha = 0.2) +
  labs(title = "Average Temperature (95% CI)", x = "Month", y = "Avg Temp") +
  theme_minimal()

# Plot 2: rel_hum with 95% CI
p2 <- ggplot(meteo_to_plot, aes(x = year_month_date, y = rel_hum)) +
  geom_line(color = "green") +
  geom_point(color = "green") +
  geom_ribbon(aes(ymin = rel_hum - 1.96 * se_hum, ymax = rel_hum + 1.96 * se_hum), fill = "green", alpha = 0.2) +
  labs(title = "Relative Humidity (95% CI)", x = "Month", y = "Rel Humidity") +
  theme_minimal()

# Plot 3: WS2M (wind speed)
p3 <- ggplot(meteo_to_plot, aes(x = year_month_date, y = WS2M)) +
  geom_line(color = "purple") +
  geom_point(color = "purple") +
  labs(title = "Wind Speed", x = "Month", y = "WS2M") +
  theme_minimal()

# Plot 4: total_precip
scale_factor <- max(meteo_to_plot$total_precip, na.rm = TRUE) / max(meteo_to_plot$consec_rainy_days, na.rm = TRUE)

p4 <- ggplot(meteo_to_plot, aes(x = year_month_date)) +
  geom_line(aes(y = total_precip, color = "Total Precip"), size = 1) +
  geom_point(aes(y = total_precip, color = "Total Precip")) +
  geom_line(aes(y = precip_max_day, color = "Max Daily Precip"), size = 1) +
  geom_point(aes(y = precip_max_day, color = "Max Daily Precip")) +
  geom_point(aes(y = consec_rainy_days * scale_factor, color = "Consec Rainy Days")) +
  geom_line(aes(y = consec_rainy_days * scale_factor, color = "Consec Rainy Days")) +

  scale_color_manual(
    name = "Legend",
    values = c(
      "Total Precip" = "darkblue",
      "Max Daily Precip" = "lightblue",
      "Consec Rainy Days" = "orange"
    )
  ) +
  scale_y_continuous(
    name = "Precipitation (mm)",
    sec.axis = sec_axis(~ . / scale_factor, name = "Days")
  ) +
  labs(
    title = "Precipitation",
    x = "Month"
  ) +
  theme_minimal() +
  theme(legend.position = "top", legend.justification = c(1, 1))


# print(p4)
# Combine into 2x2 grid
combined_plot <- plot_grid(p1, p2, p3, p4, nrow = 2, ncol = 2)
print(combined_plot)

ggsave(file.path(output_dir, "02_temporal_by_month.png"), combined_plot, width = 14, height = 8, dpi = 300)
ggsave(file.path(output_dir, "02a_temperature_by_month.png"), p1, width = 14, height = 8, dpi = 300)
ggsave(file.path(output_dir, "02b_precipitation_by_month.png"), p4, width = 14, height = 8, dpi = 300)
ggsave(file.path(output_dir, "02c_rel_hum_by_month.png"), p2, width = 14, height = 8, dpi = 300)
```

---

## Spatial Variables

Variables that describe the spatial characteristics of locations. With 1,409 spatial units (manzanas), we examine the geographic heterogeneity of the study region.

## Continuous Spatial Variables (Distribution across 1,409 locations)


```{r spatial_continuous_summary}
continuous_spatial <- c("nr_aljibes", "water_containers")

# Get unique locations to understand spatial distribution
spatial_unique <- env_all %>%
  select(all_of(c("manzana", continuous_spatial))) %>%
  distinct(manzana, .keep_all = TRUE) %>%
  drop_na()

continuous_summary <- spatial_unique %>%
  select(all_of(continuous_spatial)) %>%
  summarise(
    across(everything(),
      list(
        n_locations = ~n(),
        Mean = ~mean(., na.rm = TRUE),
        SD = ~sd(., na.rm = TRUE),
        Min = ~min(., na.rm = TRUE),
        Median = ~median(., na.rm = TRUE),
        Max = ~max(., na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value") %>%
  separate(Statistic, into = c("Variable", "Stat"), sep = "_(?=n_locations|Mean|SD|Min|Median|Max)") %>%
  pivot_wider(names_from = "Stat", values_from = "Value")

knitr::kable(continuous_summary, digits = 2, caption = "Continuous Spatial Variables - Distribution Across 1,409 Manzanas")
```

## Binary Spatial Variables

```{r binary_spatial_summary}
binary_spatial <- c("is_WI", "is_WUI", "is_urban", "has_aljibes", "water_shortage")

binary_summary <- env_all %>%
  select(all_of(binary_spatial)) %>%
  drop_na() %>%
  summarise(
    across(everything(),
      list(
        Total_n = ~n(),
        Count_TRUE = ~sum(. == TRUE, na.rm = TRUE),
        Count_FALSE = ~sum(. == FALSE, na.rm = TRUE),
        Pct_TRUE = ~sum(. == TRUE, na.rm = TRUE) / n() * 100
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value") %>%
  separate(Statistic, into = c("Variable", "Stat"), sep = "_(?=Total_n|Count_TRUE|Count_FALSE|Pct_TRUE)") %>%
  pivot_wider(names_from = "Stat", values_from = "Value")

knitr::kable(binary_summary, digits = 1, caption = "Descriptive Statistics - Binary Spatial Variables")
```

## Categorical Spatial Variables

```{r categorical_spatial_summary}
categorical_spatial <- c("landcover")

categorical_summary <- env_all %>%
  select(all_of(categorical_spatial)) %>%
  drop_na() %>%
  group_by(landcover) %>%
  summarise(
    Count = n(),
    Percentage = n() / nrow(env_all %>% select(all_of(categorical_spatial)) %>% drop_na()) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(Count))

knitr::kable(categorical_summary, digits = 1, caption = "Descriptive Statistics - Landcover (Categorical)")
```


## Spatial Variables Visualizations

```{r spatial_binary_barplot, fig.cap="Binary spatial variables frequency"}
binary_data_long <- env_all %>%
  select(all_of(binary_spatial)) %>%
  drop_na() %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable, Value) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(Percentage = paste0(round(Count / sum(Count) * 100, 1), "%"), .by = Variable)

p <- ggplot(binary_data_long, aes(x = Variable, y = Count, fill = Value)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = Percentage), position = position_dodge(width = 0.9), vjust = -0.5, size = 3.5) +
  theme_minimal() +
  labs(title = "Binary Spatial Variables Distribution among manzanas",
       x = "Variable",
       y = "Count",
       fill = "Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p)
ggsave(file.path(output_dir, "03_spatial_binary_barplot.png"), p, width = 10, height = 6, dpi = 300)
```

```{r spatial_categorical_barplot, fig.cap="Landcover distribution"}
landcover_data <- env_all %>%
  select(landcover) %>%
  drop_na() %>%
  group_by(landcover) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(Percentage = paste0(round(Count / sum(Count) * 100, 1), "%"))

p <- ggplot(landcover_data, aes(x = reorder(landcover, -Count), y = Count, fill = landcover)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Percentage), vjust = -0.5, size = 3.5) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Landcover Distribution among manzanas",
       x = "Landcover Type",
       y = "Count")
print(p)
ggsave(file.path(output_dir, "04_spatial_categorical_barplot.png"), p, width = 12, height = 6, dpi = 300)
```

### Spatial Distribution Maps

```{r spatial_maps_prep}
# Prepare spatial data - get unique manzana values for each spatial variable
spatial_map_data <- env_all %>%
  select(manzana, all_of(c(binary_spatial, continuous_spatial, categorical_spatial))) %>%
  distinct(manzana, .keep_all = TRUE)

# Join with shapefile
manzanas_spatial <- manzanas %>%
  left_join(spatial_map_data, by = c("CODIGO_" = "manzana"))
```

```{r spatial_maps_binary, fig.cap="Spatial distribution of binary variables across manzanas", fig.width=14, fig.height=12}
# Create maps for binary spatial variables
binary_plots <- list()

for (var in binary_spatial) {
  p <- ggplot() +
    geom_sf(data = municipality, fill = NA, color = "black", size = 0.8) +
    geom_sf(data = manzanas_spatial, aes(fill = !!sym(var)), color = "gray30", size = 0.1) +
    scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "gray90"),
                      na.value = "white",
                      labels = c("TRUE" = "Yes", "FALSE" = "No")) +
    theme_minimal() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(),
          plot.title = element_text(face = "bold", size = 11)) +
    labs(title = var, fill = "")
  
  binary_plots[[var]] <- p
}

# Combine into 2x2 grid
combined_binary <- cowplot::plot_grid(plotlist = binary_plots, nrow = 2, ncol = 3)
print(combined_binary)
ggsave(file.path(output_dir, "08_spatial_maps_binary.png"), combined_binary, width = 20, height = 10, dpi = 300)
```

```{r spatial_maps_continuous, fig.cap="Spatial distribution of nr_aljibes across manzanas", fig.width=10, fig.height=8}
# Map for continuous variable (nr_aljibes)
p <- ggplot() +
  geom_sf(data = municipality, fill = NA, color = "black", size = 0.8) +
  # Manzanas without aljibes (nr_aljibes == 0) in grey
  geom_sf(data = manzanas_spatial %>% filter(!is.na(nr_aljibes) & nr_aljibes == 0),
    fill = "gray70", color = "gray30", size = 0.1) +
  # Manzanas with aljibes (nr_aljibes > 0) on continuous scale
  geom_sf(data = manzanas_spatial %>% filter(!is.na(nr_aljibes) & nr_aljibes > 0),
    aes(fill = nr_aljibes), color = "gray30", size = 0.1) +
  scale_fill_viridis_c(option = "viridis", na.value = "white") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 12)) +
  labs(title = "Number of Aljibes per Manzana",
       fill = "Count")

print(p)
ggsave(file.path(output_dir, "09_spatial_map_nr_aljibes.png"), p, width = 10, height = 8, dpi = 300)
```

```{r spatial_maps_continuous water containers, fig.cap="Spatial distribution of nr of water containers across manzanas", fig.width=10, fig.height=8}
# Map for continuous variable (nr_aljibes)
p <- ggplot() +
  geom_sf(data = municipality, fill = NA, color = "black", size = 0.8) +
  # Manzanas without aljibes (nr_aljibes == 0) in grey
  geom_sf(data = manzanas_spatial %>% filter(!is.na(water_containers)),
    fill = "gray70", color = "gray30", size = 0.1) +
  # Manzanas with aljibes (nr_aljibes > 0) on continuous scale
  geom_sf(data = manzanas_spatial %>% filter(!is.na(water_containers)),
    aes(fill = log1p(water_containers)), color = "gray30", size = 0.1) +
  scale_fill_viridis_c(option = "viridis", na.value = "white") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 12)) +
  labs(title = "Number of water storage containers per Manzana",
       fill = "Log of count")

print(p)
ggsave(file.path(output_dir, "10_spatial_map_nr_water_containers.png"), p, width = 10, height = 8, dpi = 300)
```
```{r spatial_maps_categorical, fig.cap="Spatial distribution of landcover types across manzanas", fig.width=10, fig.height=8}
# Map for categorical variable (landcover)
p <- ggplot(manzanas_spatial) +
  geom_sf(data = municipality, fill = NA, color = "black", size = 0.8) +
  geom_sf(data = manzanas_spatial, aes(fill = landcover), color = "gray30", size = 0.1) +
  scale_fill_brewer(palette = "Set3", na.value = "white") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 12),
        legend.position = "bottom") +
  labs(title = "Landcover Distribution",
       fill = "Landcover Type")

print(p)
ggsave(file.path(output_dir, "11_spatial_map_landcover.png"), p, width = 10, height = 8, dpi = 300)
```

---

## Spatio-Temporal Variables

## Summary Statistics - Overall

Variables that vary in both space and time, representing vegetation and water conditions across different time points and locations.

## Summary Statistics - Overall

```{r spatiotemporal_summary ndxi}
spatiotemporal_vars <- c("mean_ndvi", "mean_ndmi", "mean_ndwi")

spatiotemporal_data <- env_all %>%
  select(all_of(spatiotemporal_vars)) %>%
  drop_na()

spatiotemporal_summary <- spatiotemporal_data %>%
  summarise(
    across(everything(),
      list(
        n = ~n(),
        Mean = ~mean(., na.rm = TRUE),
        SD = ~sd(., na.rm = TRUE),
        Min = ~min(., na.rm = TRUE),
        Q25 = ~quantile(., probs = 0.25, na.rm = TRUE),
        Median = ~median(., na.rm = TRUE),
        Q75 = ~quantile(., probs = 0.75, na.rm = TRUE),
        Max = ~max(., na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value") %>%
  separate(Statistic, into = c("Variable", "Stat"), sep = "_(?=n$|Mean|SD|Min|Q25|Median|Q75|Max)") %>%
  pivot_wider(names_from = "Stat", values_from = "Value")

knitr::kable(spatiotemporal_summary, digits = 3, caption = "Spatio-Temporal Variables - Overall Summary (Pooled across all 1,409 locations and time periods)")
```

## Distributions

```{r spatiotemporal_distributions, fig.cap="Distribution of spatio-temporal variables (pooled across all locations and months)"}
spatiotemporal_data_long <- spatiotemporal_data %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")

p <- ggplot(spatiotemporal_data_long, aes(x = Value)) +
  geom_histogram(bins = 30, fill = "mediumseagreen", alpha = 0.7, color = "black") +
  facet_wrap(~Variable, scales = "free", nrow = 1) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    axis.title = element_text(size = 10)
  ) +
  labs(title = "Distribution of Spatio-Temporal Variables",
       x = "Index Value",
       y = "Frequency")
print(p)
ggsave(file.path(output_dir, "05_spatiotemporal_distributions.png"), p, width = 14, height = 6, dpi = 300)
```

### Temporal Variation by Month

```{r spatiotemporal_by_month, fig.cap="Spatio-temporal indices by month - showing seasonal patterns"}
if ("year_month" %in% colnames(env_all)) {
  spatial_temporal_by_month <- env_all %>%
    select(all_of(c("year_month", spatiotemporal_vars))) %>%
    drop_na() %>%
    pivot_longer(all_of(spatiotemporal_vars), names_to = "Index", values_to = "Value")
  
  p <- ggplot(spatial_temporal_by_month, aes(x = year_month, y = Value, fill = year_month)) +
    geom_boxplot(alpha = 0.7) +
    facet_wrap(~Index, scales = "free_y", nrow = 1) +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 90, size = 7),
      strip.text = element_text(face = "bold")
    ) +
    labs(title = "Spatio-Temporal Indices by Month - Seasonal Patterns",
         x = "Year-Month",
         y = "Index Value")
  print(p)
  ggsave(file.path(output_dir, "06_spatiotemporal_by_month.png"), p, width = 14, height = 6, dpi = 300)
} else {
  cat("Note: year_month variable not found in dataset")
}
```

### Spatial Variation by Landcover

```{r spatiotemporal_by_landcover, fig.cap="Spatio-temporal indices by landcover"}
spatial_temporal_by_cover <- env_all %>%
  select(all_of(c("landcover", spatiotemporal_vars))) %>%
  drop_na() %>%
  pivot_longer(all_of(spatiotemporal_vars), names_to = "Index", values_to = "Value")

p <- ggplot(spatial_temporal_by_cover, aes(x = landcover, y = Value, fill = landcover)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~Index, scales = "free_y", nrow = 1) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  ) +
  labs(title = "Spatio-Temporal Indices by Landcover - Spatial Heterogeneity",
       x = "Landcover Type",
       y = "Index Value",
       fill = "Landcover")
print(p)
ggsave(file.path(output_dir, "07_spatiotemporal_by_landcover.png"), p, width = 12, height = 6, dpi = 300)
```

---

## Collinearity in Environmental data

Quick overview of correlations between all environmental numeric variables.
Using the Spearman correlation instead of the Pearson correlation we don't make any assumptions about linearity in the relation between variables. Booth et al. (1994) suggest that anything above ± 0.5 indicates high collinearity

### Correlations in numerical data - raw
NDVI, NDMI, NDWI, Temp, Precip, RH, Windspeed, nr of aljibes, water_containers

```{r numeric_vars_correlation}
all_numeric <- c(temporal_vars, continuous_spatial, spatiotemporal_vars)

correlation_data <- env_all %>%
  select(all_of(all_numeric)) %>%
  drop_na()

corr_matrix <- cor(correlation_data, method = "spearman")

# Save to file

png(file.path(output_dir, "11a_numeric_vars_correlation.png"), width = 800, height = 800, res = 150)
corrplot(corr_matrix, 
         method = "color",
         type = "upper",
         addCoef.col = "black",
         diag = FALSE,
         tl.col = "black",
         tl.cex = 0.7,
         number.cex = 0.7,
         mar = c(0, 0, 3.5, 0),
         title = "Spearman Correlation raw")
dev.off()

# Display in notebook
corrplot(corr_matrix, 
         method = "color",
         type = "upper",
         addCoef.col = "black",
         diag = FALSE,
         tl.col = "black",
         title = "Spearman Correlation raw",
         tl.cex = 0.8)
```


```{r pairwise_plot_numeric_vars, fig.cap="Pairwise scatterplot matrix of numeric explanatory variables"}
# Reuse numeric variables from previous chunk; build fresh complete-case dataset
pairwise_data <- env_all %>%
  select(all_of(all_numeric)) %>%
  drop_na()

# Optional downsampling for readability/performance with very large datasets
set.seed(123)
max_points <- 5000
if (nrow(pairwise_data) > max_points) {
  pairwise_data <- pairwise_data %>%
    slice_sample(n = max_points)
  cat("Note: pairs() plot uses a random sample of", max_points, "rows for readability.\n")
}

# Save pairwise plot to file
png(file.path(output_dir, "11a_numeric_vars_pairs_plot.png"), width = 1800, height = 1800, res = 200)
pairs(
  pairwise_data,
  pch = 16,
  cex = 0.35,
  col = rgb(0, 0, 0, 0.15),
  main = "Pairwise Scatterplot Matrix - Numeric Environmental Variables"
)
dev.off()

# Display in report
pairs(
  pairwise_data,
  pch = 16,
  cex = 0.35,
  col = rgb(0, 0, 0, 0.15),
  main = "Pairwise Scatterplot Matrix - Numeric Environmental Variables"
)
```

There is a stong collinearity between NDVI, NDMI and NDWI. The easiest solution would be to keep only one of the 3 variables or to create a linear combination of the (two or) three. 

There is also strong collinearity between Total precipitation and precip_max_daily as well as total precipitation and relative humidity. 

### Correct correlation between precip and precip_max_day

```{r residualize precip_max_day on total_precip}
library(dplyr)

# 1) Fit regression of precip_max_day on total_precip
resid_fit <- lm(precip_max_day ~ total_precip, data = env_all)

# 2) Add residualized variable
env_all <- env_all %>%
  mutate(
    precip_max_day_resid = resid(resid_fit)
  )

# 3) Quick check: should be ~uncorrelated with total_precip
cor(env_all$total_precip, env_all$precip_max_day_resid, use = "complete.obs")
```

### Correct correlation between avg_temp and total_precip 






Re-check the correlation plot with the new variable. 
```{r pairwise_plot_numeric_vars after correction, fig.cap="Pairwise scatterplot matrix of numeric explanatory variables after correction"}
all_numeric <- c(all_numeric, "precip_max_day_resid")

# remove the original variables causing issues with collinearity 
all_numeric_no_colin <- all_numeric[!all_numeric %in% c("precip_max_day", "mean_ndmi", "mean_ndwi", "WS2M")]

correlation_data <- env_all %>%
  select(all_of(all_numeric_no_colin)) %>%
  drop_na()

corr_matrix <- cor(correlation_data, method = "spearman")

# Save to file
png(file.path(output_dir, "11a_numeric_vars_correlation_corrected.png"), width = 800, height = 800, res = 150)
corrplot(corr_matrix, 
         method = "color",
         type = "upper",
         addCoef.col = "black",
         diag = FALSE,
         tl.col = "black",
         tl.cex = 0.7,
         number.cex = 0.7,
         mar = c(0, 0, 3.5, 0),
         title = "Spearman Correlation raw")
dev.off()

# Display in notebook
corrplot(corr_matrix, 
         method = "color",
         type = "upper",
         addCoef.col = "black",
         diag = FALSE,
         tl.col = "black",
         title = "Spearman Correlation raw",
         tl.cex = 0.8)
```


```{r pairwise_plot_numeric_vars without collinearity, fig.cap="Pairwise scatterplot matrix of numeric explanatory variables"}
all_numeric_no_colin <- all_numeric[!all_numeric %in% c("precip_max_day", "mean_ndmi", "mean_ndwi", "WS2M", "water_containers", "nr_aljibes")]
# Reuse numeric variables from previous chunk; build fresh complete-case dataset
pairwise_data <- env_all %>%
  select(all_of(all_numeric_no_colin)) %>%
  drop_na()

# Optional downsampling for readability/performance with very large datasets
set.seed(123)
max_points <- 5000
if (nrow(pairwise_data) > max_points) {
  pairwise_data <- pairwise_data %>%
    slice_sample(n = max_points)
  cat("Note: pairs() plot uses a random sample of", max_points, "rows for readability.\n")
}

# Save pairwise plot to file
png(file.path(output_dir, "11a_numeric_vars_pairs_plot_corrected.png"), width = 1800, height = 1800, res = 200)
pairs(
  pairwise_data,
  pch = 25,
  cex = 0.35,
  col = rgb(0, 0, 0, 0.15),
  main = "Pairwise Scatterplot Matrix - Numeric Environmental Variables"
)
dev.off()

# Display in report
pairs(
  pairwise_data,
  pch = 20,
  cex = 0.35,
  col = rgb(0, 0, 0, 0.15),
  main = "Pairwise Scatterplot Matrix - Numeric Environmental Variables"
)
```
### Correlation Among Binary Spatial Variables

```{r spatial_categorical_correlation}
# Select binary spatial variables (excluding landcover as it's multi-category categorical)
binary_vars <- c("is_WI", "is_WUI", "is_urban", "has_aljibes", "water_shortage")

# Prepare data: convert all to numeric format for correlation
spatial_binary_data <- env_all %>%
  select(all_of(binary_vars)) %>%
  drop_na() %>%
  mutate(
    across(all_of(binary_vars), ~as.numeric(.))
  )

# Calculate Spearman correlation matrix
corr_matrix_categ <- cor(spatial_binary_data, method = "spearman")

# Create correlation plot
png(file.path(output_dir, "11b_spatial_categorical_correlation.png"), width = 800, height = 800, res = 150)
corrplot(corr_matrix_categ,
         method = "color",
         type = "lower",
         addCoef.col = "black",
         diag = FALSE,
         tl.col = "black",
         tl.cex = 0.9,
         title = "Spearman Correlation - Binary Spatial Variables",
         mar = c(0, 0, 2, 0))
dev.off()

# Display in notebook
corrplot(corr_matrix_categ,
         method = "color",
         type = "lower",
         addCoef.col = "black",
         diag = FALSE,
         tl.col = "black",
         tl.cex = 0.9,
         title = "Spearman Correlation - Binary Spatial Variables",
         mar = c(0, 0, 2, 0))
```

Binary variables don't seem to have any issues with collinearity. 

```{r correlation with landcover and binary vars}
# List of binary variables
binary_vars <- c("is_WI", "is_WUI", "is_urban", "has_aljibes", "water_shortage")

run_fisher <- function(tbl) {
  tryCatch(
    fisher.test(tbl),
    error = function(e) {
      fisher.test(tbl, simulate.p.value = TRUE, B = 10000)
    }
  )
}

library(vcd)
for (var in binary_vars) {
  tbl <- table(env_all$landcover, env_all[[var]])
  print(var)
  print(tbl)
  if (any(tbl < 5)) {
    print(run_fisher(tbl))
  } else {
    print(run_fisher(tbl))
  }
}

# Visualize with bar plots
library(ggplot2)
for (var in binary_vars) {
  p <- ggplot(env_all, aes(x = landcover, fill = as.factor(.data[[var]]))) +
    geom_bar(position = "fill") +
    labs(title = paste("Proportion of", var, "by Landcover"),
         x = "Landcover", y = "Proportion", fill = var) +
    theme_minimal()
  print(p)
}
```
---

# Descriptive Statistics - Entomological Indicators
```{r load the entomo_epi_env data}
# Load data
env_epi_ento <- read_csv("/mnt/UGent/shares/bionamix/SOFI/Env_data_cuba/CF_Municipality/env_epi_entomo_data_per_manzana_2016_01_to_2019_12.csv")

# Ensure year_month is recognized as date/factor
if ("year_month" %in% colnames(env_epi_ento)) {
  env_epi_ento <- env_epi_ento %>%
    mutate(year_month = as.factor(year_month))
}
```

## Remove precip_max_day and replace by precip_max_day_resid to reduce colinnearity 
```{r create residuals for precip_max_day in linear relation to total_precip}
# na.exclude pads residuals with NA for missing rows so lengths match the dataframe
resid_fit <- lm(precip_max_day ~ total_precip, data = env_epi_ento, na.action = na.exclude)

env_epi_ento <- env_epi_ento %>%
  mutate(precip_max_day_resid = resid(resid_fit)) 
```


## Part 1: Mosquito Survey Data

### Overall Positivity Rates

```{r ento_overall_positivity}
# Calculate positivity rate across all block-month combinations
# Houses_pos_IS: number of houses with immature stages per block-month
# Inspected_houses: number of houses inspected per block (constant over time)

if (all(c("Houses_pos_IS", "Inspected_houses") %in% colnames(env_epi_ento))) {
  overall_positivity <- env_epi_ento %>%
    summarise(
      Total_Block_Months = n(),
      Total_Houses_Inspected = sum(Inspected_houses, na.rm = TRUE),
      Total_Positive_Houses = sum(Houses_pos_IS, na.rm = TRUE),
      Missing_Positivity = sum(is.na(Houses_pos_IS)),
      Overall_Positivity_Rate_Pct = sum(Houses_pos_IS, na.rm = TRUE) / sum(Inspected_houses, na.rm = TRUE) * 100
    )
  
  knitr::kable(overall_positivity, digits = 2, caption = "Overall Mosquito Survey Positivity Rates Across All Block-Months")
} else {
  cat("Note: Houses_pos_IS or Inspected_houses variable not found in dataset.")
}
```

### Total Number of Unique Houses in Database

```{r ento_total_houses}
if ("Inspected_houses" %in% colnames(env_epi_ento)) {
  # Get unique manzanas and their Inspected_houses values (constant per manzana)
  unique_houses <- env_epi_ento %>%
    select(manzana, Inspected_houses) %>%
    distinct(manzana, .keep_all = TRUE) %>%
    drop_na()
  
  total_unique_houses <- sum(unique_houses$Inspected_houses, na.rm = TRUE)
  n_unique_manzanas <- nrow(unique_houses)
  
  cat("\n**Total Number of Houses in Database:**\n\n")
  cat("- Number of unique manzanas: ", n_unique_manzanas, "\n")
  cat("- Total number of inspected houses (summed across unique manzanas): ", total_unique_houses, "\n")
  cat("- Average houses per manzana: ", round(total_unique_houses / n_unique_manzanas, 1), "\n")
} else {
  cat("Note: Inspected_houses variable not found in dataset.")
}
```

### Temporal Distribution of Positivity by Month

```{r ento_temporal_positivity}
if (all(c("Houses_pos_IS", "Inspected_houses", "year_month") %in% colnames(env_epi_ento))) {
  temporal_positivity <- env_epi_ento %>%
    select(year_month, Houses_pos_IS, Inspected_houses) %>%
    drop_na() %>%
    group_by(year_month) %>%
    summarise(
      N_Block_Months = n(),
      Total_Houses_Inspected = sum(Inspected_houses, na.rm = TRUE),
      Total_Positive = sum(Houses_pos_IS, na.rm = TRUE),
      Positivity_Rate_Pct = sum(Houses_pos_IS, na.rm = TRUE) / sum(Inspected_houses, na.rm = TRUE) * 100,
      .groups = "drop"
    ) %>%
    arrange(year_month)
  
  knitr::kable(temporal_positivity, digits = 2, caption = "Mosquito Positivity Rates by Month - Seasonal Patterns")
} else {
  cat("Note: Required variables not found in dataset.")
}
```

### Temporal Trend Visualization

```{r ento_temporal_trend_plot, fig.cap="Mosquito positivity rates over time showing seasonal patterns"}
if (all(c("Houses_pos_IS", "Inspected_houses", "year_month") %in% colnames(env_epi_ento))) {
  temporal_pos_data <- env_epi_ento %>%
    select(year_month, Houses_pos_IS, Inspected_houses) %>%
    drop_na() %>%
    group_by(year_month) %>%
    summarise(
      Total_Pos = sum(Houses_pos_IS),
      Total_n   = sum(Inspected_houses),
      Positivity_Rate = Total_Pos / Total_n,
      se = sqrt(Positivity_Rate * (1 - Positivity_Rate) / Total_n),
      lower = Positivity_Rate - 1.96 * se,
      upper = Positivity_Rate + 1.96 * se,
      N_Blocks = n(),
      .groups = "drop"
    ) 
  
  p <- ggplot(temporal_pos_data, aes(x = year_month, y = Positivity_Rate, group = 1)) +
    geom_line(color = "darkred", alpha = 0.8, linewidth = 1) +
    geom_point(size = 3, color = "darkred", alpha = 0.8) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "darkred", alpha = 0.2) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 90, size = 8),
      plot.title = element_text(face = "bold", size = 12)
    ) +
    labs(title = "Mosquito Survey Positivity Rate Over Time",
         x = "Year-Month",
         y = "Positivity Rate")
  
  print(p)
  ggsave(file.path(output_dir, "11_ento_temporal_positivity.png"), p, width = 12, height = 6, dpi = 300)
}
```

### Spatial Distribution of Positivity by Manzana

```{r ento_spatial_positivity_summary}
# Summary statistics at manzana level (aggregated across all time periods)
if (all(c("manzana", "Houses_pos_IS", "Inspected_houses") %in% colnames(env_epi_ento))) {
  manzana_positivity <- env_epi_ento %>%
    select(manzana, Houses_pos_IS, Inspected_houses) %>%
    drop_na() %>%
    group_by(manzana) %>%
    summarise(
      N_Block_Months = n(),
      Total_Houses_Inspected = sum(Inspected_houses, na.rm = TRUE),
      Total_Positive = sum(Houses_pos_IS, na.rm = TRUE),
      Positivity_Rate = sum(Houses_pos_IS, na.rm = TRUE) / sum(Inspected_houses, na.rm = TRUE) * 100,
      .groups = "drop"
    ) %>%
    arrange(desc(Positivity_Rate))
  
  # Summary of manzana-level statistics
  manzana_summary <- manzana_positivity %>%
    summarise(
      N_Manzanas = n(),
      Mean_Positivity = mean(Positivity_Rate, na.rm = TRUE),
      SD_Positivity = sd(Positivity_Rate, na.rm = TRUE),
      Min_Positivity = min(Positivity_Rate, na.rm = TRUE),
      Max_Positivity = max(Positivity_Rate, na.rm = TRUE),
      Median_Positivity = median(Positivity_Rate, na.rm = TRUE),
      Zero_Positive_Manzanas = sum(Positivity_Rate == 0, na.rm = TRUE),
      High_Transmission_Manzanas = sum(Positivity_Rate > 50, na.rm = TRUE)
    )
  
  knitr::kable(manzana_summary, digits = 2, caption = "Spatial Heterogeneity in Mosquito Positivity - Manzana Level Summary (Aggregated Across All Time Periods)")
  
  # Show top 20 highest positivity manzanas
  cat("\n\n**Top 20 Manzanas with Highest Mosquito Positivity Rates:**\n")
  knitr::kable(head(manzana_positivity, 20), digits = 2)
} else {
  cat("Note: Required variables for spatial analysis not found in dataset.")
}
```

### Positivity Distribution Across Manzanas

```{r ento_positivity_histogram, fig.cap="Distribution of mosquito positivity rates across manzanas"}
if (all(c("manzana", "Houses_pos_IS", "Inspected_houses") %in% colnames(env_epi_ento))) {
  manzana_pos_dist <- env_epi_ento %>%
    select(manzana, Houses_pos_IS, Inspected_houses) %>%
    drop_na() %>%
    group_by(manzana) %>%
    summarise(
      Positivity_Rate = sum(Houses_pos_IS) / sum(Inspected_houses) * 100,
      .groups = "drop"
    )
  
  p <- ggplot(manzana_pos_dist, aes(x = Positivity_Rate)) +
    geom_histogram(bins = 30, fill = "darkred", alpha = 0.7, color = "black") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 12)) +
    labs(title = "Distribution of Mosquito Positivity Rates Across Manzanas",
         x = "Positivity Rate (%) grouped by manzana",
         y = "Number of Manzanas")
  
  print(p)
  ggsave(file.path(output_dir, "12_ento_positivity_distribution.png"), p, width = 10, height = 6, dpi = 300)
}
```

### Spatial Map of Mosquito Positivity

```{r ento_spatial_map, fig.cap="Spatial distribution of mosquito positivity rates across manzanas - handled for zero-inflation"}
if (all(c("manzana", "Houses_pos_IS", "Inspected_houses") %in% colnames(env_epi_ento))) {
  # Calculate positivity rate per manzana (aggregated across time)
  manzana_pos_map <- env_epi_ento %>%
    select(manzana, Houses_pos_IS, Inspected_houses) %>%
    drop_na() %>%
    group_by(manzana) %>%
    summarise(
      Positivity_Rate = sum(Houses_pos_IS) / sum(Inspected_houses) * 100,
      .groups = "drop"
    )
  
  # Join with shapefile
  manzanas_ento <- manzanas %>%
    left_join(manzana_pos_map, by = c("CODIGO_" = "manzana"))
  
  # Create quantile-based categories to handle zero-inflation
  # Separate zeros and non-zeros
  manzanas_ento <- manzanas_ento %>%
    mutate(
      Positivity_Category = case_when(
        is.na(Positivity_Rate) ~ "No Data",
        Positivity_Rate == 0 ~ "Zero",
        Positivity_Rate > 0 & Positivity_Rate <= quantile(Positivity_Rate[Positivity_Rate > 0], 0.33, na.rm = TRUE) ~ "Low (1st tertile)",
        Positivity_Rate > quantile(Positivity_Rate[Positivity_Rate > 0], 0.33, na.rm = TRUE) & 
        Positivity_Rate <= quantile(Positivity_Rate[Positivity_Rate > 0], 0.67, na.rm = TRUE) ~ "Medium (2nd tertile)",
        Positivity_Rate > quantile(Positivity_Rate[Positivity_Rate > 0], 0.67, na.rm = TRUE) ~ "High (3rd tertile)",
        TRUE ~ "No Data"
      )
    )
  
  # Create map with categorical coloring
  p <- ggplot() +
    geom_sf(data = municipality, fill = NA, color = "black", linewidth = 0.8) +
    geom_sf(data = manzanas_ento, aes(fill = Positivity_Category), color = "gray40", linewidth = 0.1) +
    scale_fill_manual(
      values = c(
        "Zero" = "white",
        "Low (1st tertile)" = "#deebf7",
        "Medium (2nd tertile)" = "#9ecae1",
        "High (3rd tertile)" = "#3182bd",
        "No Data" = "grey"
      ),
      na.value = "white",
      breaks = c("Zero", "Low (1st tertile)", "Medium (2nd tertile)", "High (3rd tertile)", "No Data")
    ) +
    theme_minimal() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(face = "bold", size = 12),
      legend.position = "right",
      legend.text = element_text(size = 9)
    ) +
    labs(
      title = "Mosquito Survey Positivity Rates Across Manzanas",
      subtitle = "(Stratified by tertiles due to zero-inflation)",
      fill = "Positivity\nCategory"
    )
  
  print(p)
  ggsave(file.path(output_dir, "13_ento_spatial_map_positivity.png"), p, width = 11, height = 9, dpi = 300)
  
  # Print summary of categories for reference
  cat("\n**Positivity Category Breakdown:**\n")
  category_summary <- manzanas_ento %>%
    st_drop_geometry() %>%
    group_by(Positivity_Category) %>%
    summarise(
      Count = n(),
      Mean_Rate = mean(Positivity_Rate, na.rm = TRUE),
      Min_Rate = min(Positivity_Rate, na.rm = TRUE),
      Max_Rate = max(Positivity_Rate, na.rm = TRUE),
      .groups = "drop"
    )
  print(knitr::kable(category_summary, digits = 2))
} else {
  cat("Required columns not found in env_epi_ento dataset.\n")
}
```

### Positivity by Manzana Over Time (Spatio-Temporal)

```{r ento_positivity_by_manzana_time, fig.cap="Mosquito positivity rates by manzana and month - heatmap of all manzanas", fig.width=16, fig.height=10}
if (all(c("manzana", "year_month", "Houses_pos_IS", "Inspected_houses") %in% colnames(env_epi_ento))) {
  # Calculate positivity for each manzana-month combination
  manzana_month_positivity <- env_epi_ento %>%
    select(manzana, year_month, Houses_pos_IS, Inspected_houses) %>%
    drop_na() %>%
    group_by(manzana, year_month) %>%
    summarise(
      Positivity_Rate = sum(Houses_pos_IS) / sum(Inspected_houses) * 100,
      N_Houses_Inspected = sum(Inspected_houses),
      .groups = "drop"
    )
  
  # Get all manzanas ordered by mean positivity (for ordering in plot)
  manzana_order <- manzana_month_positivity %>%
    group_by(manzana) %>%
    summarise(Mean_Pos = mean(Positivity_Rate, na.rm = TRUE), .groups = "drop") %>%
    arrange(manzana) %>%
    pull(manzana)
  
  # Prepare data with ordered manzanas
  heatmap_data <- manzana_month_positivity %>%
    mutate(
      manzana = factor(manzana, levels = manzana_order),
      year_month = factor(year_month)
    )
  
  # Create heatmap
  p <- ggplot(heatmap_data, aes(x = manzana, y = year_month, fill = Positivity_Rate)) +
    geom_tile(color = "gray30", linewidth = 0.1) +
    scale_fill_viridis_c(option = "YlOrRd", direction = 1, na.value = "white", 
                         limits = c(0, 110), trans = "log1p") +
    theme_minimal() +
    theme(
      axis.text.x = element_blank(),
      axis.text.y = element_text(vjust = 0.5, hjust = 1, size = 8),
      plot.title = element_text(face = "bold", size = 12),
      legend.title = element_text(size = 10),
      panel.grid = element_blank()
    ) +
    labs(
      title = "Mosquito Positivity Rates Over Time - All Manzanas (Ordered by Mean Positivity)",
      x = "Manzana",
      y = "Year-Month",
      fill = "Positivity\nRate (%)"
    )
  
  print(p)
  ggsave(file.path(output_dir, "14_ento_positivity_heatmap_manzana_time.png"), p,width = 20, height = 10, dpi = 300)
}
```

### Positivity by CMF over time (spatio-temporal)

First create a unique Area_CMF indicator 
```{r create CMF indicator Area_CMF}
env_epi_ento_CMF <- env_epi_ento %>% 
  mutate(Area_CMF = paste(AREA, CMF, sep = "_"))
```

```{r ento_positivity_by_cmf_time, fig.cap="Mosquito positivity rates by CMF and month - heatmap of all CMFs", fig.width=16, fig.height=10}
if (all(c("Area_CMF", "year_month", "Houses_pos_IS", "Inspected_houses") %in% colnames(env_epi_ento_CMF))) {
  # Calculate positivity for each CMF-month combination (aggregate to CMF level)
  cmf_month_positivity <- env_epi_ento_CMF %>%
    select(Area_CMF, year_month, Houses_pos_IS, Inspected_houses) %>%
    drop_na() %>%
    group_by(Area_CMF, year_month) %>%
    summarise(
      Positivity_Rate = sum(Houses_pos_IS) / sum(Inspected_houses) * 100,
      N_Houses_Inspected = sum(Inspected_houses),
      .groups = "drop"
    )
  
  # Get all Area_CMFs ordered first by Area_CMF, then by mean positivity
  cmf_order <- cmf_month_positivity %>%
    group_by(Area_CMF) %>%
    summarise(Mean_Pos = mean(Positivity_Rate, na.rm = TRUE), .groups = "drop") %>%
    arrange(Area_CMF, desc(Mean_Pos)) %>%
    pull(Area_CMF)
  
  # Prepare data with ordered Area_CMFs
  cmf_heatmap_data <- cmf_month_positivity %>%
    mutate(
      Area_CMF = factor(Area_CMF, levels = cmf_order),
      year_month = factor(year_month)
    )
  
  # Create heatmap
  p <- ggplot(cmf_heatmap_data, aes(x = Area_CMF, y = year_month, fill = Positivity_Rate)) +
    geom_tile(color = "gray30", linewidth = 0.1) +
    scale_fill_viridis_c(option = "YlOrRd", direction = 1, na.value = "white", 
                         limits = c(0, 110), trans = "log1p") +
    theme_minimal() +
    theme(
      axis.text.y = element_text(vjust = 0.5, hjust = 1, size = 8),
      axis.text.x = element_text(angle = 90, size = 9),
      plot.title = element_text(face = "bold", size = 12),
      legend.title = element_text(size = 10),
      panel.grid = element_blank()
    ) +
    labs(
      title = "Mosquito Positivity Rates Over Time - All Area_CMFs (Ordered by Area then Mean Positivity)",
      x = "Area_CMF",
      y = "Year-Month",
      fill = "Positivity\nRate (%)"
    )
  
  print(p)
  ggsave(file.path(output_dir, "15_ento_positivity_heatmap_cmf_time.png"), p, width = 20, height = 10, dpi = 300)
} else {
  cat("Note: CMF or required variables not found in dataset.\n")
}
```

## Temporal autocorrelation

Check whether IS posivity rate within manzanas show autocorrelation to their previous timestep.
temporal correlation is calculated across all manzanas, comparing each month’s positivity rate to the previous month’s rate within the same manzana 
```{r temporal autocorrelation in positivity index}
# # Temporal autocorrelation for IS positivity index per manzana
library(dplyr)
library(ggplot2)

# Assume df contains columns: block (manzana), year_month_date, y_bt (positives), n_bt (total tested)
# If your data frame is named differently, adjust 'df' accordingly
ento <- env_epi_ento
ento <- ento %>%
  mutate(pos_rate = Houses_pos_IS/Inspected_houses) %>%
  arrange(manzana, year_month) %>%
  group_by(manzana) %>%
  mutate(pos_rate_lag1 = lag(pos_rate, 1)) %>%
  ungroup()

# Check autocorrelation: correlation between pos_rate and pos_rate_lag1
autocorr_result <- ento %>%
  filter(!is.na(pos_rate) & !is.na(pos_rate_lag1)) %>%
  summarise(correlation = cor(pos_rate, pos_rate_lag1))
print(autocorr_result)

# Optionally, plot for visual inspection
ggplot(ento, aes(x = pos_rate_lag1, y = pos_rate)) +
  geom_point(alpha = 0.5) +
  labs(title = "IS Positivity Rate Autocorrelation (Lag 1)",
       x = "Previous Month Positivity Rate",
       y = "Current Month Positivity Rate",
      subtitle = paste("Correlation: ", round(autocorr_result$correlation, 3))) +
  theme_minimal()
```

Let's re-test the temporal autocorrelation but this time taking out all null-values (both at time t and time t-1)

```{r re-test temp autocorr positivity rate after removal null values}
ento_nonull <- ento %>%
  filter(pos_rate != 0 & pos_rate_lag1 != 0)

autocorr_result <- ento_nonull %>%
  filter(!is.na(pos_rate) & !is.na(pos_rate_lag1)) %>%
  summarise(correlation = cor(pos_rate, pos_rate_lag1))
print(autocorr_result)

# Optionally, plot for visual inspection
ggplot(ento_nonull, aes(x = pos_rate_lag1, y = pos_rate)) +
  geom_point(alpha = 0.5) +
  labs(title = "IS Positivity Rate Autocorrelation (Lag 1) after removal null values",
       x = "Previous Month Positivity Rate",
       y = "Current Month Positivity Rate", 
       subtitle = paste("Correlation: ", round(autocorr_result$correlation, 3))) +
  theme_minimal()
```
After removal of null values (98% of the data), the correlation is very large ...

## Spatial auto-correlation 

Next to collinearity in the explanatory variables, spatial auto-correlation forms an important issue to check before building regression models. Positive spatial auto-correlation violates the assumption of independent sampling. We will use Moran's I to measure spatial auto-correlation. However, we need to consider how to define *neighbors*. This can be done based on 
1. touching polygons (border or corner = Queen or Rook respectively), 
  Pro's: simple and common in areal models (CAR, BYM)
  Con's: not good if blocks are uneven in size, isolated blocks may have no neighbors
2. distance-based (manzanas are neighbors if they are within x meters of eachother), 
  Pro's: Biologically meaningfull if you take for instance average (or max) flight range of *Aedes* mosquitoes. Handles irregular spacing. No issue for isolated blocks.
  Con's: requires choosing threshold. 
3. or **inverse-distance weighting** were all blocks influence each other $w_{i,j} = \frac{1}{d_{i,j}}$. 
  Pro's: No need to choose threshold. Continuous spatial decay. Ecologically realistic. 
  Con's Dense weight matrix. Harder to interprete CAR framework. 

We will apply an inverse-distance weighting ~with a 400m cut-off for maximal mosquito flight range~. I had to remove the cut-off because it caused many manzanas to become isolated, breaking the logic of Moran's I. Instead I used a stronger weighting for distance $w{ij} = \frac{1}{d_{ij}^2}$. 

We will test this two ways: 
1. On raw counts Houses_Pos_IS
2. On proportions Houses_Pos_IS/Inspected_houses
3. Ideally on model residuals -> this will have to wait until after fitting 

### Moran's I
```{r join mosquito data with shapefile}
mosquito_manzana <- env_epi_ento %>%
  select(manzana, Houses_pos_IS, Inspected_houses, cases) %>%
  drop_na() %>%
  group_by(manzana) %>%
  summarise(
    Total_Houses_Inspected = sum(Inspected_houses),
    Total_Positive_Houses = sum(Houses_pos_IS),
    Positivity_Rate = sum(Houses_pos_IS) / sum(Inspected_houses) * 100,
    Total_Cases = sum(cases),
    .groups = "drop"
  )

# Join with manzanas shapefile
manzanas_mosquito <- manzanas %>%
  left_join(mosquito_manzana, by = c("CODIGO_" = "manzana"))
```

Check the Coordinate Reference System (CRS)

```{r check_crs}
# Check CRS of manzanas_mosquito
cat("CRS of manzanas_mosquito:\n")
print(st_crs(manzanas_mosquito))

# Check if CRS is projected (needed for distance calculations in meters)
if (st_is_longlat(manzanas_mosquito)) {
  cat("\nWarning: CRS is geographic (lon/lat), not projected.\n")
  cat("Distance calculations will be in degrees, not meters.\n")
  cat("Consider transforming to a projected CRS for meaningful distance metrics.\n")
} else {
  cat("\nCRS is projected. Distance calculations will be in the units of the CRS.\n")
}
```

Create the spatial neighbors matrix 

```{r spatial neighbors creation}
library(spdep)

# Get centroids (keep only geometry, drop attributes)
centroids <- st_geometry(manzanas_mosquito) %>%
  st_centroid()
  
# Calculate pairwise distances (in meters, assuming projected CRS)
distances <- st_distance(centroids)

# Convert to matrix and drop units
dist_matrix <- units::drop_units(as.matrix(distances))

# Create inverse-distance weights (1/distance)
# Set diagonal to 0 (no self-loops)
inv_dist_weights <- 1 / dist_matrix^2
diag(inv_dist_weights) <- 0
inv_dist_weights[is.infinite(inv_dist_weights)] <- 0

# # Apply 400m cutoff for mosquito maximal flight range
# d_max <- 400
# inv_dist_weights[dist_matrix > d_max] <- 0

# Normalize row-wise
# Rows with no neighbors (isolated blocks) will have row_sum = 0
row_sums <- rowSums(inv_dist_weights)
inv_dist_weights_normalized <- inv_dist_weights / row_sums

# Handle NaN/Inf values (from division by 0 for isolated blocks)
inv_dist_weights_normalized[is.nan(inv_dist_weights_normalized)] <- 0
inv_dist_weights_normalized[is.infinite(inv_dist_weights_normalized)] <- 0

# Report isolated blocks
n_isolated <- sum(row_sums == 0)
if (n_isolated > 0) {
  cat("Note: ", n_isolated, " manzanas have no neighbors within 400m and are isolated.\n")
}

# Convert to listw format for spdep
weights_idw <- mat2listw(inv_dist_weights_normalized, style = "W")  
```

```{r calculate Morans I on positivity rate}
# Remove NAs manually
valid_idx <- !is.na(manzanas_mosquito$Positivity_Rate)
positivity_clean <- manzanas_mosquito$Positivity_Rate[valid_idx]

# Check how many NAs were removed
cat("Total manzanas in shapefile: ", nrow(manzanas_mosquito), "\n")
cat("Manzanas with valid Positivity_Rate: ", sum(valid_idx), "\n")
cat("Manzanas with NA Positivity_Rate: ", sum(!valid_idx), "\n")
cat("Percentage with valid data: ", round(sum(valid_idx) / nrow(manzanas_mosquito) * 100, 1), "%\n")

# Subset weights to match (only keep rows/cols for valid observations)
# Convert to matrix, subset, then back to listw
weights_matrix <- listw2mat(weights_idw)
weights_matrix_clean <- weights_matrix[valid_idx, valid_idx]
weights_idw_clean <- mat2listw(weights_matrix_clean, style = "W")

moran_global <- moran.test(positivity_clean,
                           weights_idw_clean,
                           na.action = na.exclude,
                           zero.policy = TRUE)

print(moran_global)
```

With a Moran's I of 0.14 with significance, we can say that there is strong evidence for spatial auto-correlation (spatial clustering) in the positivity rate! 

The next step will be to test the Moran's I on the residuals fromt he model, after accounting for the spatial explanantory variables. If the Moran's I is no longer singificant after that, then we don't need to account for additional spatial autocorrelation in the model. If it is, then we should include a spatial random effect in the model (likely). 

### Moran's I Correlogram (distance bands) with inverse squared weighting
Now I would like to know at what distance the spatial clustering remains. To know up to what distance are manzanas positively correlated. 

```{r morans I correlogram}
# Get coordinates for valid observations
coords_valid <- centroids[valid_idx]

# Pairwise distance matrix (meters)
dist_matrix_valid <- units::drop_units(as.matrix(st_distance(coords_valid)))
diag(dist_matrix_valid) <- NA_real_

# Test at non-cumulative distance bands (annuli)
distance_breaks <- seq(0, 1000, by = 50)  # 0-50, 50-100, ..., 950-1000 m
moran_by_distance <- data.frame()

for (i in 2:length(distance_breaks)) {
  d_low <- distance_breaks[i - 1]
  d_high <- distance_breaks[i]

  inv_dist_sq_weights <- 1 / (dist_matrix_valid^2)
  diag(inv_dist_sq_weights) <- 0
  inv_dist_sq_weights[is.infinite(inv_dist_sq_weights)] <- 0
  # Keep only links in current distance band (d_low, d_high]
  inv_dist_sq_weights[is.na(dist_matrix_valid) | dist_matrix_valid <= d_low | dist_matrix_valid > d_high] <- 0

  # Skip empty bands with no links
  if (sum(inv_dist_sq_weights > 0, na.rm = TRUE) == 0) {
    moran_by_distance <- rbind(
      moran_by_distance,
      data.frame(
        Distance_from_m = d_low,
        Distance_to_m = d_high,
        Distance_mid_m = (d_low + d_high) / 2,
        Morans_I = NA_real_,
        P_Value = NA_real_,
        Significant = NA
      )
    )
    next
  }
  
  # Convert to listw with zero.policy = TRUE
  weights_d <- mat2listw(inv_dist_sq_weights, style = "W", zero.policy = TRUE)
  
  moran_d <- moran.test(positivity_clean, weights_d, zero.policy = TRUE)
  
  moran_by_distance <- rbind(moran_by_distance, 
    data.frame(Distance_from_m = d_low,
               Distance_to_m = d_high,
               Distance_mid_m = (d_low + d_high) / 2,
               Morans_I = moran_d$estimate[1],
               P_Value = moran_d$p.value,
               Significant = moran_d$p.value < 0.05))
}

# View and plot results
print(moran_by_distance)
plot(moran_by_distance$Distance_mid_m, moran_by_distance$Morans_I, 
     type = "b", pch = 16, xlab = "Distance (m)", ylab = "Moran's I",
     main = "Spatial Autocorrelation by Distance Band (Inverse-Distance-Squared Weighting)")
abline(h = 0, lty = 2, col = "gray")
grid()
```

Interpretation: 
- at 50 m, neighborhoods are too sparse (many blocks have few/no neighbors), so Moran’s I is unstable/low;
- by 100 m, enough spatial links appear, so clustering is detected strongly;
- from 200–1000 m, the process is already “fully connected,” so I plateaus around ∼0.16 instead of decaying.


```{r plot morans I correlogram}
plot(moran_by_distance$Distance_mid_m, moran_by_distance$Morans_I, 
     type = "b", pch = 16, cex = 1.5, lwd = 2,
     xlab = "Distance (m)", ylab = "Moran's I",
  main = "Spatial Autocorrelation Correlogram by Distance Band\n(Inverse-Squared-Distance Weighting)")

# Color points by significance
colors <- ifelse(moran_by_distance$Significant, "red", "blue")
points(moran_by_distance$Distance_mid_m, moran_by_distance$Morans_I, 
       pch = 16, cex = 1.5, col = colors)

abline(h = 0, lty = 2, col = "gray", lwd = 2)
legend("topright", legend = c("p < 0.05", "p ≥ 0.05"), 
       col = c("red", "blue"), pch = 16)
grid()

# Save the current base plot
dev.copy(
  png,
  filename = file.path(output_dir, "16_squared_inv_moransI_correlogram_distance_band.png"),
  width = 1200,
  height = 800,
  res = 150
)
dev.off()
```

Moran's I remains significant all throughout distances from 100m to 1000m. 

### Moran's I with exponontial distance weighting 

$$I = \frac{n}{S_0};
\frac{\sum_{i=1}^{n}\sum_{j=1}^{n} w_{ij}(x_i-\bar x)(x_j-\bar x)}
{\sum_{i=1}^{n}(x_i-\bar x)^2}$$

where

$$ w_{ij}=\exp!\left(-\frac{d_{ij}}{\phi}\right),\quad i\neq j,\qquad w_{ii}=0 $$

and

$$S_0=\sum_{i=1}^{n}\sum_{j=1}^{n} w_{ij}.$$

```{r morans i exponential weighting}
# Test at non-cumulative distance bands (annuli), for multiple exponential decay values
distance_breaks <- seq(0, 1000, by = 50)  # 0-50, 50-100, ..., 950-1000 m
phi_values <- c(100, 200, 400, 800)

moran_by_distance <- data.frame()

for (phi in phi_values) {
  for (i in 2:length(distance_breaks)) {
    d_low <- distance_breaks[i - 1]
    d_high <- distance_breaks[i]

    exp_weights <- exp(-dist_matrix_valid / phi)
    diag(exp_weights) <- 0
    exp_weights[is.na(dist_matrix_valid)] <- 0

    # Keep only links in current distance band (d_low, d_high]
    exp_weights[dist_matrix_valid <= d_low | dist_matrix_valid > d_high] <- 0

    # Skip empty bands with no links
    if (sum(exp_weights > 0, na.rm = TRUE) == 0) {
      moran_by_distance <- rbind(
        moran_by_distance,
        data.frame(
          phi = phi,
          Distance_from_m = d_low,
          Distance_to_m = d_high,
          Distance_mid_m = (d_low + d_high) / 2,
          Morans_I = NA_real_,
          P_Value = NA_real_,
          Significant = NA
        )
      )
      next
    }

    # Convert to listw with zero.policy = TRUE
    weights_d <- mat2listw(exp_weights, style = "W", zero.policy = TRUE)

    moran_d <- moran.test(positivity_clean, weights_d, zero.policy = TRUE)

    moran_by_distance <- rbind(
      moran_by_distance,
      data.frame(
        phi = phi,
        Distance_from_m = d_low,
        Distance_to_m = d_high,
        Distance_mid_m = (d_low + d_high) / 2,
        Morans_I = moran_d$estimate[1],
        P_Value = moran_d$p.value,
        Significant = moran_d$p.value < 0.05
      )
    )
  }
}

# View results
print(moran_by_distance)

# Plot all phi curves in different colors
plot_data <- moran_by_distance %>%
  dplyr::filter(!is.na(Morans_I)) %>%
  dplyr::mutate(phi = factor(phi, levels = sort(unique(phi))))

p_moran_phi <- ggplot(plot_data, aes(x = Distance_mid_m, y = Morans_I, color = phi, group = phi)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_minimal() +
  labs(
    title = "Moran's I Correlogram by Distance Band",
    subtitle = "Exponential distance weighting: w = exp(-d / phi)",
    x = "Distance band midpoint (m)",
    y = "Moran's I",
    color = "phi (m)"
  )

print(p_moran_phi)
ggsave(file.path(output_dir, "16_exp_moransI_correlogram_multi_phi.png"), p_moran_phi, width = 11, height = 7, dpi = 300)
```

### Variogram 
```{r variogram_positivity_rate, fig.cap="Empirical variogram of mosquito positivity rate across manzanas", fig.width=10, fig.height=6}
# library(sf)
library(gstat)
# library(dplyr)
# library(ggplot2)

# Use the manzanas object already joined with positivity:
# manzanas_mosquito (created in your Moran section)

# Keep valid rows only
vario_data <- manzanas_mosquito %>%
  filter(!is.na(Positivity_Rate)) %>%
  st_make_valid()

# Project to metric CRS if needed (important for distance bins in meters)
if (st_is_longlat(vario_data)) {
  vario_data <- st_transform(vario_data, 32617)  # UTM 17N (Cuba area); adjust if needed
}

# Empirical variogram
vg_emp <- variogram(
  Positivity_Rate ~ 1,
  data = vario_data,
  cutoff = 1000,   # max distance (m)
  width  = 50      # lag bin width (m)
)

# Optional fitted model (spherical as a starting point)
vg_model_init <- vgm(psill = var(vg_emp$gamma, na.rm = TRUE), model = "Sph", nugget = 0)
vg_fit <- fit.variogram(vg_emp, vg_model_init)

# Plot
p <- ggplot(vg_emp, aes(x = dist, y = gamma)) +
  geom_point(size = 2, color = "steelblue") +
  geom_line(color = "steelblue", alpha = 0.6) +
  labs(
    title = "Empirical Variogram - Positivity Rate",
    x = "Distance (m)",
    y = "Semivariance"
  ) +
  theme_minimal()

# Add fitted line if fit succeeded
if (!any(is.na(vg_fit$psill))) {
  vg_line <- variogramLine(vg_fit, maxdist = max(vg_emp$dist, na.rm = TRUE), n = 200)
  p <- p + geom_line(data = vg_line, aes(x = dist, y = gamma), color = "darkred", linewidth = 1)
}

print(p)
ggsave(file.path(output_dir, "16_variogram_positivity_rate.png"), p, width = 10, height = 6, dpi = 300)
```
# Descriptive statistics - Cases

## Part 1: purely temporal 

```{r cases per year_month, fig.cap="Nr of dengue and zika cases per month", fig.width=16, fig.height=10}
cases_temporal <- env_epi_ento %>%
  group_by(year_month) %>%
  summarise(Total_cases = sum(cases))

p <- ggplot(cases_temporal, aes(x = year_month, y = Total_cases)) +
  geom_col(fill = "darkred", alpha = 0.8) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, size = 8),
    plot.title = element_text(face = "bold", size = 12)
  ) +
  labs(title = "Dengue/ZIKV cases over time",
        x = "Year-Month",
        y = "Total dengue case count")

print(p)
ggsave(file.path(output_dir, "17_dengue_cases_temp.png"), p, width = 16, height = 10, dpi = 300)

```

### Dengue Cases by Manzana Over Time (Spatio-Temporal Heatmap)

```{r dengue_cases_by_manzana_time, fig.cap="Dengue cases by manzana and month - heatmap of all manzanas", fig.width=16, fig.height=10}
if (all(c("manzana", "year_month", "cases") %in% colnames(env_epi_ento))) {
  # Calculate total cases for each manzana-month combination
  manzana_month_cases <- env_epi_ento %>%
    select(manzana, year_month, cases) %>%
    drop_na() %>%
    group_by(manzana, year_month) %>%
    summarise(
      Total_Cases = sum(cases),
      .groups = "drop"
    )
  
  # Get all manzanas ordered by mean case count
  manzana_order <- manzana_month_cases %>%
    group_by(manzana) %>%
    summarise(Mean_Cases = mean(Total_Cases, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(Mean_Cases)) %>%
    pull(manzana)
  
  # Prepare data with ordered manzanas
  heatmap_cases_data <- manzana_month_cases %>%
    mutate(
      manzana = factor(manzana, levels = manzana_order),
      year_month = factor(year_month)
    )
  
  # Create heatmap
  p <- ggplot(heatmap_cases_data, aes(x = manzana, y = year_month, fill = Total_Cases)) +
    geom_tile(color = "gray30", linewidth = 0.1) +
    scale_fill_viridis_c(option = "YlOrRd", direction = 1, na.value = "white", 
                         trans = "log1p") +
    theme_minimal() +
    theme(
      axis.text.x = element_blank(),
      axis.text.y = element_text(vjust = 0.5, hjust = 1, size = 8),
      plot.title = element_text(face = "bold", size = 12),
      legend.title = element_text(size = 10),
      panel.grid = element_blank()
    ) +
    labs(
      title = "Dengue Cases Over Time - All Manzanas (Ordered by Mean Case Count)",
      x = "Manzana",
      y = "Year-Month",
      fill = "Total Cases"
    )
  
  print(p)
  ggsave(file.path(output_dir, "18_dengue_cases_heatmap_manzana_time.png"), p, width = 16, height = 10, dpi = 300)
} else {
  cat("Note: Required columns not found in dataset.\n")
}
```

# One-on-one correlation between positivity rate and environmental var's 

```{r reload the data}
env_epi_ento <- read_csv("/media/rita/New Volume/Documenten/DI-MOB/Other Data/Env_data_cuba/data/env_epi_entomo_data_per_manzana_2016_01_to_2019_12_noColinnearity.csv")

env_all <- read_csv("/media/rita/New Volume/Documenten/DI-MOB/Other Data/Env_data_cuba/data/all_environmental_data_per_manzana_2016_01_to_2019_12_imputed.csv")
```


### Create the one-on-one data + Pearson correlation 

#### Numerical temporal variables
```{r env_vs_positivity_one_on_one, fig.width=16, fig.height=12, fig.cap="Temporal variables vs mosquito positivity rate"}
# Temporal-only associations (corrected: one point per month)

temporal_only_vars <- c("avg_temp", "total_precip", "rel_hum", "WS2M", "precip_max_day", "precip_max_day_resid")

# 1) Monthly outcome (weighted by inspected houses)
monthly_pos <- env_epi_ento %>%
  group_by(year_month) %>%
  summarise(
    Positivity_Rate = sum(Houses_pos_IS, na.rm = TRUE) / sum(Inspected_houses, na.rm = TRUE) * 100,
    .groups = "drop"
  )

# 2) Monthly environmental predictors (deduplicate month-level repetition)
monthly_env <- env_epi_ento %>%
  group_by(year_month) %>%
  summarise(across(all_of(temporal_only_vars), ~mean(.x, na.rm = TRUE)), .groups = "drop")

# 3) Join + long format
temporal_corr_data <- monthly_pos %>%
  left_join(monthly_env, by = "year_month") %>%
  pivot_longer(all_of(temporal_only_vars), names_to = "Variable", values_to = "Value") %>%
  filter(!is.na(Value), !is.na(Positivity_Rate))

# 4) Spearman rho per variable (now based on months, not block-month duplicates)
rho_tbl <- temporal_corr_data %>%
  group_by(Variable) %>%
  summarise(rho = cor(Value, Positivity_Rate, method = "spearman"), .groups = "drop") %>%
  mutate(label = paste0("rho = ", round(rho, 2)))

plot_data <- temporal_corr_data %>% left_join(rho_tbl, by = "Variable")
```

Plot the result for numeric temporal variables
```{r plot the one-on-one relations and their spearman correlation}
p_temporal <- ggplot(plot_data, aes(x = Value, y = Positivity_Rate)) +
  geom_point(color = "darkred", alpha = 0.8, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue", linewidth = 0.8) +
  facet_wrap(~Variable, scales = "free_x", ncol = 3) +
  geom_text(
    data = plot_data %>% distinct(Variable, label),
    aes(x = -Inf, y = Inf, label = label),
    hjust = -0.1, vjust = 1.2, inherit.aes = FALSE, size = 3.5
  ) +
  theme_minimal() +
  labs(
    title = "Monthly environmental variables vs monthly mosquito positivity",
    subtitle = "One point per month (pseudo-replication removed)",
    x = "Environmental value",
    y = "Positivity rate (%)"
  )

print(p_temporal)
ggsave(file.path(output_dir, "19_temporal_env_vs_monthly_positivity.png"), p_temporal, width = 14, height = 8, dpi = 300)
```

Actually, the correlation with windspeed and total_precipitation is very very low 0.03 and 0.01 respectively, while the correlation with relative humidity is higher (0.28). 

**Plot the result for numeric temporal variables with 0 & 1 & 2 month lag**
```{r calculate the one-on-one relations and their spearman correlation with lag}
# 1) Create lagged variables for lag 0, 1, and 2
monthly_env_lagged <- monthly_env %>%
  arrange(year_month) %>%
  mutate(
    across(all_of(temporal_only_vars), ~.x, .names = "{.col}_lag0"),
    across(all_of(temporal_only_vars), ~lag(.x, n = 1), .names = "{.col}_lag1"),
    across(all_of(temporal_only_vars), ~lag(.x, n = 2), .names = "{.col}_lag2")
  )


# 2) Join lagged predictors to outcome
lagged_data <- monthly_pos %>%
  left_join(monthly_env_lagged, by = "year_month")

# 3) Pivot to long format for plotting (all lags)
lagged_long <- lagged_data %>%
  pivot_longer(matches("_lag[012]$"), names_to = "Variable", values_to = "Value") %>%
  filter(!is.na(Value), !is.na(Positivity_Rate)) %>%
  mutate(
    Lag = case_when(
      grepl("_lag0$", Variable) ~ "0-month lag",
      grepl("_lag1$", Variable) ~ "1-month lag",
      grepl("_lag2$", Variable) ~ "2-month lag"
    ),
    Variable = gsub("_lag[012]$", "", Variable)
    ) %>%
    mutate(
      Variable = factor(
        Variable,
        levels = c("avg_temp", "total_precip", "precip_max_day", "precip_max_day_resid", "rel_hum", "WS2M")
      )
    )

# 4) Calculate Spearman rho for each variable and lag
rho_tbl <- lagged_long %>%
  group_by(Variable, Lag) %>%
  summarise(rho = cor(Value, Positivity_Rate, method = "spearman"), .groups = "drop") %>%
  mutate(label = paste0("rho = ", round(rho, 2)))

plot_data <- lagged_long %>% left_join(rho_tbl, by = c("Variable", "Lag"))
```

```{r plot the one-on-one temporal vars with different month lag and their spearman correlation}
p_temporal_lagged <- ggplot(plot_data, aes(x = Value, y = Positivity_Rate)) +
  geom_point(color = "firebrick", alpha = 0.7) +
  geom_smooth(method = "lm", color = "steelblue") +
  facet_grid(Lag ~ Variable, scales = "free_x") +
  geom_text(data = rho_tbl, aes(x = -Inf, y = Inf, label = label), hjust = -0.1, vjust = 1.2, inherit.aes = FALSE) +
  labs(
    title = "Monthly environmental variables (1- and 2-month lag) vs monthly mosquito positivity",
    subtitle = "One point per month (pseudo-replication removed)",
    x = "Lagged Environmental value",
    y = "Positivity rate (%)"
  )

print(p_temporal_lagged)
ggsave(file.path(output_dir, "19b_temporal_env_vs_monthly_lag_0_1_2_positivity.png"), p_temporal_lagged, width = 14, height = 8, dpi = 300)
```

### Testing non-linear transformations of numeric variables 
#### Quadratic

```{r temperature quadratic transformations}
p_temporal_lagged_quad <- ggplot(plot_data, aes(x = Value, y = Positivity_Rate)) +
  geom_point(color = "firebrick", alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), color = "steelblue") +
  facet_grid(Lag ~ Variable, scales = "free_x") +
  labs(
    title = "Monthly environmental variables (1- and 2-month lag, quadratic fit) vs monthly mosquito positivity",
    subtitle = "Quadratic model: y = β₀ + β₁x + β₂x² + ε",
    x = "Lagged Environmental value",
    y = "Positivity rate (%)"
  )

print(p_temporal_lagged_quad)

ggsave(file.path(output_dir, "19b_temporal_env_vs_monthly_lag_0_1_2_positivity_quadratic.png"), p_temporal_lagged_quad, width = 14, height = 8, dpi = 300)

```

#### Spline
```{r temperature spline transformations}
library(mgcv) # Required for GAM splines

p_temporal_lagged_spline <- ggplot(plot_data, aes(x = Value, y = Positivity_Rate)) +
  geom_point(color = "firebrick", alpha = 0.7) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), color = "steelblue") +
  facet_grid(Lag ~ Variable, scales = "free_x") +
  labs(
    title = "Monthly environmental variables (1- and 2-month lag, spline fit) vs monthly mosquito positivity",
    subtitle = "Spline model: y = β₀ + f(x) + ε",
    x = "Lagged Environmental value",
    y = "Positivity rate (%)"
  )

print(p_temporal_lagged_spline)
ggsave(file.path(output_dir, "19b_temporal_env_vs_monthly_lag_0_1_2_positivity_spline.png"), p_temporal_lagged_spline, width = 14, height = 8, dpi = 300)
```

The spline is definitely not worth it... not better 


#### Binary spatial variables 

```{r env_vs_positivity_one_on_one spatial}

# 1) Manzana-level outcome (aggregated across time)
manzana_outcome <- env_epi_ento %>%
  group_by(manzana) %>%
  summarise(
    Positivity_Rate = sum(Houses_pos_IS, na.rm = TRUE) / sum(Inspected_houses, na.rm = TRUE) * 100,
    .groups = "drop"
  )

# 2) One row per manzana for spatial predictors
spatial_vars <- c("is_WI", "is_WUI", "is_urban", "has_aljibes", "water_shortage","landcover", "nr_aljibes", "water_containers")

manzana_spatial <- env_all %>%
  select(manzana, all_of(spatial_vars)) %>%
  distinct(manzana, .keep_all = TRUE)

# 3) Join
spatial_df <- manzana_outcome %>%
  left_join(manzana_spatial, by = "manzana")
```


```{r plot the one-on-one relations posivity rate vs binary env var}
binary_vars <- c("is_WI", "is_WUI", "is_urban", "has_aljibes", "water_shortage")

binary_long <- spatial_df %>%
  select(Positivity_Rate, all_of(binary_vars)) %>%
  pivot_longer(all_of(binary_vars), names_to = "Variable", values_to = "Group") %>%
  filter(!is.na(Group), !is.na(Positivity_Rate)) %>%
  mutate(Group = factor(Group, levels = c(FALSE, TRUE), labels = c("No", "Yes")))

p_bin <- ggplot(binary_long, aes(x = Group, y = Positivity_Rate, fill = Group)) +
  geom_violin(alpha = 0.25, trim = FALSE) +
  geom_boxplot(width = 0.2, outlier.alpha = 0.25) +
  facet_wrap(~Variable, scales = "free_x") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Positivity by binary spatial predictors",
       x = "", y = "Positivity rate (%)")

# Wilcoxon tests
wilcox_results <- binary_long %>%
  group_by(Variable) %>%
  wilcox_test(Positivity_Rate ~ Group) %>%
  adjust_pvalue(method = "BH")
print(wilcox_results)

# Add stars to significant plots
star_tbl <- wilcox_results %>%
  mutate(star = ifelse(p.adj < 0.05, "*", "")) %>%
  select(Variable, star)

# Prepare annotation data: place star at top of each plot
annot_df <- binary_long %>%
  group_by(Variable) %>%
  summarise(y = max(Positivity_Rate, na.rm = TRUE) + 2) %>%
  left_join(star_tbl, by = "Variable")

p_bin_star <- p_bin +
  geom_text(
    data = annot_df,
    aes(x = 1.5, y = y, label = star),
    inherit.aes = FALSE,
    size = 8,
    color = "black"
  )

print(p_bin_star)
ggsave(file.path(output_dir, "20_spatial_binary_vs_positivity.png"), p_bin_star, width = 12, height = 8, dpi = 300)
```

The Wilcox test should signficant differences in positivity rate between yes/no of all binary variables (except water_shortage).  

#### Categorical variables (landcover)
```{r one-on-one association to landcover}
land_df <- spatial_df %>%
  filter(!is.na(landcover), !is.na(Positivity_Rate))

p_land <- ggplot(land_df, aes(x = landcover, y = Positivity_Rate, fill = landcover)) +
  geom_boxplot(outlier.alpha = 0.25) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Positivity by landcover", x = "Landcover", y = "Positivity rate (%)")
print(p_land)
ggsave(file.path(output_dir, "21_landcover_vs_positivity.png"), p_land, width = 11, height = 6, dpi = 300)

kruskal_res <- kruskal_test(land_df, Positivity_Rate ~ landcover)
print(kruskal_res)

# If Kruskal-Wallis is significant, identify which landcover groups differ
dunn_res <- land_df %>%
  dunn_test(Positivity_Rate ~ landcover, p.adjust.method = "BH")

print(dunn_res)

# Optional: only show significant pairwise differences after BH correction
dunn_sig <- dunn_res %>%
  filter(p.adj < 0.05)

print(dunn_sig)
```

Kruskal-Wallis test shows singificant difference between groups (or at least some).  
According to the Dunn pairwise table, there is a signficant difference in positivity rate between built-up and mangrove locations and between built-up and tree cover locations. All other pairwise comparisons are not significant at the p (adjusted for multiple comparison) < 0.05 level.

#### Numerical spatial 
```{r one-on-one association numerical spatial vars to positivity rate}
# Build grouped/ordered factor (collapse sparse high levels)
aljibes_df <- spatial_df %>%
  filter(!is.na(nr_aljibes), !is.na(Positivity_Rate)) %>%
  mutate(
    nr_aljibes_grp = factor(nr_aljibes,
                            levels = c("0","1","2","3"),
                            ordered = TRUE)
  )
wc_df <- spatial_df %>%
  filter(!is.na(water_containers), !is.na(Positivity_Rate))

# Plot
p_alj <- ggplot(aljibes_df, aes(x = nr_aljibes_grp, y = Positivity_Rate)) +
  geom_boxplot(outlier.alpha = 0.25, fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Positivity by number of aljibes (ordered groups)",
    x = "nr_aljibes (grouped)",
    y = "Positivity rate (%)"
  )

print(p_alj)
ggsave(file.path(output_dir, "22_nr_aljibes_ordered_vs_positivity.png"),
       p_alj, width = 10, height = 6, dpi = 300)

# Global group difference test
kruskal_test(aljibes_df, Positivity_Rate ~ nr_aljibes_grp)

# One-on-one association: water_containers
wc_df <- spatial_df %>%
  filter(!is.na(water_containers), !is.na(Positivity_Rate))

p_wc <- ggplot(wc_df, aes(x = water_containers, y = Positivity_Rate)) +
  geom_point(alpha = 0.6, color = "darkred", position = position_jitter(width = 0.2)) +
  theme_minimal() +
  labs(title = "Positivity vs water containers (count per manzana)",
       x = "Estimated water containers",
       y = "Positivity rate (%)")

print(p_wc)
ggsave(file.path(output_dir, "23_water_containers_vs_positivity.png"), p_wc, width = 10, height = 6, dpi = 300)

# Spearman correlation
cor(wc_df$water_containers, wc_df$Positivity_Rate, method = "spearman")

# Combine both plots in a 1x2 grid
library(cowplot)
combined_numeric_spatial <- cowplot::plot_grid(p_alj, p_wc, nrow = 1, ncol = 2, labels = c("A", "B"))
print(combined_numeric_spatial)
ggsave(file.path(output_dir, "24_numeric_spatial_vs_positivity_combined.png"), combined_numeric_spatial, width = 18, height = 7, dpi = 300)
```
Again, the Kruskal-wallis test shows that my ordinal 

#### Spatio-temporal variables (NDXI's)

```{r}

manzana_outcome <- env_epi_ento %>%
  group_by(manzana) %>%
  summarise(
    Positivity_Rate = sum(Houses_pos_IS, na.rm = TRUE) / sum(Inspected_houses, na.rm = TRUE) * 100,
    .groups = "drop"
  )

# 2) One row per manzana for spatial predictors
spatiotemporal_vars <- c("mean_ndvi", "mean_ndmi", "mean_ndwi")

manzana_spatio_temp <- env_all %>%
  select(manzana, all_of(spatiotemporal_vars)) %>%
  distinct(manzana, .keep_all = TRUE)

# 3) Join
spatiotemp_df <- manzana_outcome %>%
  left_join(manzana_spatio_temp, by = "manzana")

st_df <- spatiotemp_df %>%
  select(Positivity_Rate, all_of(spatiotemporal_vars)) %>%
  pivot_longer(all_of(spatiotemporal_vars), names_to = "Variable", values_to = "Value") %>%
  filter(!is.na(Value), !is.na(Positivity_Rate))


# Plot
p_st <- ggplot(st_df, aes(x = Value, y = Positivity_Rate)) +
  geom_point(alpha = 0.6, color = "darkred") +
  facet_wrap(~Variable, scales = "free_x", ncol = 3) +
  theme_minimal() +
  labs(title = "Positivity rate vs NDVI, NDMI, NDWI",
       x = "Index value",
       y = "Positivity rate (%)")

# Compute correlation and p-value for annotation
st_rho_p <- st_df %>%
  group_by(Variable) %>%
  summarise(
    rho = cor(Value, Positivity_Rate, method = "spearman"),
    p = cor.test(Value, Positivity_Rate, method = "spearman")$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    star = ifelse(p < 0.05, "*", ""),
    label = paste0("Spearman Correlation rho = ", round(rho, 2), star)
  )

# Find y position for annotation
annot_df <- st_df %>%
  group_by(Variable) %>%
  summarise(y = max(Positivity_Rate, na.rm = TRUE) + 2) %>%
  left_join(st_rho_p, by = "Variable") %>%
  filter(Variable %in% c("mean_ndvi", "mean_ndmi", "mean_ndwi"))

# Add annotation to plot
p_st_ann <- p_st +
  geom_text(
    data = annot_df,
    aes(x = -Inf, y = y, label = label),
    hjust = -0.1, vjust = 1.2, inherit.aes = FALSE, size = 4.5
  )

print(p_st_ann)
ggsave(file.path(output_dir, "25_spatiotemporal_vs_positivity.png"), p_st_ann, width = 14, height = 6, dpi = 300)
```

Significant correlations over all three variables, positive for NDWI, negative for NDVI and NDMI