---
title: "Variability analysis entomological indicators"
author: "Rita Verstraeten"
date: "`r Sys.Date()`"
output:
  pagedown::html_paged:
    toc: true
    # change to true for a self-contained document, but it'll be a litte slower for Pandoc to render
    self_contained: false
lot: true # insert a list of tables
lof: true # insert a list of figures
# uncomment this line to produce HTML and PDF in RStudio:
#knit: pagedown::chrome_print
---

```{r setup, include=FALSE}
source("renv/activate.R")
# renv::init()  # creates renv/ folder and lockfile
# Install necessary packages
# if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
# if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
library(readxl)
library(dplyr)
# install.packages(c("lubridate", "readr"))
library(lubridate)
library(readr)
library(DT)
library(ggplot2)
# if (interactive() && isTRUE(capabilities("httpgd"))) {
#   httpgd::hgd(silent = TRUE)  # use httpgd only in interactive sessions
# }
library(httpgd)
library(sf)
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)

results_dir <- "/home/rita/PyProjects/DI-MOB-BionamiX/results/Entomo/images"
```

# Introduction

This file checks the entomological indicator named: "Houses_pos_IS" for temporal variability both at the manzana and at the CMF level. 
The variability can be decomposed into two components: 
  1. Whether Houses_pos_IS > 0 (positivity)
  2. The actual value of Houses_pos_IS when > 0 

The original database has been filled such that every year-month combination is present. In case there was no entomological finding for 
that year-month combination in a particular manzana, the manzana was added with Houses_pos_IS = 0.

# Load data

### Entomological data
```{r load the entomological data}
entomo_path <- "/media/rita/New Volume/Documenten/DI-MOB/Data Sharing/WP2_Entomological_Data/Entomo_surveillance_11-24_secondary_DB_Month_Mz_v05082025.xlsx"
entomo <- suppressWarnings(read_excel(entomo_path, sheet = "Data"))
colnames(entomo)
entomo <- entomo %>%
  select(Year, Month, CODIGO_DIMOB, CMF, CP, Area, Houses_pos_IS, Inspected_houses)
head(entomo)
```

### Shapefiles 
```{r load the shapefiles}
mz_file_path <- "/media/rita/New Volume/Documenten/DI-MOB/Data Sharing/WP1_Cartographic_data/Administrative borders/Manzanas vectores 21_11_2024_CMF column uncleaned/manzanero_AS_1a8_19112024.shp"
cmf_file_path <- "/media/rita/New Volume/Documenten/DI-MOB/Data Sharing/WP1_Cartographic_data/Administrative borders/CMF/Poligonos CMF Cienfuegos_28032025.shp"

mz_shape <- st_read(mz_file_path)
cmf_shape <- st_read(cmf_file_path)

```

### Manzana allocation to CMF file 
```{r load the manzana allocation file}
mz_allocation_path <- "/media/rita/New Volume/Documenten/DI-MOB/Data Sharing/WP1_ Demographic_Data/Mzlist_overview CMF allocation_Shared between Areas_Original Mzlist_PerArea.xlsx"
mz_allocation <- suppressWarnings(read_excel(mz_allocation_path, sheet = "unique CMF allocation")) %>%
  select(1:10) %>% # selects first 10 columns (A-J)
  select(CODIGO_DIMOB, Area...2, MAP_PC, CMF_unico, MAP_Universo, Poblacion_3) %>%
  rename(CP = MAP_PC, 
  AREA = Area...2,
  CMF = CMF_unico,
  UNIVERSE = MAP_Universo)

head(mz_allocation)
# print(colnames(mz_allocation))
```

## Add the correct CMF to the entomological data based on the mz allocation file
```{r correction CMF entomo data}
entomo <- entomo %>%
  left_join(mz_allocation %>% select(CODIGO_DIMOB, CMF), 
            by = "CODIGO_DIMOB") %>%
            rename(CMF_ento = CMF.x, 
                   CMF_alloc = CMF.y)

head(entomo)
colnames(entomo)
```

```{r check the differences between CMF in entomo and the allocation file}
# inspect the differences between CMF_ento and CMF_alloc
entomo$CMF_ento = as.character(entomo$CMF_ento)
entomo$CMF_alloc = as.character(entomo$CMF_alloc)
entomo_diff <- entomo %>%
  filter(CMF_ento != CMF_alloc) %>%
  select(CODIGO_DIMOB, CMF_ento, CMF_alloc) 

nrow(entomo_diff)  # 168 differences
print(unique(entomo_diff))  # A8APGuMzFcPSOL 17       16
```

The differences in allocation between the CMFs in the entomological database and in the manzana allocation file are all about the same manzana: 
A8APGuMzFcPSOL, which is allocated to CMF 17 in the entomological database and to CMF 16 in the allocation file.

I will keep the allocations from the allocation file, as these are more up to date.

```{r update CMF in entomo data}
entomo <- entomo %>%
  select(-CMF_ento) %>%
  rename(CMF = CMF_alloc)%>%
  relocate(CMF, .after = CODIGO_DIMOB)
head(entomo)
```

# Entomo data MANZANA LEVEL - descriptive analysis 
### Check for missing values in Houses_pos_IS, CMF and Inspected_houses
```{r check missing variables CMF and Houses_inspected}
skimr::skim(entomo)

message("Number of missing Houses_pos_IS:", sum(is.na(entomo$Houses_pos_IS)))  # 0
message("Number of missing CMFs:", sum(is.na(entomo$CMF)))  # 672
message("Number of missing Inspected_houses:", sum(is.na(entomo$Inspected_houses)))  # 168

# check the manzanas with missing CMFs
missing_cmf_mz <- entomo %>%
  filter(is.na(CMF)) %>%
  select(CODIGO_DIMOB)

message("\n======= Missing CMF analysis =======")
message("Number entomo data entries with missing CMF value:", nrow(missing_cmf_mz))  # 672 manzanas with missing CMF
message("Number of unique manzanas with missing CMF value:", length(unique(missing_cmf_mz$CODIGO_DIMOB)))  # 4 manzanas with missing CMF
message("Manzanas with missing CMF value:", unique(missing_cmf_mz$CODIGO_DIMOB))  # A6Mz71   , A6MzFcJUCA , A6MzPASSAC , DD  

message("\n======= Missing Inspected_houses analysis =======")
missing_insp_houses <- entomo %>%
  filter(is.na(Inspected_houses)) %>%
  select(CODIGO_DIMOB)

message("Number entomo data entries with missing Inspected_houses value:", nrow(missing_insp_houses))  # 168 manzanas with missing Inspected_houses
message("Number of unique manzanas with missing Inspected_houses value:", length(unique(missing_insp_houses$CODIGO_DIMOB)))  # 1 manzana with missing Inspected_houses
message("Manzana with missing Inspected_houses value:", unique(missing_insp_houses$CODIGO_DIMOB))  # DD


message("\n======= Missing CODIGO_DIMOB analysis =======")
print(sum(is.na(entomo$CODIGO_DIMOB)))  # 0)
message("Number of entomo entries with MZ == DD:", nrow(entomo %>% filter(CODIGO_DIMOB == "DD")))  # 168
# missing_mz <- entomo %>%
#   filter(is.na(CODIGO_DIMOB)) %>%

# print(head(missing_mz))
```

### For now, remove the entries with MZ == DD 
```{r plot the Houses_pos_IS time series at manzana level}
entomo <- entomo %>%
  filter(CODIGO_DIMOB != "DD")

message("final database size: ", nrow(entomo))  # 236712 entries
length(unique(entomo$CODIGO_DIMOB))  # 1409 unique manzanas
```

### Filter for 2015-2018
```{r filter entomo data for 2015-2018}
entomo <- entomo %>%
  filter(Year >= 2015 & Year <= 2018)
message("final database size (2015-2018): ", nrow(entomo))  # 56688 entries
```

## Plot time series of Houses_pos_IS at manzana level 

### Heatmap of Houses_pos_IS over time for all manzanas

Create the helper variable is_positive to indicate whether Houses_pos_IS > 0. 
Follow this up with a logarithmic transformation of Houses_pos_IS (when is_positive) to better visualise the differences.
```{r plot Houses_pos_IS at manzana level}

# create the date to plot with
entomo_heatmap <- entomo %>%
  mutate(
    date = make_date(Year, Month, 1),
    is_positive = Houses_pos_IS > 0,
    Houses_pos_IS_log = ifelse(Houses_pos_IS > 0,
                               log1p(Houses_pos_IS), # log(1+x)
                               NA_real_)
  )

# # order manzanas by frequency of positivity
# entomo_heatmap_order <- entomo_heatmap %>%
#   group_by(CODIGO_DIMOB) %>%
#   summarise(
#     prop_positive = mean(is_positive, na.rm = TRUE)
#   ) %>%
#   arrange(prop_positive) %>%
#   pull(CODIGO_DIMOB)

# entomo_heatmap <- entomo_heatmap %>%
#   mutate(
#     CODIGO_DIMOB = factor(CODIGO_DIMOB, levels = entomo_heatmap_order)
#   )

entomo_heatmap <- entomo_heatmap %>%
  arrange(Area, CMF) %>%
  mutate(
    CODIGO_DIMOB = factor(CODIGO_DIMOB, levels = unique(CODIGO_DIMOB))
  )
```


```{r heatmap log Houses_pos_IS manzana level}

ggplot(entomo_heatmap, aes(x = date, y = CODIGO_DIMOB)) +
  geom_tile(aes(fill = Houses_pos_IS_log)) +
  scale_fill_viridis_c(
    name = "log(1 + Houses_pos_IS)",
    na.value = "grey95",
    option = "C"
  ) +
  labs(
    x = "Time",
    y = "CODIGO_DIMOB",
    title = "Temporal variability of Houses_pos_IS\n(zeros shown as background)"
  ) +
  theme_minimal() +
  theme(
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank()
  )
ggsave(file.path(results_dir, "Entomo_Heatmap_Houses_pos_IS_Mz_level.png"), width = 12, height = 12)
```

## Occurance vs intensity decomposition

```{r calculate manzana-level occurance metrics}
entomo <- entomo %>%
  mutate(
    date = make_date(Year, Month, 1),
    is_positive = Houses_pos_IS > 0
  )

occ_pos <- entomo %>%
  group_by(CODIGO_DIMOB) %>%
  summarise(
    prop_positive = mean(is_positive, na.rm = FALSE),
    n_positive = sum(is_positive, na.rm = FALSE),
    occurrence_sd = sd(is_positive),
    .groups = "drop"
  )

View(occ_pos)
```

```{r calculate manzana-level intensity metrics}
entomo_pos <- entomo %>%
  filter(Houses_pos_IS > 0)

int_pos <- entomo_pos %>%
  group_by(CODIGO_DIMOB) %>%
  summarise(
    mean_intensity = mean(Houses_pos_IS, na.rm = FALSE),
    intensity_sd = sd(Houses_pos_IS, na.rm = FALSE),
    cv_intensity = intensity_sd / mean_intensity,
    max_intensity = max(Houses_pos_IS, na.rm = FALSE),
    .groups = "drop"
  )
```
cv_intensity = sd(houses_pos) / mean(houses_pos) filtered for houses_pos > 0
So we scale variability relative to the mean. This measures the intrinsic variability of the positive values.

```{r combine occurance and intensity metrics at manzana level}
manzana_varianca_decomp <- occ_pos %>%
  left_join(int_pos, by = "CODIGO_DIMOB")

manzana_varianca_decomp %>%
  summarise(
    median_prop_positive = median(prop_positive, na.rm = TRUE),
    median_occurrence_sd = median(occurrence_sd, na.rm = TRUE),
    median_cv_intensity = median(cv_intensity, na.rm = TRUE)
  )

print(manzana_varianca_decomp)
```

1. The median manzana is positive 0.0625 months (i.e., 6.25% of months).
2. The median occurance sd is near 0.25 -> presence/absence fluctuates STRONGLY. 
3. The median cv_intensity is moderate -> when positivity does occur, the intensity varies moderately (so reasonably stable). 

**At the manzana level, the variability is almost entirely driven by rare, sporadic occurrence of posivity, while the magnitude is relatively stable when it does happen.**

# Entomo data CMF LEVEL - descriptive analysis 

### Aggregate the MZ-level counts to CMF level
```{r aggregate entomo data to CMF level}
entomo_cmf <- entomo %>%
  group_by(Year, Month, Area, CMF) %>%
  summarise(
    Houses_pos_IS = sum(Houses_pos_IS, na.rm = TRUE),
    Inspected_houses = sum(Inspected_houses, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Area_CMF = paste0("A", Area, "CMF", CMF))
``` 

### Check missing values in CMF

```{r check cmf entomo for missing values}
skimr::skim(entomo_cmf)
# View(entomo_cmf)
message("Unique CMFs in entomo_cmf data:", length(unique(entomo_cmf$Area_CMF)))  # 152 unique CMFs
```

## Plot the Heatmap per CMF 

Prep the data for heatmap plotting: add a date, order the CMFs by frequency of positivity. 

```{r plot Houses_pos_IS at CMF level}
# create the date to plot with
entomo_cmf_heatmap <- entomo_cmf %>%
  mutate(
    date = make_date(Year, Month, 1),
    is_positive = Houses_pos_IS > 0,
    Houses_pos_IS_log = ifelse(Houses_pos_IS > 0,
                               log1p(Houses_pos_IS), # log(1+x)
                               NA_real_)
  )

# order cmfs by Area (roman numeral) then CMF number
entomo_cmf_heatmap <- entomo_cmf_heatmap %>%
  arrange(Area, CMF) %>%
  mutate(
    Area_CMF = factor(Area_CMF, levels = unique(Area_CMF))
  )
```

```{r heatmap log Houses_pos_IS CMF level}

ggplot(entomo_cmf_heatmap, aes(x = date, y = Area_CMF)) +
  geom_tile(aes(fill = Houses_pos_IS_log)) +
  scale_fill_viridis_c(
    name = "log(1 + Houses_pos_IS)",
    na.value = "grey95",
    option = "C"
  ) +
  labs(
    x = "Time",
    y = "Area_CMF",
    title = "Temporal variability of Houses_pos_IS at CMF level\n(zeros shown as background)"
  ) +
  theme_minimal() +
  theme(
    axis.text.y  = element_text(size = 6)
  )

ggsave(file.path(results_dir, "Entomo_Heatmap_Houses_pos_IS_CMF_level.png"), width = 12, height = 12)
```


## Occurance vs intensity decomposition

```{r calculate CMF-level occurance metrics}
entomo_cmf <- entomo_cmf %>%
  mutate(
    date = make_date(Year, Month, 1),
    is_positive = Houses_pos_IS > 0
  )

occ_pos_cmf <- entomo_cmf %>%
  group_by(Area_CMF) %>%
  summarise(
    prop_positive = mean(is_positive, na.rm = FALSE),
    n_positive = sum(is_positive, na.rm = FALSE),
    occurrence_sd = sd(is_positive),
    .groups = "drop"
  )

View(occ_pos_cmf)
```

```{r calculate cmf-level intensity metrics}
entomo_pos_cmf <- entomo_cmf %>%
  filter(Houses_pos_IS > 0)

int_pos_cmf <- entomo_pos_cmf %>%
  group_by(Area_CMF) %>%
  summarise(
    mean_intensity = mean(Houses_pos_IS, na.rm = FALSE),
    intensity_sd = sd(Houses_pos_IS, na.rm = FALSE),
    cv_intensity = intensity_sd / mean_intensity,
    max_intensity = max(Houses_pos_IS, na.rm = FALSE),
    .groups = "drop"
  )
```
cv_intensity = sd(houses_pos) / mean(houses_pos) filtered for houses_pos > 0
So we scale variability relative to the mean. This measures the intrinsic variability of the positive values.

```{r combine occurance and intensity metrics at cmf level}
cmf_varianca_decomp <- occ_pos_cmf %>%
  left_join(int_pos_cmf, by = "Area_CMF")

cmf_varianca_decomp %>%
  summarise(
    median_prop_positive = median(prop_positive, na.rm = TRUE),
    median_occurrence_sd = median(occurrence_sd, na.rm = TRUE),
    median_cv_intensity = median(cv_intensity, na.rm = TRUE)
  )

print(cmf_varianca_decomp)
```

1. The median CMF is positive 0.479 months (i.e., 50% of months). -> no longer a rare-event at CMF level
2. The median occurance sd is near 0.47 -> presence/absence fluctuates STRONGLY. 
3. The median cv_intensity is moderate -> when positivity does occur, the intensity is more variable towards the mean than at Mz level.  

**Aggregation transformed the system from a rare-event, occurrence-dominated process into a frequently positive process, at the cost of increased relative variability in intensity due to spatial heterogeneity within CMFs**

# Autocorrelation analysis at both scales 

Most entomological models will assume memory. So if there are vectors in Mz x at timepoint t, then the chance is big that there will also be at t+1. 

Because the Mz-level data is zero-inflated, there will be 2 kinds of memory: 
1. **occurance*** memory -> this is more of a conditional probability than a correlation coeff 
  At Mz level we saw that occurance was rare (median prop_positive = 0.0625), so occurance memory will be low.
  At CMF level, occurance is more frequent (median prop_positive = 0.479), so occurance memory will be higher.
2. **intensity** memory -> closer to a classic autocorr. 
  At Mz level, 
  At CMF level, 

```{r prep the data for autocorr analysis}

prep_autocorr <- function(df, unit_col) {
  df %>%
    rename(unit = {{ unit_col }}) %>%
    mutate(
      date = make_date(Year, Month, 1),
      is_positive = Houses_pos_IS > 0
    ) %>%
    arrange(unit, date)
}
entomo_mz_p <- prep_autocorr(entomo, CODIGO_DIMOB)
entomo_cmf_p   <- prep_autocorr(entomo_cmf, Area_CMF)
```

### Occurance persistance
*If it's positive this month, does it stay positive next month?*

```{r define occurance persistence function}
occurrence_persistence <- function(df) {
  df %>%
    group_by(unit) %>%
    mutate(lag_pos = lag(is_positive)) %>% # default is lag by 1
    summarise(
      p_pos_given_pos = mean(is_positive & lag_pos, na.rm = TRUE) / # prop of still pos when pos last time
                        mean(lag_pos, na.rm = TRUE),
      p_pos_given_zero = mean(is_positive & !lag_pos, na.rm = TRUE) / # prop of pos when zero last time
                         mean(!lag_pos, na.rm = TRUE),
      p_neg_given_zero = mean(!is_positive & !lag_pos, na.rm = TRUE) / # prop of neg when zero last time
                         mean(!lag_pos, na.rm = TRUE),
      persistence_diff = p_pos_given_pos - p_pos_given_zero,  # difference in persistence
      .groups = "drop"
    )
}

occ_mz <- occurrence_persistence(entomo_mz_p)
occ_cmf <- occurrence_persistence(entomo_cmf_p)
```

```{r print occurance persistence results}
summary(occ_mz)

summary(occ_mz$persistence_diff)
summary(occ_cmf$persistence_diff)
```

### Intensity autocorrelation (lag 1)

*When it is positive, does the number of positive houses carry over to the next month?*

```{r compute lag-1 autocorrelation for intensity}
intensity_autocorr <- function(df, unit) {
  df %>%
    filter(Houses_pos_IS > 0) %>%
    group_by(unit) %>%
    summarise(
      acf1 = ifelse(
        n() >= 3,
        cor(
          log1p(Houses_pos_IS[-n()]),
          log1p(Houses_pos_IS[-1]),
          use = "complete.obs"
        ),
        NA_real_
      ),
      .groups = "drop"
    )
}
acintensity_mz <- intensity_autocorr(entomo_mz_p, CODIGO_DIMOB)
acintensity_cmf <- intensity_autocorr(entomo_cmf_p, Area_CMF)
```

There are some warnings coming up when calculating the autocorr, this is because there are measurements that are consistently the same (likely 0). 
So when there is no variance, the correlation cannot be calculated.

```{r print intensity autocorr results}
summary(acintensity_mz$acf1)
summary(acintensity_cmf$acf1)
```

There is a stronger autocorrelation at the MZ level. However, this is negative! 
This makes sense given that we know there are entomological interventions happening at the MZ level, which will likely reduce the intensity in the next month.

At the CMF level, the autocorrelation is positive but weak (0.04). 

```{r visualize intensity autocorr results}
library(ggplot2)
library(dplyr)

# Combine DIMOB and CMF for comparison
acf_combined <- bind_rows(
  acintensity_mz %>% mutate(scale = "CODIGO_DIMOB"),
  acintensity_cmf   %>% mutate(scale = "Area_CMF")
)

ggplot(acf_combined, aes(x = acf1, fill = scale)) +
  geom_histogram(alpha = 0.5, position = "identity", bins = 30) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Lag-1 autocorrelation (log1p intensity)",
    y = "Number of units",
    title = "Distribution of intensity autocorrelation by spatial scale",
    fill = "Scale"
  ) +
  theme_minimal()

ggplot(acf_combined, aes(x = scale, y = acf1, fill = scale)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.1, outlier.size = 1, alpha = 0.7) +
  labs(
    x = "Spatial scale",
    y = "Lag-1 autocorrelation (log1p intensity)",
    title = "Intensity autocorrelation by scale"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r visualize ocurrence persistence}
# Combine DIMOB and CMF
occ_combined <- bind_rows(
  occ_mz %>% mutate(scale = "CODIGO_DIMOB"),
  occ_cmf   %>% mutate(scale = "Area_CMF")
)

# Gather for plotting
occ_long <- occ_combined %>%
  select(scale, p_pos_given_pos, p_pos_given_zero) %>%
  tidyr::pivot_longer(cols = c(p_pos_given_pos, p_pos_given_zero),
                      names_to = "conditional", values_to = "prob")

ggplot(occ_long, aes(x = conditional, y = prob, fill = scale)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.1, outlier.size = 1, alpha = 0.7) +
  labs(
    x = "Condition",
    y = "Probability of positivity",
    title = "Occurrence persistence by spatial scale",
    fill = "Scale"
  ) +
  scale_x_discrete(labels = c("Given previous positive", "Given previous zero")) +
  theme_minimal()

```