\documentclass[12pt,a4paper]{article}

% Packages
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage[normalem]{ulem}
\usepackage{lastpage}
\usepackage{fancyhdr}

% Page layout
\geometry{margin=1in}

\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\fancyfoot[C]{\thepage/\pageref{LastPage}}

% Document information
\title{Bayesian Hierarchical State-Space Models:\\
\large Per-CMF AR(1) with Block RE and ICAR Spatial Effects}
\author{Rita Verstraeten}
\date{\today}

\begin{document}

\maketitle

\section{Introduction}

This document describes two Bayesian models for estimating the mosquito immature stages presence probability $p_{bt}$ at the CMF (circumscripción) level in Cienfuegos, Cuba, over monthly time periods.
Both models are implemented in Stan and differ only in how spatial heterogeneity between CMFs is captured:

\begin{description}
  \item[\texttt{AR\_perCMF\_blockRE}] The spatial component is an i.i.d.\ random intercept per CMF,
    capturing stable differences in baseline mosquito risk independently of neighbourhood structure.
  \item[\texttt{AR\_perCMF\_ICAR}] The spatial component is an Intrinsic CAR (ICAR) random effect
    that borrows strength from geographically adjacent CMFs, encoding the assumption that
    neighbouring areas tend to have similar residual risk.
\end{description}

All other model components --- the latent ecological process, environmental effects, temporal
dynamics, reactive surveillance correction, and observation model --- are identical.

\section{Model Structure}

\subsection{Latent Ecological Process}

For CMF $b$ at time $t$, the true mosquito presence probability is modelled as a latent
spatio-temporal process on the logit scale:
\begin{equation}
  \text{logit}(p_{bt}) = \alpha
  + \underbrace{\sum_{k=1}^{K} \sum_{\ell=0}^{L} w_{k\ell}\, X_{k,b,t-\ell}}_{\text{distributed lag effects}}
  + \underbrace{\sum_{k=1}^{K_u} w_k^u\, X_{k,b}^u}_{\text{static spatial covariates}}
  + v_{bt}
  + s_b,
  \label{eq:lp}
\end{equation}
where $\alpha$ is the global intercept, $v_{bt}$ is the per-CMF AR(1) temporal random effect
(Section~\ref{sec:ar}), and $s_b$ is the spatial random effect (Section~\ref{sec:spatial}).
Dengue cases $C_{bt}$ do not enter this equation; they affect only the observation process.

\subsection{Environmental Covariate Effects}

Environmental covariates enter the linear predictor in two ways.

\textbf{Lagged covariates} ($K$ variables, maximum lag $L = 2$ months) capture delayed ecological
responses to climatic drivers. Each combination of variable $k$ and lag $\ell$ receives an
independent weight $w_{k\ell}$:
\begin{equation}
  \text{lagged effect} = \sum_{k=1}^{K} \sum_{\ell=0}^{L} w_{k\ell}\, X_{k,b,t-\ell}.
\end{equation}
Continuous lagged variables are $z$-standardized before lag construction. Candidate lagged
variables include climatic drivers of Aedes larval development: rainfall indicators
(\texttt{total\_rainy\_days}, \texttt{total\_precip}, \texttt{precip\_max\_day},
\texttt{consec\_rainy\_days}, or categorical variants), vapour pressure deficit
(\texttt{avg\_VPD}), and temperature (\texttt{avg\_temp} or \texttt{temp\_cat}).
NDVI (\texttt{mean\_ndvi}) is always placed in the unlagged set.

\textbf{Unlagged (static spatial) covariates} ($K_u$ variables) capture persistent structural
features of the CMF that do not vary month to month (except for ndvi):
\begin{equation}
  \text{static effect} = \sum_{k=1}^{K_u} w_k^u\, X_{k,b}^u.
\end{equation}
These include \texttt{mean\_ndvi} (greenness, a proxy for landscape type), binary land-use indicators (\texttt{is\_urban}, wild-land urban interface \texttt{is\_WUI}, water-interface \texttt{is\_WI}), and domestic water infrastructure variables (\texttt{has\_aljibes}, \texttt{water\_shortage}, \texttt{water\_containers}). Binary variables are not standardized.

\subsection{Per-CMF AR(1) Temporal Effect}
\label{sec:ar}

Each CMF follows its own AR(1) trajectory, capturing block-specific seasonality and persistence
in mosquito presence:
\begin{align}
  v_{b,1} &= \frac{\sigma_v\, \varepsilon_{b,1}}{\sqrt{1 - \rho^2}}, \\
  v_{b,t} &= \rho\, v_{b,t-1} + \sigma_v\, \varepsilon_{b,t}, \quad \varepsilon_{b,t} \overset{\text{iid}}{\sim} \text{Normal}(0,1), \quad t \geq 2.
\end{align}
The first time point is initialised at the stationary distribution of the AR(1) process.
The AR(1) coefficient $\rho \in (-1,1)$ and innovation SD $\sigma_v > 0$ are \textbf{shared across
all CMFs} (partial pooling), so each CMF trajectory is an independent realisation of the same
stochastic process. This lets each CMF track its own seasonal rhythm while the global persistence
and variability are informed by all CMFs jointly (see Table~\ref{tab:priors}).

\section{Spatial Random Effects}
\label{sec:spatial}

\subsection{Model 1 --- I.I.D.\ Block Random Effect (\texttt{AR\_perCMF\_blockRE})}

Each CMF receives a static scalar offset drawn independently from a common distribution:
\begin{equation}
  s_b = u_b^{\text{block}} = \sigma_{\text{block}}\, z_b, \qquad z_b \overset{\text{iid}}{\sim} \text{Normal}(0,1).
\end{equation}
This accounts for stable, unmeasured CMF-level risk factors (e.g.\ sewer infrastructure, housing
density). Because the offsets are independent, this model does not assume that neighbouring CMFs
are more similar than distant ones --- it only shrinks all CMF effects toward zero.

\subsection{Model 2 --- ICAR Spatial Effect (\texttt{AR\_perCMF\_ICAR})}

The ICAR prior penalises differences between adjacent CMFs, encoding spatial smoothness:
\begin{equation}
  p(\tilde{\mathbf{u}}) \propto \exp\!\left(-\frac{1}{2} \sum_{(b,b') \in \mathcal{E}} \bigl(\tilde{u}_b - \tilde{u}_{b'}\bigr)^2\right),
\end{equation}
where $\mathcal{E}$ is the set of undirected edges of the CMF adjacency graph (shared boundaries,
100\,m snap tolerance). The scaled effect is $s_b = u_b^{\text{ICAR}} = \sigma_{\text{ICAR}}\,\tilde{u}_b$.

Because the ICAR prior is improper (it only constrains \emph{differences}, not absolute levels),
an identifiability constraint is imposed:
\begin{equation}
  \sum_{b=1}^{B} \tilde{u}_b \sim \text{Normal}(0,\, 0.001\,B).
\end{equation}
In Stan this is coded as a single sparse dot product over the edge list, costing $O(|\mathcal{E}|)$
per HMC step. The ICAR model is preferred when residual spatial autocorrelation remains after
accounting for covariates --- for example, due to short-range Aedes dispersal across CMF
boundaries or spatially clustered unmeasured risk factors.

\section{Reactive Surveillance and Observation Model}

\subsection{Reactive Inspection Bias}

Dengue cases $C_{bt}$ trigger $\kappa$ additional household inspections beyond routine surveillance.
Because these inspections target locations associated with cases, they inflate observed positivity
relative to the ecological probability $p_{bt}$. The reactive observation probability is:
\begin{equation}
  p_{bt}^R = \text{logit}^{-1}\!\bigl(\text{logit}(p_{bt}) + \delta_1\, C_{bt}\bigr),
  \quad \delta_1 \geq 0,
\end{equation}
where $\delta_1 \geq 0$ scales the log-odds boost per additional dengue case. The total inspection count is $n_{bt} = N_b^{HH} + \kappa C_{bt}$ with $\kappa = 2$
fixed, and the reactive fraction is $\omega_{bt} = \min(1,\, \kappa C_{bt} / n_{bt})$.

\subsection{Effective Observation Probability and Likelihood}

The effective probability that a randomly selected inspection is positive mixes the two
surveillance regimes:
\begin{equation}
  \pi_{bt} = \begin{cases}
    p_{bt} & \text{if } C_{bt} = 0, \\
    (1 - \omega_{bt})\,p_{bt} + \omega_{bt}\,p_{bt}^R & \text{if } C_{bt} > 0.
  \end{cases}
\end{equation}
The observation model is a Beta-Binomial to accommodate overdispersion beyond the Binomial:
\begin{equation}
  y_{bt} \sim \text{BetaBinomial}\!\left(n_{bt},\; \pi_{bt}\phi,\; (1-\pi_{bt})\phi\right),
\end{equation}
where $\phi > 0$ is the concentration parameter controlling overdispersion ($\phi \to \infty$
recovers the Binomial). It is either fixed from a data-driven estimate or estimated from data
(see Table~\ref{tab:priors}).

\section{Prior Distributions}

Table~\ref{tab:priors} summarises all prior distributions. Parameters marked with an asterisk
are model-specific; all others are shared between \texttt{AR\_perCMF\_blockRE} and
\texttt{AR\_perCMF\_ICAR}.

\begin{table}[ht]
\centering
\small
\begin{tabular}{p{2.0cm}p{3.2cm}p{3.0cm}p{5.8cm}}
\toprule
\textbf{Parameter} & \textbf{Role} & \textbf{Prior} & \textbf{Reasoning} \\
\midrule
$\alpha$ & Global intercept & Normal$(-7.0,\; 1.5)$ & Centred on low baseline prevalence (0.5--5\% positivity corresponds to logit $\approx -5.3$ to $-2.9$). \\
\addlinespace
$w_{k\ell}$ & Lag weight for covariate $k$ at lag $\ell$ & Normal$(0,\; 1.0)$ & Weakly regularising; allows effects of up to $\pm 2$ log-odds per SD shift. \\
\addlinespace
$w_k^u$ & Unlagged covariate weight & Normal$(0,\; 0.5)$ & Tighter than lag weights; static spatial predictors typically have smaller marginal effects. \\
\addlinespace
$\rho$ & AR(1) coefficient (shared) & Normal$(0.35,\; 0.1)$, $\rho \in (-1,1)$ & Centred on moderate positive persistence, consistent with seasonal mosquito data; constrained to stationarity. \\
\addlinespace
$\sigma_v$ & AR(1) innovation SD (shared) & Normal$^+(0,\; 0.3)$ & Allows moderate month-to-month variability within each CMF trajectory; half-normal via lower bound. \\
\addlinespace
$\sigma_{\text{block}}^*$ & SD of i.i.d.\ block offsets & Normal$^+(0,\; 0.5)$ & Mean $\approx 0.4$ log-odds; allows substantial between-CMF heterogeneity while shrinking toward zero. \\
\addlinespace
$\sigma_{\text{ICAR}}^*$ & Marginal SD of ICAR effect & Normal$^+(0,\; 0.3)$ & Similar scale to $\sigma_{\text{block}}$; slightly tighter because the ICAR prior already pools across neighbours. \\
\addlinespace
$\delta_1$ & Reactive surveillance boost & Normal$^+(0,\; 0.5)$ & Constrained positive (reactive inspections inflate, not deflate, positivity); weakly regularised. \\
\addlinespace
$\phi$ & Beta-Binomial concentration & Gamma$(13,\; 0.1)$ or fixed & Centred on the empirically observed posterior (mean 130, mode 120, SD $\approx 36$); still weakly informative. Fixed to a data-driven estimate when \texttt{fix\_phi = TRUE}. \\
\bottomrule
\end{tabular}
\caption{Prior distributions for all model parameters. $^*$Model-specific: $\sigma_{\text{block}}$ applies to \texttt{AR\_perCMF\_blockRE}; $\sigma_{\text{ICAR}}$ applies to \texttt{AR\_perCMF\_ICAR}. Normal$^+$ denotes a half-normal (lower bound 0 in Stan).}
\label{tab:priors}
\end{table}

\section{Model Selection via LOO-IC}

\subsection{Leave-One-Out Information Criterion}

To compare model variants --- differing in spatial component (block RE vs.\ ICAR) or in
covariate composition --- we use the Leave-One-Out Information Criterion (LOO-IC), estimated by
Pareto-smoothed importance sampling \mbox{(PSIS-LOO;} Vehtari et al.\ 2017;
\href{https://mc-stan.org/loo/reference/elpd.html}{Stan : Understanding ELPD an LOO}):
\begin{equation}
  \text{LOO-IC} = -2\,\widehat{\text{elpd}}_{\text{loo}},
  \qquad
  \widehat{\text{elpd}}_{\text{loo}} = \sum_{i=1}^{N} \log \hat{p}(y_i \mid y_{-i}).
\end{equation}
Each summand estimates how well the fitted model predicts a held-out observation, integrated
over the posterior. The LOO-IC is computed from \texttt{log\_lik[i]} stored in the Stan
\texttt{generated quantities} block.

\subsection{Comparing Models}

\texttt{loo::loo\_compare} ranks models by $\widehat{\text{elpd}}_{\text{loo}}$ (higher is better)
and reports, for each model $m$ relative to the best:
\begin{equation}
  \Delta\widehat{\text{elpd}}_m \leq 0,
  \qquad
  z_m = \frac{\Delta\widehat{\text{elpd}}_m}{\text{SE}(\Delta\widehat{\text{elpd}}_m)}.
\end{equation}
A rule of thumb: $|z_m| > 2$ indicates a practically meaningful difference in predictive
performance. Models with $|z_m| \leq 2$ are considered equivalent within the uncertainty of
the PSIS-LOO estimates. In the variable sweep, each covariate combination is fitted with both
spatial structures; the LOO comparison across all runs simultaneously identifies the better
spatial component \emph{and} the better covariate set.

\end{document}